---
title: "Compare streambank & Midwest samples"
author: "Christy Dolph"
date: "2026-08-31"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Overview

This script compares TP, depth and other geospatial predictors representative of soils, climate and land use among streambank and Midwest datasets.

# Workspace Setup

Load packages and set directories.

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace 
rm(list=ls())


#install and load packages

#list of necessary packages 
want <- c('data.table', 'tidyverse', 'readxl', 'writexl', 'data.table', 'mapview', 'sf', 'terra', 'tmap','geodata', 'RSelenium', 'rvest', 'httr', 'dataRetrieval', 'DT', 'FedData', 'raster', 'nhdplusTools', 'tigris', 'janitor', 'segmented', 'mcp','grDevices', 'sp', 'Hmisc','tools', 'randomForest', 'plyr', 'devtools', 'pbapply', 'digest', 'grid')


need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)


#Install StreamCatTools (if not already installed):
# First install remotes if needed
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
# Install StreamCatTools
if (!requireNamespace("StreamCatTools", quietly = TRUE)) {
  remotes::install_github("USEPA/StreamCatTools")
}
# Load the package
library(StreamCatTools)



# Create subdirectory shortcuts

#Main directory
dir.root<-getwd()

#Input data
dir.data<-paste(dir.root, "/Streambank_P_measured_data", sep="")

#Spatial data
dir.spatial<-paste(dir.root, "/Spatial_data", sep="")

#Output data (produced by script)
dir.output<-paste0(dir.root, "/output_data")

dir.figures<-paste0(dir.root, '/Maps_figures')

# Create additional folder destinations, if they don't already exist:
dir.create(file.path(getwd(), "Spatial_data"))
dir.create(file.path(dir.spatial, "nlcd"))
dir.create(file.path(dir.spatial,"ssurgo"))
dir.create(file.path(dir.root,"output_data"))
dir.create(file.path(dir.root, "Maps_figures"))


```

# Compare TP among streambank samples (by major watershed) and Midwestern samples

```{r}


#Margenot et al IL streambank P data, with limited ssurgo attributes to retain all data, for independent streambank model or for running through existing soil TP model (predictors are matched below): 
model.input<-"/Margenot_sample_attributes_for_modelling_wSSURGO_LIMITED_comid"

streambank <-read.csv(paste0(dir.output, model.input, '.csv'))


#Count number of streambank sample locations by NHD stream reach
names(streambank)

Num.samples.bycomid<-streambank |> dplyr::select(x,y,comid) |> distinct() |> dplyr::select(comid) |> group_by(comid) |> count()
Num.samples.bycomid

summary(Num.samples.bycomid$freq)



#For streambank data, get NHD attributes, including HUC8

#create shapefile of sample locations:

streambank_sf<-st_as_sf(streambank, coords=c("x", "y"), crs=4269, remove=FALSE)


#creating bounding box for sample locations

bb = st_as_sfc(st_bbox(streambank_sf))


#Get HUC8 info
huc8_polygon <- get_huc(AOI = bb, type = "huc08")

# 4. Extract the HUC8 Code and Name
# drop the spatial geometry layer to return a clean data frame
huc8_info <- huc8_polygon %>%
  st_drop_geometry() %>%
  dplyr::select(huc8, name)

#OR, spatial join huc8 ID directly to streambank locations, if are plotting by HUC8

streambank_huc8<-st_join(streambank_sf, huc8_polygon)
head(streambank_huc8)

#Count number of streambank sample locations by HUC8

names(streambank_huc8)

Num.samples.byHUC8<-streambank_huc8 |> dplyr::select(x,y,huc8,name) |> st_drop_geometry() |> distinct() |> dplyr::select(huc8, name) |> group_by(huc8) |> count()
Num.samples.byHUC8

write.csv(Num.samples.byHUC8, paste0(dir.figures, "/Table_streambank_samples_by_HUC8.csv"), row.names=FALSE)


#Mean number of samples per watershed
summary(Num.samples.byHUC8$freq)

write.csv(Num.samples.byHUC8, paste0(dir.figures, "/Table_streambank_samples_by_HUC8.csv"), row.names=FALSE)


#Modern Midwest Soil P database:
model.input2<-'/USGS_NCSS_soils_database_attributes_alldepths_for_modelling_wSSURGO_LIMITED'


#Modern Midwest soil P database but with limited ssurgo attributes that should encompass water areas?
#model.input<-'/USGS_NCSS_soils_database_attributes_alldepths_for_modelling_wSSURGO_LIMITED_comid'

Midwest <-read.csv(paste0(dir.output, model.input2, '.csv'))


#Remove any invalid P measurements
Midwest<-Midwest %>% filter(P_mgkg>0)

#combine Midwest and streambank values

streambank.sel<-streambank_huc8 %>% mutate(dataset="Streambank") |> dplyr::select(name, P_mgkg, Depth_cm, dataset)

Midwest.sel<-Midwest %>% mutate(dataset="Midwest") |> dplyr::select(P_mgkg, Depth_cm, dataset) %>% mutate(name="Midwest")

both<-bind_rows(streambank.sel, Midwest.sel)

#fix order of labels, put Midwest last
library(forcats)

both$name<-fct_relevel(factor(both$name), "Midwest", after = Inf)

#Plot by HUC8 and depth as well

depth.plot<-ggplot(both)+
  geom_point(aes(Depth_cm, P_mgkg))+
  geom_smooth(aes(Depth_cm, P_mgkg), method="lm", se=FALSE)+
  facet_wrap(~name)
depth.plot

#Plot by depth only
depth.plot<-ggplot(both)+
  geom_point(aes(log10(Depth_cm), log10(P_mgkg)), alpha=0.3)+
  #geom_smooth(aes(log(Depth_cm), log(P_mgkg)), method="lm")+
  #scale_shape_manual(values = c(0))+
  facet_wrap(~dataset)+
  labs(x="Log depth (cm)",y = bquote("Log TP mgkg"^-1))+
  #xlab("Log soil depth (cm)")+
  #ylab("Log TP (mg/kg)")+
  #ylim(1,3.5)+
  #labs(color = "Dataset")+
  theme_bw()+
  theme(strip.text=element_text(size=14), axis.title = element_text(size = 14),
        axis.text = element_text(size = 14), legend.title=element_text(size=14), legend.text=element_text(size=14))+
  scale_y_continuous(breaks = scales::breaks_width(0.5))
depth.plot
  
#write to file
png(paste0(file.name=dir.figures, "/Fig_streambank_vs_Midwest_TP_DEPTH.png"), width = 2800, height = 1400, res = 300)    

grid.newpage() 
print(depth.plot) 

dev.off()   



#Plot across HUC8s

TP.compare<-ggplot(both, aes(factor(name), P_mgkg))+
  geom_jitter(width = 0.15, height = 0, alpha = 0.1, color = "darkblue") +
  geom_violin(trim = FALSE, fill = "lightblue", alpha = 0.5, quantiles=0.5, quantile.linetype = "solid")+
  geom_abline(slope=0, intercept=512.96, linetype="dashed")+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size=14),
    axis.text.y=element_text(size=14),
      axis.title=element_text(size=16),
        panel.grid.major.y=element_line(color="grey90", a),
        panel.grid.minor=element_line(color="grey95"), 
        panel.background=element_rect(fill = "transparent", colour = "black"))+
  labs(x="",y = bquote("Total Phosphorus mgkg"^-1))
TP.compare

#write to file
png(paste0(file.name=dir.figures, "/Fig_streambank_vs_Midwest_TP.png"), width = 2900, height = 1800, res = 300)    

grid.newpage() 
print(TP.compare) 

dev.off()  

```

# Compare geology and soils across streambank and Midwest samples

```{r}

#Subset to same attributes across both datasets

streambank.sub<-streambank |> dplyr::select(any_of(colnames(Midwest))) |> mutate(dataset="Streambank") |> mutate(NLCD_Land=as.factor(NLCD_Land))

names(streambank.sub)

Midwest<-Midwest |> mutate(dataset="Midwest")|> mutate(NLCD_Land=as.factor(NLCD_Land))

all<-bind_rows(streambank.sub, Midwest) 

names(all)

#assign states to locations
#read in USA state boundaries using tigris package
states<-states()
states_sf <- st_as_sf(states) #convert to sf object
st_bbox(states_sf) #bounding box
states_sf

st_crs(states_sf) #check reference system

#create spatial object for sample locations:
all_sf<-st_as_sf(all, coords=c("x", "y"), crs=4269, remove=FALSE)

all_state<-st_join(all_sf, states_sf)
names(all_state)


#select a subset of attributes to compare
geo.sel<-all_state |> dplyr::select(P_mgkg, dataset, STUSPS, rocknws,pctglactilcrsws, pcthydricws, compstrgthws, caows, p2o5ws, hydricon, hydricrating, drainagecl, taxpartsize, taxceactcl,taxreaction, pctagdrainagews, pctagslphigh2011ws, septicws, wwtpmajordensws, manurews, elevws,sio2ws, na2ows, k2ows, fe2o3ws, p2o5ws, al2o3ws, caows, mgows, hydrlcondws, kffactws, pctglactilloamws, pctglactilcrsws, pcthydricws, pctcrop2011ws, pestic1997ws, precip8110ws, precip9120ws, tmean9120ws, clayws, sandws, omws) |> st_drop_geometry()


#attributes without data for streambank samples:
#caco3_r
#claytotal_r
#freeiron_r
#ksat_r, kffact
#sandtotal_r 
#silttotal_r 

#hm, if data is missing for attributes that are used in Midwest model, that might explain poor performance somewhat??
#make an Midwest model that includes only attributes that are non NA for streambanks?

sapply(geo.sel, class)

class(geo.sel)


#Look at numeric attributes first

#need to make nice plot labels, or define them in legend?

#Option to select Subset of states
#Option to select specific attributes for geology, soils, climate, land use






geo.sel.long<-geo.sel |> 
#if want to select states in immediate region:
#filter(STUSPS %in% c('IL', 'OH', 'IA', 'IN')) |> 
  dplyr::select(where(is.numeric), dataset) |> 
  pivot_longer(
    cols = -c(dataset, P_mgkg),         # Pivot every column EXCEPT your factor column and TP
    names_to = "variable",     # New column holding the old column names
    values_to = "value"        # New column holding the numeric values
  ) |> 
#select specific attributes
  dplyr::filter(variable %in% c("clayws", "sandws", "pctglactilloamws", "pcthydricws", "pctcrop2011ws", "pctagdrainagews", "pestic1997ws", "precip9120ws", "tmean9120ws"))

head(geo.sel.long)

#make nice labels for plot
levels(factor(geo.sel.long$variable))

geo.sel.long$variable<-factor(geo.sel.long$variable, 
  levels=c("clayws", "sandws", "pctglactilloamws", "pcthydricws", "pctcrop2011ws", "pctagdrainagews", "pestic1997ws", "precip9120ws", "tmean9120ws"),
  labels=c("% Clay", "% Sand", "% Glacial till - loamy", "% Hydric soils", "% Crop cover, 2011", "% Ag Drainage",  "Pesticide use - 1997", "Average annual precip, 1991-2020", "Average annual temp, 1991-2020"))

compare.plot<-
ggplot(geo.sel.long, aes(x = dataset, y = value, fill = dataset)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y", labeller = label_wrap_gen(width = 20)) +  # 'free_y' allows each plot to have its own Y-axis range
  theme(axis.text.x=element_blank(), axis.text.y=element_text(size=16), axis.title=element_text(size=16), legend.text=element_text(size=16), legend.title=element_text(size=16), strip.text=element_text(size=16)) +
  labs(x = "Sample Type", y = "Value")+
  scale_fill_manual(values=c("#7570b3","#d95f02"))
compare.plot

#Write to file 
png(paste0(file.name=dir.figures, "/Fig_compare_streambank_vs_upland_attributes.png"), width = 3000, height = 2400, res = 300)    

print(compare.plot)

dev.off()  

```

# TP vs key predictors

Load in imputed training data for streambank and Midwest datasets

```{r}

#Read in imputed data, so aren't dealing with NAs


model.input1<-"/Margenot_sample_attributes_for_modelling_wSSURGO_LIMITED"

model.input2<-'/USGS_NCSS_soils_database_attributes_alldepths_for_modelling_wSSURGO_LIMITED'

#READ IMPUTED DATA from Model Tuning module: 
dir.final.model.data<-paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets")


#(note this is training data only)

#Streamabank data, with soil: 

P.Predictors1<-read.csv(paste0(dir.final.model.data, model.input1, "_train_imputed_SOIL_08132026.csv")) |> mutate(dataset="Streambank")

P.Predictors2<-read.csv(paste0(dir.final.model.data, model.input2, "_train_imputed_groupsplit_08252026.csv")) |> mutate(dataset="Midwest")

# Combine datasets into a single master data frame
P.Predictors <- bind_rows(P.Predictors1, P.Predictors2)

P.Predictors1 |> filter(Depth_cm<30) |> nrow()

```

Calculate R2 and p values for each dataset

```{r}

target_col <- "P_mgkg"

# 1. Pivot numeric predictors long, grouping by dataset and variable
results_df <- P.Predictors |> 
  dplyr::select(dataset, all_of(target_col), where(is.numeric)) |> 
  pivot_longer(
    cols = -c(dataset, all_of(target_col)), 
    names_to = "Variable", 
    values_to = "Predictor_Value"
  ) |> 
  drop_na(Predictor_Value, all_of(target_col)) |> 
  group_by(dataset, Variable) |> 
  # Fit linear model per dataset and predictor variable
  summarise(
    fit = list(lm(as.formula(paste0("`", target_col, "` ~ Predictor_Value")), data = cur_data())),
    .groups = "drop"
  ) |> 
  # Extract R2 and p-value from model summary
  mutate(
    sum_fit   = map(fit, summary),
    R_Squared = map_dbl(sum_fit, ~ round(.x$r.squared, 2)),
    P_Value   = map_dbl(sum_fit, ~ {
      co <- .x$coefficients
      if (nrow(co) >= 2) round(co[2, "Pr(>|t|)"], 4) else NA_real_
    })
  ) |> 
  dplyr::select(dataset, Variable, R_Squared, P_Value)

head(results_df)

View(results_df)

#Write R2 and p values to table
write_xlsx(results_df, paste0(dir.figures, "/Table_R2_p_values_predictors_vs_TP.xlsx"))

# Check count of significant relationships per dataset
results_df |> 
  filter(P_Value < 0.05) |> 
  count(dataset)

```

TP vs predictors, for variables with highest R2 and significant p value

```{r}
#For regression, select numeric attributes
df <- P.Predictors |>  dplyr::select(where(is.numeric))
target_col <- "P_mgkg"    # Replace with your target column name

# 1. Get the names of all other columns
other_cols <- setdiff(names(df), target_col)

# 2. Run regression across all columns
results_list <- lapply(other_cols, function(col) {
  formula_str <- as.formula(paste0("`", target_col, "` ~ `", col, "`"))
  
  fit <- lm(formula_str, data = df)
  sum_fit <- summary(fit)
  
  r2 <- sum_fit$r.squared
  # Extract the p-value for the predictor variable safely
  p_val <- if (nrow(sum_fit$coefficients) >= 2) {
    sum_fit$coefficients[2, "Pr(>|t|)"]
  } else {
    NA
  }
  
  data.frame(
    Variable = col, 
    R_Squared = round(r2, 2), 
    P_Value = round(p_val, 4),
    stringsAsFactors = FALSE
  )
})

# 3. Combine into a master summary table
results_df <- do.call(rbind, results_list)

results_df<-results_df |> mutate(dataset="Midwest")

head(results_df)

#How many significant relationships?
results_df |> filter(P_Value<0.05) |> nrow()

nrow(results_df)

View(results_df)

# 4. Filter variables with R2 > 0.10 and sort descending by R2
stronger <- results_df |> 
  filter(R_Squared >= 0.04) |> 
  arrange(desc(R_Squared))

stronger

# Make nicer labels for plotting
# Note: Ensure the replacement vector matches the order of 

levels(factor(stronger$Variable))

#For MIDWEST
stronger$Label <- factor(
  stronger$Variable, 
  labels = c(
  "Habitat score",
  "Hydrologic regulation score",
  "% Crop - cat",
  "% Crop - ws",
  "% Coarse eolian sed, cat",
  "% Coarse eolian sed, ws",
  "Soil permeability",
  "Annual precip - 2009, cat",
  "Annual precip - 2009, ws",
  "Soil organic carbon, 0-20cm",
  "Soil organic carbon, 0-30cm",
  "Soil organic carbon, 5-20cm",
  "Mean stream temp"))
  
  
#FOR STREAMBANK:
stronger$Label <- factor(
  stronger$Variable, 
  labels = c(
    "drainagecl",
    "Anthropogenic N inputs",
    "NLCD_Land",
    "NPDES permit density",
    "N surplus",
    "% Organic matter",
    "% Clay - measured",
    "% Sand - measured",
    "% Silt - measured",
    "% Colluvial sediment",
    "% Herbaceous wetland",
    "% Hydric soils",
    "% Open Water",
    "Mean bedrock depth",
    "Water input",
    "WWTP density"
  )
)

# 5. Create formatted facet labels using original Variable names as keys
facet_labels <- setNames(
  paste0(stronger$Label, " (R² = ", sprintf("%.2f", stronger$R_Squared), ")"),
  stronger$Variable # Map FROM original column name TO display string
)

# 6. Pivot selected variables and reorder levels by R2
streambank.sel <- P.Predictors |> 
  dplyr::select(all_of(stronger$Variable), P_mgkg) |> 
  dplyr::select(-any_of(c("soc5_20", "soc0_20", "soc0_30"))) |> 
  mutate(across(everything(), as.numeric)) |> 
  pivot_longer(
    cols = -c(P_mgkg),
    names_to = "variable",
    values_to = "value"
  ) |> 
  # Fixed typo: stronger$Variable instead of stronger$variable
  mutate(variable = factor(variable, levels = stronger$Variable))

# 7. Plot ordered facets with R2 in the top header strip
scatterplot<-ggplot(streambank.sel, aes(x = value, y = P_mgkg)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed") +
  facet_wrap(
    ~variable, 
    scales = "free", 
    labeller = labeller(variable = facet_labels)
  ) +
  theme_minimal() +
  labs(
    x = "Predictor Value",
    y = "P_mgkg",
    title = "Linear Relationships with TP - Midwest samples"
  )
scatterplot


#Write to file 
png(paste0(file.name=dir.figures, "/Fig_variables_vs_TP_scatterplots_MIDWEST.png"), width = 3200, height = 2400, res = 300)    

print(scatterplot)

dev.off()  
```
