---
title: "Assigning spatial attributes to sample locations"
author: "Christy Dolph"
date: "2024-07-17"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Overview: Generating spatial attributes for sites of interest

This script generates spatial attributes associated with specific point locations.

Attributes are from the following sources:

-   the NHDPlusv2 stream network (NHD)

-   the National Wetlands Inventory (NWI),

-   National Land Cover Dataset (NLCD)

-   gSSURGO soils data

-   EPA StreamCat

This approach is updated and adapted from the scripts available at: <https://github.com/cldolph/SoilP>

## Steps:

1.  Merge locations to NHD, NLCD, and NWI attributes + SSURGO map unit key (mukey)
2.  Compile and process SSURGO attributes and merge
3.  Compile and process StreamCat attributes and merge

# National Hydrography Dataset (NHD)

Use nhdplusTools package to link sample locations to NHD COMID (unique identifier for each stream reach of the NHD).

The COMID can then be used to link the location to NHD attributes (stream type, stream order, etc) as well as EPA StreamCat attributes.

To do this, we will join sample locations to NHD catchments. This avoids issues with sample location accuracy relative to flowlines (if you tried to join sample points to NHD flowlines rather than catchments).

\*\*Note that 'featureid' of an NHD Catchment is equal to COMID of an NHD flowline; see p. 47 of the NHDPlusV2 User Guide:

[https://www.epa.gov/system/files/documents/2023-04/NHDPlusV2_User_Guide.pdf](https://www.epa.gov/system/files/documents/2023-04/NHDPlusV2_User_Guide.pdf,)

## Download NHD catchment boundaries for area of interest

```{r}

#Note, this can take considerable time

Sys.time()

NHD_catchment<-get_nhdplus(
  AOI = HUC4_AOI,
  comid = NULL,
  nwis = NULL,
  realization = "catchment",
  streamorder = NULL,
  t_srs = NULL
)

Sys.time()

names(NHD_catchment)

#View(NHD_catchment)

#Create COMID column
NHD_catchment$comid<-NHD_catchment$featureid

#Quick map check:
tm_shape(NHD_catchment)+
  tm_borders()

#write to file:
st_write(NHD_catchment, paste0(dir.spatial, "/nhd_catchment.shp"), append=FALSE)

```

## Identify NHD COMIDs associated with sample locations

see NHDv2Plus User Guide: <https://nhdplus.com/NHDPlus/NHDPlusV2_documentation.php>

```{r}
#first transform pour point and flowlines to same projection: 
#continental US (Conus Albers NAD83)  
loc_proj <- st_transform(loc_sf, 5070) 
NHD_catch_proj<-st_transform(NHD_catchment, 5070)

#Join catchment attributes to soil sample locations
loc_NHD <- st_join(loc_proj, NHD_catch_proj %>% dplyr::select(comid, areasqkm), join = st_intersects)  

View(loc_NHD)

```

# National Wetlands Inventory (NWI)

See access to NWI data at: <https://www.fws.gov/program/national-wetlands-inventory/data-download>

Download data by state: <https://www.fws.gov/program/national-wetlands-inventory/download-state-wetlands-data>

### Load in wetlands data:

```{r}

#US states where samples occur (corresponding to study area):
states<-list(States_sub$STUSPS)
states


##To download NWI files for every state in the your study area of interest, loop GET url for multiple states:

# Base URL where NWI data is available
base_url <- "https://documentst.ecosphere.fws.gov/wetlands/data/State-Downloads"

# Loop through state abbreviations and download corresponding zip files

for (i in states) {
  zip_name <- paste0(i, "_geodatabase_wetlands.zip")
  url <- paste0(base_url, "/", zip_name)
  destfile <- file.path("./spatial_data", zip_name)
  
  if (!file.exists(destfile)) {
    message("Downloading: ", url)
    tryCatch({
      response <- GET(url, write_disk(destfile, overwrite = TRUE),
                      user_agent("Mozilla/5.0 (Windows NT 10.0; Win64; x64)"))
      if (response$status_code != 200) {
        warning("Download failed for ", states, " with status ", response$status_code)
      }
    }, error = function(e) {
      warning("Failed to download ", states, ": ", e$message)
    })
  } else {
    message("Already downloaded: ", states)
  }
}

# Unzip geodatabase(s)

#List of zipped files:
zip_wetlands <- list.files("./spatial_data", pattern = "wetlands.*\\.zip$", full.names = TRUE)
zip_wetlands

#Unzip files to spatial data directory:
ldply(.data = zip_wetlands, .fun = unzip, exdir = "./spatial_data")


##Read in wetlands layer from every geodatabase (from each different state) in study region
  

# Initialize a list to hold each state's wetlands layer
wetlands_list <- list()
wetlands_list
# Loop over state abbreviations and read each layer into the list

for (i in states) {
  gdb_path <- file.path("spatial_data", paste0(i, "_geodatabase_wetlands.gdb"))
  layer_name <- paste0(i, "_Wetlands")

  if (dir.exists(gdb_path)) {
    message("Reading layer for state: ", i)
    tryCatch({
      wetlands_list[[i]] <- st_read(dsn = gdb_path, layer = layer_name)
    }, error = function(e) {
      warning("Failed to read ", i, ": ", e$message)
    })
  } else {
    warning("GDB file does not exist for ", i, " at path: ", gdb_path)
  }
}


# Combine all shapefiles into one
combined_wetlands <- do.call(rbind, wetlands_list)


# Check spatial extent of wetlands data loaded in (the bounding box)
tmap_mode("view")
tm_shape(st_as_sfc(st_bbox(combined_wetlands)))+
  tm_borders()+
tm_shape(States_AOI)+
  tm_borders()

#Remove zipped geodatabase files
zip_wetlands
for (i in zip_wetlands) {
  file.remove(i)
  cat("Removed:", i, "\n")
}

```

### Join NWI attributes to sample locations:

```{r}

#Project locations into NAD83 Albers to match wetlands layer:
st_crs(combined_wetlands)
loc_NHDproj<-st_transform(loc_NHD, crs=st_crs(combined_wetlands))

Sys.time()
#Join selected wetland attributes to sample locations:
loc_NWI <- st_join(loc_NHDproj, combined_wetlands %>% dplyr::select(ATTRIBUTE, WETLAND_TYPE, ACRES), join = st_intersects)
Sys.time()

names(loc_NWI)

#View(loc_NWI)

#write intermediate sf object to file, to save all NWI info:
st_write(loc_NWI, paste0(dir.spatial, "/loc_NHD_NWI.shp"), delete_dsn = T)


```

# National Land Cover Database (NLCD)

Extract the NLCD value in the 30m cell where the sample was collected.

One option is to use get_nlcd() from the FedData package. The url this draws data from, www.mrlc.gov, can sometimes go down and be non functional.

Another option is to download NLCD data directly to the 'Spatial_data' folder. You can access the NLCD by state here: <https://datagateway.nrcs.usda.gov/GDGHome_DirectDownLoad.aspx>

Accessed 7.24.25

### Load NLCD data

**Option 1: using get_nlcd()**

Note: this function frequently encounters problems with then the mlrc site goes down - recommend using Option 2

```{r}
# Download the NLCD data for the specified AOI
#Note you are using 2019 NLCD data here (may want to adjust based on when data was collected):

#Sys.time()
#nlcd_data <- get_nlcd(template = States_AOI, label = "NLCD_AOI", year = 2019, dataset = "landcover")
#head(nlcd_data)
#Sys.time()
```

**Option 2: load after manually downloading nlcd rasters for each state to your project folder in a directory called 'nlcd'**

Note: if you run this for a new spatial extent, you will want to update the catched combined raster file (ie, merge it from scratch, not rely on the existing file)

```{r}

library(terra)
library(pbapply)
library(digest)   # for hashing file lists

# Full path to the nlcd folder
nlcd_dir <- file.path(dir.spatial, "nlcd")

# Find all .tif files in subfolders that start with "nlcd"
tif_files <- list.files(
  nlcd_dir,
  pattern = "^nlcd.*\\.tif$",
  recursive = TRUE,
  full.names = TRUE
)

tif_files

if (length(tif_files) == 0) stop("No rasters found in ", nlcd_dir)

#If the combined cache file (the merged raster) already exists (the raster was already written), use that; otherwise load rasters and merge)

combined_file <- list.files(
    nlcd_dir, 
    pattern="^nlcd_combined.*\\.tif$", 
    recursive=FALSE, 
    full.names=TRUE)
combined_file

if (file.exists(combined_file)) {
  # Load from cache
  message("Loading cached raster from: ", combined_file)
  
  nlcd_combined <- rast(combined_file)

} else {
  
  # Create a unique cache filename based on the rasters being used
  hash_val <- digest(tif_files)
  cache_file <- file.path(nlcd_dir, paste0("nlcd_combined_", hash_val, ".tif"))
  
  # Load rasters with progress bar
  message("Loading rasters...")
  
  nlcd_rasters <- pblapply(tif_files, rast)
  names(nlcd_rasters) <- NULL
 
#First project to same CRS
 # Target CRS - same as NWI (EPSG:6269)
  target_crs <- crs(combined_wetlands)

  # Reproject all rasters
  message("Projecting rasters to EPSG:6269...")
  
  nlcd_rasters <- pblapply(nlcd_rasters, function(r) {
    if (crs(r) != target_crs) {
      project(r, target_crs)
    } else {
      r
    }
  })
  
  # Merge if needed
  message("Combining rasters...")
  
  if (length(nlcd_rasters) == 1) {
    nlcd_combined <- nlcd_rasters[[1]]
    message("Only one raster found; skipping merge.")
  } else {
    nlcd_combined <- do.call(terra::merge, nlcd_rasters)
    message("Merged ", length(nlcd_rasters), " rasters.")
  }
 
  # Save to cache
  message("Saving merged raster to cache...")
  writeRaster(nlcd_combined, cache_file, overwrite = TRUE)
  message("Raster cached at: ", cache_file)
}


#check reference system
st_crs(nlcd_combined)
       

#Check map to make sure it looks OK
tmap_mode("view")

tm_shape(nlcd_combined)+
  tm_raster(col="NLCD_Land")+
  tm_shape(loc_NWI)+
  tm_dots(col="black")

```

### Join NLCD attributes to sample locations

```{r}

# Extract the raster value at the point location
loc_NLCD <- extract(nlcd_combined, loc_NWI)

#Bind NLCD attributes to other attributes:
loc_NLCD<-cbind(loc_NWI, loc_NLCD)

#Save intermediate attribute file: 
st_write(loc_NLCD, "./spatial_data/loc_NHD_NWI_NLCD.shp", append=FALSE)

#Look at attribute table containing NHD, NWI and NLCD attributes
#View(loc_NLCD)
```

# SSURGO dataset

Need to identify map unit key ('mukey') for each soil sample location. Will need ssurgo raster data for all states in study area.

Download and unzip state gSSURGO files from <https://www.nrcs.usda.gov/resources/data-and-reports/gridded-soil-survey-geographic-gssurgo-database>

see: "State Databases - soils - powered by Box"

Put unzipped geodatabase folders in "spatial_data/ssurgo" folder

If you want to look in more detail at soil series, NRCS has a query table by state where you can look at map unit soil series definitions:

<https://www.nrcs.usda.gov/publications/query-by-state.html>

### Load SSURGO data

**Load** **after manually downloading gSSURGO geodatabases for each state to your project folder in a directory called 'ssurgo'**

```{r}

#test for multiple states
#states<-c('IA','IL')

#otherwise assign states to study region
states<-list(States_sub$STUSPS)

# Read in ssurgo layer from every geodatabase (from each different state) in study region
  

# Full path to the ssurgo folder
ssurgo_dir <- file.path(dir.spatial, "ssurgo")


# Initialize a list to hold each state's ssurgo map unit layer
ssurgo_list <- list()

# Loop over state abbreviations and read each layer into the list

for (i in states) {
  gdb_path <- file.path(ssurgo_dir, paste0("gSSURGO_",i, ".gdb"))
  layer_name <- "MUPOLYGON"

  if (dir.exists(gdb_path)) {
    message("Reading layer for state: ", i)
    tryCatch({
      ssurgo_list[[i]] <- st_read(dsn = gdb_path, layer = layer_name)
    }, error = function(e) {
      warning("Failed to read ", i, ": ", e$message)
    })
  } else {
    warning("GDB file does not exist for ", i, " at path: ", gdb_path)
  }
}

head(ssurgo_list)

# Combine all shapefiles into one
combined_ssurgo <- do.call(rbind, ssurgo_list)

names(combined_ssurgo)

# Check spatial extent of ssurgo data loaded in (the bounding box)
tmap_mode("view")
tm_shape(st_as_sfc(st_bbox(combined_ssurgo)))+
  tm_borders()+
tm_shape(States_AOI)+
  tm_borders()

```

### Join SSURGO attributes to sample locations

```{r}

#Load in sample locations (if not already loaded)
#Read in table as needed: 
#Note that column names get abbreviated when you write the shapefile
loc_NLCD<-st_read("./spatial_data/loc_NHD_NWI_NLCD.shp")

names(loc_NLCD)

#Note slightly different CRS
#Get CRS to match exactly

loc_NLCD_NAD83ConusAlbers<-st_transform(loc_NLCD, crs=st_crs(combined_ssurgo))
crs(loc_NLCD_NAD83ConusAlbers)

#Join gSSURO mapunit attributes to soil sampl locations:
loc_ssurgo <- st_join(loc_NLCD_NAD83ConusAlbers, combined_ssurgo %>% dplyr::select(MUKEY, MUSYM), join = st_intersects)

nrow(loc_ssurgo)

View(loc_ssurgo)

#Write spatial data with all attribute keys to file:
st_write(loc_ssurgo, paste0(dir.spatial,"/loc_NHD_NWI_NLCD_gSSURGO.shp"), append=FALSE)

#Just pull out attributes:
loc_ssurgo_data<-loc_ssurgo %>% 
  st_drop_geometry() %>% 
  data.frame() 

head(loc_ssurgo_data)
write.csv(loc_ssurgo_data, paste0(dir.output, "./sample_attributes.csv"), row.names=FALSE)

```

# Add additional StreamCat and SSURGO attributes based on unique IDs (COMID and MUKEY)

Now that attribute keys for NHD, NWI, NLCD, gSSURGO and StreamCat have been added, you can follow the modules from Dolph et al 2023 to do add additional attributes from StreamCat and gSSURGO. This entails following modules 2 and 4 from the github repository (with some updates for coding efficiency): <https://github.com/cldolph/SoilP/>

Load sample points with all attributes attached so far:

```{r}

#set P sample file input name:

#For Streambank P:
#file.input<-"/sample_attributes"

#For soil P database:
file.input<-"/USGS_NCSS_soils_database_attributes_alldepths"

#Read in data:
P.att<-st_read(paste0(dir.output, file.input, '.csv'))
nrow(P.att)
head(P.att)
names(P.att)

#number of unique comids
P.att %>% 
  dplyr::select(comid) %>% 
  distinct() %>% 
  nrow()

```

## Get StreamCat attributes using StreamCatTools

The `StreamCatTools` package was designed to simplify the use of [StreamCat](https://www.epa.gov/national-aquatic-resource-surveys/streamcat-dataset) data in R, leveraging the new [API for StreamCat](https://api.epa.gov/StreamCat/streams/metrics).

There are different ways to request data: by area of interest, by state, or by a list of COMIDs.

Note 1: StreamCat attributes apply to two different scales: 1) catchment scale (called 'Cat') and 2) watershed scales (called 'Ws'). Each attribute will have 'Cat' or 'Ws' in the column name depending on which scale it applies to.

StreamCat Variable list can be seen here (however, note that not all the variables in the streamcat list are in this explainer variable list): <https://www.epa.gov/national-aquatic-resource-surveys/streamcat-metrics-and-definitions>

Some additional metadata on new Streamcat attributes: <https://pasteur.epa.gov/uploads/10.23719/1532282/documents/StreamCatNNI_readme.txt>

**Streamlining StreamCat request (for large datasets)**

```{r}

#Get a list of all Streamcat attributes available: 
All.Streamcat<-sc_get_params(param="metric_names")
All.Streamcat


#Subset the Streamcat attributes you want to use
#eliminate fire related, containing "burn", "sev", "fire", "procmask"
#remove veg change, "vegresp", forest loss "frstloss"
#select relevant NLCD year & eliminate all other years
#the following attributes are all based on NANI and are available every year, retain one of each variable for the year of interest, exclude all others: "n_ff, p_ff, n_uf,p_uf,p_hw,n_hw,n_lw,p_lw,n_cr,p_cr,n_cf,n_lwr,p_lwr,n_ags,p_ags"

All.Streamcat<-as.data.frame(All.Streamcat)

Streamcat.select<- All.Streamcat %>% 
  filter(!grepl('burn|sev|fire|procmask|vegresp|frstloss', All.Streamcat)) %>% 
  filter(!grepl('2001|2004|2006|2008|2013|2016|2019', All.Streamcat)) %>% 
  filter(!grepl( "_19|_20", All.Streamcat))
  
Streamcat.select

#Create list of select attributes in appropriate format for sc_get_data()
metric.list=paste0(Streamcat.select$All.Streamcat, collapse=",")
metric.list


#now get list of COMIDs for sample sites

#Original approach (this works fine for smallish datasets)
#comid.list=paste0(P.att$comid, collapse=",")
#comid.list

#Get streamcat attributes for this list of comids:
#sc_data<-sc_get_data(metric=metric.list, aoi='cat, ws', #comid=comid.list) #only a few seconds!


#For large data, a very large data request to sc_get_data appears to crash it out; 
#So split data into equal chunks and apply function, then combine (split, apply, combine)

# Define chunk size
chunk_size <- 300   # adjust if needed

# Split COMIDs into chunks
comid_chunks <- split(P.att$comid, 
                      ceiling(seq_along(P.att$comid) / chunk_size))
comid_chunks

# Iterate over chunks, call sc_get_data, and combine results
#(takes a few minutes)

sc_data_list <- lapply(comid_chunks, function(chunk) {
  comid.list <- paste0(chunk, collapse = ",")
  sc_get_data(metric = metric.list, aoi = "cat, ws", comid = comid.list)
})


# Combine all results into one dataframe
sc_data <- do.call(rbind, sc_data_list)

# Preview
head(sc_data)
View(sc_data)
nrow(sc_data) #Note that some values were dropped - these may be sink catchments (not associated with NHD flowline)
```

Merge StreamCat attributes by COMID to sampling location:

```{r}
P.streamcat<-merge(P.att, sc_data, by=c("comid"))
names(P.streamcat)
nrow(P.streamcat)
#View(P.streamcat)

#write to file:

write.csv(P.streamcat, paste0(dir.output, file.input %>% gsub(".csv", "", .), "_streamcat.csv"))
```

## Get SSURGO attributes by MUKEY

This part of the script uses ssurgo geodatabases downloaded to the project folder as specified above. Note downloading gSSURGO data first and then working with it will likely be faster than trying to pull data in from servers.

The following code is broken down by processing required for individual attribute tables in the gSSURGO dataset. At the end, attributes are subsetted and combined at the map unit scale.

Download and unzip state gSSURGO files from <https://www.nrcs.usda.gov/resources/data-and-reports/gridded-soil-survey-geographic-gssurgo-database>

see: "State Databases - soils - powered by Box"

Put unzipped geodatabase folders in "spatial_data/ssurgo" folder

### General Set Up

```{r}
#can ignore warnings about simple feature geometries not being present

options(warn=-1)

#Set directory for ssurgo data
ssurgo_dir<-paste0(dir.spatial, "/ssurgo")
ssurgo_dir

#ID states in study region
states<-as.list(States_sub$STUSPS)
states

#Read in soil sample locations, for file of interest

#Read in data:
P.streamcat<-read.csv(paste0(dir.output, file.input, '_streamcat', '.csv'))

nrow(P.streamcat)
```

### Read in all the SSURGO layers you will need:

Includes the following layers: "component", "chorizon", "chtexture", "chtexturegrp", "copm", "copmgrp", "Valu1"

```{r}

#Function to read one layer across all states
read_layer <- function(layer) {
  # Use map to create a list, then row-bind
  res_list <- map(states, ~ {
    gdb_path <- file.path(ssurgo_dir, paste0("gSSURGO_", .x, ".gdb"))
    message("Reading ", layer, " for state: ", .x)
    st_read(dsn = gdb_path, layer = layer, quiet = TRUE) %>%
      mutate(State = .x)
  })
  
  # Combine list into one data frame
  list_rbind(res_list)
}

# Read multiple layers into a named list
layers <- c("component", "chorizon", "chtexture", "chtexturegrp", "chtexturegrp", "copm", "copmgrp", "Valu1")

layer_data <- map(set_names(layers), read_layer)

```

### Further processing of ssurgo attributes at the map unit (mukey) scale

1.  Some SSURGO attributes are already summarized at the map unit scale. Others are summarized at scales smaller than the mapunit. To make everything consistent, attributes that occur at sub-map unit scales are combined (as weighted averages etc) to match the map unit scale.
2.  Attributes are also subsetted to desired ssurgo attributes - ie, those likely to be related to P content - as per Records et al 2016 <https://doi.org/10.1016/j.earscirev.2016.04.010>. Attributes ending in "\_r" are selected as these are considered "representative" (according to SSURGO manual).
3.  Note that the list of SSURGO attributes does include total P values (where available; most locations do not have total P values available)

#### Component table: different soil types in a map unit

'cokey' is the unique unit identifier for the component layer.

```{r}

#Create Component layer:
Component<-layer_data$component


#Check if component layers are unique --> they are
nrow(Component)

Component %>% 
  dplyr::select(cokey) %>% 
  distinct() %>% 
  nrow()


#Number of map units
Component %>% dplyr::select(mukey) %>% distinct() %>% nrow()


#Check map units against sample locations:
P.streamcat %>% 
  filter(!MUKEY %in% unique(Component$mukey))
#-->All sample mapunits are in component table

Component %>% 
  filter(mukey %in% P.streamcat$MUKEY) %>% 
  dplyr::select(cokey) %>% 
  distinct()
#-->are multiple components per map unit 

#Find mukey associated with specific cokey
Component %>% filter(mukey=="407496") %>% 
  dplyr::select(mukey, cokey)

##Subset to desired ssurgo attributes - ie, those likely to be related to P content (as per Records et al 2016) AND to mapunits that overlap with study sites

#Subset Component attributes 
names(Component)
Component.list<-c('comppct_r', 'slope_r', 'slopelenusle_r','runoff',
                  'erocl', 'hydricon', 'hydricrating', 'drainagecl', 'elev_r', 'geomdesc',
                  'airtempa_r', 'reannualprecip_r', 'ffd_r', 'taxceactcl', 'taxreaction', 'taxpartsize',
                  'mukey', 'cokey')


Component.sub<-Component %>% 
  dplyr::select(any_of(Component.list)) %>%  #select desired attributes
  filter(mukey %in% P.streamcat$MUKEY) %>%  #subset mapunits to those overlapping with sample sites
  dplyr::rename(MUKEY=mukey) #format MUKEY to match sample sites

```

#### Chorizon table (Soil chemistry and texture by horizon)

Chorizon does not have unique cokey. The multiple subcomponents within each component (cokey) for Chorizon are multiple soil horizons (at different depths).

-   hzdept_r - top depth of soil horizon

-   hzdepb-r - bottom depth of soil horizon

-   Note: SSURGO table metadata reports depth units as cm

Note: Originally, I sought to assign 1 horizon layer to each cokey, by doing the following:

-   eliminate samples from the very shallow surface that are likely to be plant material? nrow(Chorizon.surface) Chorizon.soil\<-Chorizon.surface %\>% filter(hzdepb_r\>3

-   choose uppermost surface layer, to avoid component attributes that include plant material at the surface rather than soil

However, in the module below, I retain all horizon depths for each cokey and then later, select the horizon depth in each cokey that most closely matches the depth of the actual soil sample

```{r}

#Create Chorizon layer
Chorizon<-layer_data$chorizon 


#check if chorizon layers map 1:1 with mukey 
nrow(Chorizon)

Chorizon %>% 
  dplyr::select(cokey) %>% 
  distinct() %>% 
  nrow()

#-->does not map 1:1



#Exploration of Chorizon table: 
View(Chorizon %>% 
       dplyr::select(cokey, hzname, hzdepb_r, hzdept_r) %>% 
      arrange(cokey, hzdepb_r))

#Look at the depth range of the uppermost horizon across all map units
#min.bottom.depth<-Chorizon %>% 
#    group_by(cokey) %>% 
#    dplyr::slice(which.min(hzdepb_r)) %>% 
#    dplyr::select(cokey, hzdepb_r, hzdept_r)

#View(min.bottom.depth)

#summary(min.bottom.depth$hzdepb_r)
#summary(min.bottom.depth$hzdept_r)

#Note that a minority of samples of the top horizon have a very deep depth; all horizons start at the surface (depth of 0)      

#OPTION: subset Chorizon data to shallowest layer in each cokey
#NOTE: this is NOT the approach I used for data manipulation below, where we have info on depth of soil samples

#Chorizon.surface<-Chorizon %>% 
#    group_by(cokey) %>% 
#    dplyr::slice(which.min(hzdepb_r)) %>% 
#    ungroup()

nrow(Chorizon)
nrow(min.bottom.depth)
#nrow(Chorizon.surface)


#does this still leave you with duplicates per site?
#(ie, multiple depths above 30cm)
#length(Chorizon.surface$cokey)
#length(unique(Chorizon.surface$cokey))

#Now only one record per cokey (the top most surface measurement)


#Check to see Chorizon measures for sample locations
# First make list of cokeys associated with map units of sample locations
cokey.list<-Component %>% 
  filter(mukey %in% unique(P.streamcat$MUKEY)) %>% 
  dplyr::select(cokey) 

nrow(cokey.list)

#some cokeys dropped when merge to Chorizon.surface
#These cokeys were never in the Chorizon table 
#They were not filtered out as the result of taking the minimum depth 
#I wonder if it could be related to points on stream line? (no Chorizon info?)


#Check for this list of cokeys in Chorizon.surface
View(Chorizon %>% 
  filter(cokey %in% unique(cokey.list$cokey)) %>% 
  distinct() %>% 
  dplyr::select(cokey, chkey, hzname, hzdepb_r))

Chorizon.cokey.list<-Chorizon %>% 
  dplyr::select(cokey) %>% 
  filter(cokey %in% unique(cokey.list$cokey)) %>% 
  distinct()


#Find difference between Component and Chorizon cokeys for sample locations:

Component %>% 
  filter(cokey %in% cokey.list$cokey) %>% 
  filter(!cokey %in% Chorizon.cokey.list$cokey) %>% 
  dplyr::select(cokey) %>% 
  distinct()

#Looking at the component/map unit names for these missing cokeys - they tend to be associated with water, flooded areas, or marsh areas 

#Look at the values that are not in Chorizon table:
Chorizon %>% 
  dplyr::select(cokey) %>% 
  filter(!cokey %in% unique(cokey.list$cokey)) %>% 
  distinct()
#These cokeys don't exist in the Chorizon table!
#Note: one map unit is in marsh/aquolls/ponded - i wonder if these don't have horizon values associated!

#Limit Chorizon to components that overlap with sample locations:
nrow(Chorizon)
Chorizon<-Chorizon %>% 
  filter(cokey %in% Component.sub$cokey)
nrow(Chorizon)

#Note: for now this will retain duplicate cokeys, for all horizon layers. Later we will subset horizons to those that match sample locations/depths

```

#### Chtexture & chtexturegrp table (Detailed soil texture)

Notes:

-   chkey - primary key in horizon table - unique identifier for each soil horizon

-   chtexture - contains info on soil texture (silt, clay, sand etc) for a given soil horizon

    -   there are different textures per horizon; for the different subcomponents - see <https://sdmdataaccess.nrcs.usda.gov/documents/ReturningSoilTextureRelatedAttributes.pdf>

-   chtkey - links soil texture (in chtexture) to soil horizon

-   chtexturegrp - contains complex texture group names

-   'rvindicator' column in chtexturegrp - from Table Column Descriptions: 'A yes/no field that indicates if a value is representative for the component'

-   Select rvindicator = Yes - leaves you with unique value, this is the 'representative' component

```{r}
#Create Chtexture Table (texture/grain size)
#Can be merged to Chorizon, & with Chtexturegrp

Chtexture<-layer_data$chtexture
Chtexturegrp<-layer_data$chtexturegrp

names(Chtexture)
names(Chtexturegrp)
#chtgkey is the common key
#Chtexturegrp has chkey (connects to Chorizon)

#Are all chkeys from Chorizon in Chtexture?
#names(Chorizon.surface)
#View(Chorizon.surface %>% 
#  filter(!chkey %in% Chtexturegrp$chkey) %>% 
#  dplyr::select(chkey, cokey))

#for some reason, a small # of rows in horizon that are not in texture group (I wonder if IA texture group data is not inclusive or something??)


Texture<-merge(Chtexture, Chtexturegrp %>% dplyr::select(!State), by=c('chtgkey'))
nrow(Texture)

#View(Texture)

#Join to Texture to Chorizon, uisng chkey

#first check if each chkey has a rvindicator=yes value?

Texture %>%
  group_by(chkey) %>%
  summarise(has_yes = any(rvindicator == "Yes")) %>%
  filter(!has_yes)
#--> yes, each chkey has at least one representative component     

#NOTE: previously had Chorizon.surface here: 

Horizon<-merge(Chorizon, Texture[Texture$rvindicator=="Yes",], by=c('chkey'))
#head(Horizon)
nrow(Chorizon)
nrow(Horizon)

#dropping some values here b/c of chkeys not in Chtexturegrp (relatively small #s)


#do I have duplicate measurements in Chtexture or Chtexturegrp?? (ie, subcomponets?)
length(Horizon$cokey)
length(unique(Horizon$cokey))

#Subset Horizon attributes 
names(Horizon)
Horizon.list<-c('cokey', 'hzdept_r', 'hzdepb_r','hzname', 'sieveno4_r','sieveno10_r','sieveno40_r','sieveno200_r',
                'sandtotal_r','silttotal_r','claytotal_r','claysizedcarb_r','om_r','ksat_r',
                'kffact','caco3_r','gypsum_r', 'freeiron_r','feoxalate_r','extral_r','aloxalate_r','pbray1_r', 'poxalate_r', 
                'partdensity', 'ph2osoluble_r', 'ptotal_r', 'texcl','texture', 'texdesc')

Horizon.sub<-Horizon %>% 
  dplyr::select(any_of(Horizon.list))
names(Horizon.sub)  
  
```

#### Copm and copmgrp (Soil Parent Origin)

```{r}

#turn warnings back on, now that all attribute tables are loaded
options(warn=0)

#Aggregate parent material from the start 

Parent<-merge(layer_data$copm, layer_data$copmgrp %>% dplyr::select(!State), by=c('copmgrpkey'))
head(Parent)
nrow(Parent)
#same number of rows as Copm; so was a 1:1 join

#check for duplicate cokeys
length(Parent$cokey)
length(unique(Parent$cokey))
#yes are duplicates, not 1:1 with cokey

#As for Texture, select rvindicator=Yes to avoid duplicates
names(Parent)
Parent2<-Parent[Parent$rvindicator=="Yes",]

#now check for duplicate cokeys
length(Parent2$cokey)
length(unique(Parent2$cokey))

#still have duplicates -
#find duplicates:
head(Parent2)
nrow(Parent2[duplicated(Parent2$cokey)|duplicated(Parent2$cokey, fromLast=TRUE),])
duplicates.Parent<-Parent2[duplicated(Parent2$cokey)|duplicated(Parent2$cokey, fromLast=TRUE),]
duplicates.Parent[1:20,]

#pmorder - the sequence in which the parent material occurs, when more than one parent material exists for one soil profile
#Note: will select pmorder ==1; assume this is the material closes to the surface in the soil profile
#Note that when there is only one parent material, no entry is required for pmorder, so also allow NA

Parent3<-Parent2[Parent2$pmorder=="1"|is.na(Parent2$pmorder),]
nrow(Parent3)

#now check for duplicate cokeys
length(Parent3$cokey)
length(unique(Parent3$cokey))
#still a smaller number of duplicates
duplicates.Parent3<-Parent3[duplicated(Parent3$cokey)|duplicated(Parent3$cokey, fromLast=TRUE),]

#View(duplicates.Parent3[1:20,])
#are multiple pmkind in the same order/pmgroupname
#pmgroupname seems to specify all possible combinations for each component
levels(factor(Parent3$pmgroupname)) #over 5000 levels

#Pick one pmkind for each horizon - 
#since they are all 'representative'
#this isn't perfect, but subsets 1 type to each horizon
Parent4 <- Parent3[!duplicated(Parent3$cokey),]
nrow(Parent3)
nrow(Parent4)
length(Parent4$cokey)
length(unique(Parent4$cokey))
#Now there are no duplicates for cokey

#Subset Parent attributes
names(Parent4)
head(Parent4)
nrow(Parent4)
Parent4.list<-c('cokey', 'pmkind', 'pmorigin')

Parent.sub<-Parent4 %>% 
  dplyr::select(any_of(Parent4.list))

```

#### Valu1 Table

These are commonly used attributes already aggregated in SSURGO at the map unit scale

```{r}
valu1<-layer_data$Valu1

names(valu1)
#rename MUKEY column
valu1<-valu1 %>% 
  dplyr::rename(MUKEY=mukey)
```

### Merge SSURGO attributes to the map unit scale.

Merge everything to mukey; using cokey

Need comppct from component table for % contribution of each component to each mukey (see below)

NOTE: not all cokeys associated with samplie locations have attributes in other tables, so use all.x=TRUE to keep all sample locations

```{r}

#COMPLETE LIST OF SSURGO TABLES NEEDED:
names(Component) #component unit (cokey)
names(Horizon) #component unit (cokey)
names(Parent4) #component unit (cokey)
names(valu1) #map unit (mukey)
```

```{r}

#Merge Component and Horizon
Comp1<-merge(Component.sub, Horizon.sub, by=c('cokey'), all.x=TRUE)
head(Comp1)
nrow(Comp1)
names(Comp1)

View(Comp1)

#Check that mukeys for sample locations were included
if(nrow(P.streamcat)==
P.streamcat %>% 
  filter(MUKEY %in% unique(Comp1$mukey)) %>% 
  nrow()
)
{message("yes all sample locations have mapunit data after merging with horizon table")} else{
  message("some mapunits missing")
}



#check for duplicates
length(Comp1$cokey)
length(unique(Comp1$cokey))


#Merge in Parent table
Comp2<-merge(Comp1, Parent.sub, by=c('cokey'))
names(Comp2)
head(Comp2)
nrow(Comp2)
#check for duplicates
length(Comp2$cokey)
length(unique(Comp2$cokey))


#Check that mukeys for sample locations were included
if(nrow(P.streamcat)==
P.streamcat %>% 
  filter(MUKEY %in% unique(Comp2$mukey)) %>% 
  nrow()
)
{message("yes all sample locations have mapunit data after merging with Parent table")} else{
  message("some mapunits missing")
}


```

### Summarize/aggregate gSSURGO attributes at the map unit scale

Steps:

1.  Check column types to make sure each is designated as correct type (either numeric or factor).
2.  Simplify/reclassify categorical attributes with many levels.
3.  Summarize component scale attributes at the map unit scale.
4.  For numeric variables, take weighted average for each map unit.
5.  For categorical variable, take component with highest % of map unit for each sample. (For ties, the first row in each group is arbitrarily selected)

Weights are 'comppct_r' - i.e., the percent of each component in the map unit (the mukey)

**Check attribute type and reclassify attributes as needed:**

```{r}

sapply(Comp2, class)
head(Comp2)
names(Comp2)

#Assign correct class to each attribute

#ID numeric variables
Numeric.list <- c(
  "slope_r", "slopelenusle_r", "elev_r", "airtempa_r", "reannualprecip_r",
  "ffd_r", "sieveno4_r", "sieveno10_r", "sieveno40_r", "sieveno200_r",
  "sandtotal_r", "silttotal_r", "claytotal_r", "claysizedcarb_r", "om_r",
  "partdensity", "ksat_r", "kffact", "caco3_r", "gypsum_r", "freeiron_r",
  "feoxalate_r", "extral_r", "aloxalate_r", "pbray1_r", "poxalate_r",
  "ph2osoluble_r", "ptotal_r", "comppct_r"
)


#ID categorical variables
Categorical.list <- c("runoff", "erocl", "hydricon", "hydricrating", "drainagecl", "taxpartsize", "taxceactcl", "taxreaction", "texcl", "pmkind", "pmorigin", "texdesc", "texture", "MUKEY", "hzname")


#For categorical variables
#check levels for each categorical variables
#note - random forest models can only handle categorical variables with a certain number of levels (53)

levels(factor(Comp2$runoff)) # 6 levels
levels(factor(Comp2$erocl)) # 5 levels
levels(factor(Comp2$hydricon)) # 3 levels
levels(factor(Comp2$hydricrating)) # 3 levels
levels(factor(Comp2$drainagecl)) # 8 levels
levels(factor(Comp2$geomdesc)) #many levels!! -->Exclude
levels(factor(Comp2$taxpartsize)) #many levels --> Combine? Or Exclude;
#^^might be represented by other variables, e.g., % clay
levels(factor(Comp2$taxceactcl)) #5 levels
levels(factor(Comp2$taxreaction)) #7 levels
levels(factor(Comp2$texcl)) #many levels, could substitute for taxpartsize maybe
levels(factor(Comp2$pmkind)) #many levels -->could aggregate some levels
levels(factor(Comp2$pmorigin)) 

#Variables to exclude or modify (eg, aggregate or code dummy variables):
#geomdesc - too many levels


#Transform numeric and categorical attributes as appropriate:
Comp2 <- Comp2 %>%
  mutate(across(all_of(Numeric.list), as.numeric)) %>% 
  mutate(across(all_of(Categorical.list), as.factor)) %>% 
  mutate(kffact=as.numeric(kffact), hzname=as.factor(hzname), MUKEY=as.factor(MUKEY), pmorigin=as.factor(pmorigin), pmkind=as.factor(pmkind), texdesc=as.factor(texdesc), texture=as.factor(texture), texcl=as.factor(texcl), taxpartsize=as.factor(taxpartsize), taxreaction=as.factor(taxreaction), taxceactcl=as.factor(taxceactcl))

summary(Comp2)




```

**Summarize Attributes:**

For each soil sample, calculate the weighted average of numeric attributes at the map unit scale that occur at the component scale. Select the categorical attributes with the highest component area.

Note that it is possible to have more than one sample at the same location but at different sample depths. Thus, samples are also assigned the corresponding most closely matching horizon depths from the Chorizon table in SSURGO prior to calculating weighted averages. To do this, before summarizing component attributes, assign horizon depth to sample sites. This should result in the assignment of one of *each* the cokeys in a map unit (MUKEY) to each sample. (Otherwise the sample will be assigned to multiple horizons).

```{r}

#Steps:
#Join SSURGO attributes to sample data (P.streamcat) based on mukey
#then select rows where actual measured Depth_cm is between Chorizon table hzdepth values
#hzdepb_r = bottom depth of a horizon
#hzdept_r = top depth of a horizon 


#Note, must use allow.cartersian=TRUE or otherwise get an error, because of the one to many join
P.ssurgo1<-merge(P.streamcat, Comp2, by=c("MUKEY"), allow.cartesian=TRUE)

#filter if Depth of soil sample is between hz levels
P.ssurgo.depth<-P.ssurgo1 %>%
  filter(Depth_cm <= hzdepb_r & Depth_cm >= hzdept_r)

#Check that it worked for a samples in a select mapunit: 
View(P.ssurgo.depth %>% dplyr::select(MUKEY, cokey, SampleID, comppct_r, hzname, hzdept_r, hzdepb_r, Depth_cm, kffact) %>% filter(MUKEY=="100040"))



#For numeric attributes - 
#calculate weighted mean for each sample in each map unit (MUKEY). Weights are % contribution of each component (comppct_r) to the total map unit area. 
#calculate for multiple attributes at once

Mukey.mean<-P.ssurgo.depth %>% group_by(MUKEY, SampleID) %>% 
  summarise_at(Numeric.list, 
               list(~weighted.mean(., comppct_r)))
names(Mukey.mean)

#Check result:
View(Mukey.mean %>% dplyr::select(MUKEY, SampleID, kffact, comppct_r) %>% filter(MUKEY=="100040"))


#For categorical attributes - 
#select component level with the highest percentage area in the map unit, for each sample

#group by mukey and sample ID
#apply just across columns in Categorical.list
#select attribute levels in each group based on max compcct_r
#note this will just pick the first value if there is a tie:
#(somewhat arbitrary)

Mukey.top_cat<-P.ssurgo1 %>% 
  dplyr::select(MUKEY, SampleID, comppct_r, all_of(Categorical.list)) %>%
  group_by(MUKEY, SampleID) %>% 
  dplyr::slice(which.max(comppct_r))

# View the result
View(Mukey.top_cat %>% dplyr::select(MUKEY, SampleID, pmorigin, comppct_r) %>% filter(MUKEY=="100040"))

##So now you have:
#weighted average for numeric variables for each sample site in each map unit
#categorical attribute with highest percentage area for each sample site in map unit

```

Merge numeric and categorical SSURGO attributes to all other attributes and write to file:

```{r}
#Merge the two;
SSURGO.att<-merge(Mukey.mean, Mukey.top_cat, by=c('MUKEY', 'SampleID'))
head(SSURGO.att)
#View(SSURGO.att)

#Add in attributes from Valu1 table
SSURGO.all<-merge(SSURGO.att, valu1, by=c('MUKEY'))
head(SSURGO.all)

#Check that mukeys for sample locations were included
if(nrow(P.streamcat)==
P.streamcat %>% 
  filter(MUKEY %in% unique(SSURGO.all$mukey)) %>% 
  nrow()
)
{message("yes all sample locations have mapunit data after merging with valu1 table")} else{
  message("some mapunits missing")
}

#Note: some mukeys may have been omitted if they were not in all SSURGO tables - this seems to occur for map units associated with water areas


```

### Merge ssurgo attributes to sample locations

```{r}

#Merge back to all other attributes previously added for sample locations:
soilP.att<-merge(P.streamcat, SSURGO.all, by=c("SampleID", "MUKEY"))
View(soilP.att)


#write to table;
#these are unique attributes for each map unit, from multiple SSURGO tables

write.table(soilP.att, paste0(dir.output, file.input,'_SSURGO_attributes_mapunit.csv'), sep=",", row.names=FALSE)

```

# Clean up all attributes -

Sample datasets have had the following pre-processing:

-   NHD stream type from NHDv2Plus

-   NWI wetland type, if within wetland polygon

-   NLCD 2019 Land Use category in 30m cell where sample occurs

-   EPA StreamCat attributes

-   gSSURGO attributes

Additional processing below:

-   remove extraneous attributes (not likely applicable to study area)

-   remove attributes with no information (all 0 or all NA)

## Remove extraneous attributes

```{r}

names(soilP.att)

#Remove administrative attributes: "MUKEY", "comid", "MUSYM", "ID", 'State', 'dataset', 'ACRES', 'field_1', 'SampleID'

soilP.less<-
soilP.att %>% 
  dplyr::select(-contains(c('State', 'SampleID', 'MUKEY', 'comid', 'MUSYM', 'ACRES', 'ID', 'dataset', 'Site', 'replicates', 'field_1'))) %>% 
  dplyr::select(-starts_with("X", ignore.case = FALSE))

ncol(soilP.att)
ncol(soilP.less)

names(soilP.less)

#View(soilP.less)
```

## Remove columns with no information

```{r}

#Check for columns where all values are 0 or NA - will need to assign numeric vs categorical 
#Check for missing values

# Identify columns where all values are NA or 0
cols_to_remove <- soilP.less %>%
  dplyr::select(where(~ all(is.na(.) | . == 0))) %>%
  names()

# Print the names of the identified columns
print(cols_to_remove)

#Note! No info about total P for these samples

#Remove columns were all are NA

soilP.data<-soilP.less %>% 
  dplyr::select(-contains (cols_to_remove))

names(soilP.data)
```

## Combine replicate streambank measurements (if applicable)

Note: if you have depth measurements for streambank samples and you want to retain this as a predictor, you could try to keep separate. But also note that so far ssurgo attributes have been selected for the surface layer.

**If no replicates, just do the following:**

```{r}
P.data<-soilP.data
```

**If replicates:**

```{r}

soilP.avg<-soilP.data %>% 
  group_by(x,y) %>% 
  dplyr::summarise(meanP_mgkg = mean(P_mgkg), meanBD_gcm3=mean(BD_gcm3)) %>% 
      ungroup()

#Merge back to whole dataset

P.data<-merge(soilP.data, soilP.avg, by=c("x", "y"))

P.data<-P.data %>% 
  dplyr::select(-c(P_mgkg, BD_gcm3)) %>% 
  distinct()


View(P.data)
```

## Create non wetland category for wetland attributes

```{r}

P.data<-P.data %>% 
  mutate(WETLAND_TYPE=ifelse(is.na(WETLAND_TYPE), "Non wetland", WETLAND_TYPE)) %>% 
  mutate(WETLAND_ATTRIBUTE=ifelse(is.na(ATTRIBUTE), "Non wetland", ATTRIBUTE)) %>% 
  dplyr::select(-ATTRIBUTE)

#View(soilP.avg)
```

## Write final data to file

```{r}
#Write to file

write.table(P.data, paste0(dir.output, file.input, '_for_modelling.csv'), sep=",", row.names=FALSE)
```
