---
title: "Assigning spatial attributes to sample locations"
author: "Christy Dolph"
date: "2024-07-17"
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Overview: Generating spatial attributes for sites of interest

This script generates spatial attributes associated with specific point locations.

Attributes are from the following sources:

-   the NHDPlusv2 stream network (NHD)

-   the National Wetlands Inventory (NWI),

-   National Land Cover Dataset (NLCD)

-   gSSURGO soils data

-   EPA StreamCat

This approach is updated and adapted from the scripts available at: <https://github.com/cldolph/SoilP>

## Steps:

1.  Merge locations to NHD, NLCD, and NWI attributes + SSURGO map unit key (mukey)
2.  Compile and process SSURGO attributes and merge
3.  Compile and process StreamCat attributes and merge

# National Hydrography Dataset (NHD)

Use nhdplusTools package to link sample locations to NHD COMID (unique identifier for each stream reach of the NHD)

The COMID can then be used to link the location to NHD attributes (stream type, stream order, etc) as well as EPA StreamCat attributes.

Note that 'featureid' of an NHD Catchment is equal to COMID of an NHD flowline; see p. 47 of the NHDPlusV2 User Guide:

[https://www.epa.gov/system/files/documents/2023-04/NHDPlusV2_User_Guide.pdf](https://www.epa.gov/system/files/documents/2023-04/NHDPlusV2_User_Guide.pdf,)

### Download flowlines for NHD stream network

```{r}

#Get NHD data for HUC4s of interest via web service, using nhdplusTools:

NHD_HUC4<-get_nhdplus(
  AOI = HUC4_AOI)

names(NHD_HUC4)

st_crs(NHD_HUC4) #check reference system

#plot NHD flowlines (note this can take several minutes)
#tm_shape(NHD_HUC4)+
 # tm_lines()

View(head(NHD_HUC4))

```

### Identify NHD COMIDs associated with sample locations

```{r}

#first transform pour point and flowlines to same projection:
#continental US (Conus Albers NAD83)

loc_trans <- st_transform(loc_sf, 5070)
NHD_trans<-st_transform(NHD_HUC4, 5070)

Sys.time()
loc_snap <- st_snap(loc_trans, NHD_trans, tolerance = 100)
Sys.time() #4 min for IA sites

#Quick plot check
plot(loc_snap$geometry)


#The function discover_nhdplus_id() can pull comids for single point locations; to get comids for all points you must loop this function over every point

# Initialize a vector to store COMIDs
comids <- numeric(nrow(loc_snap))

# Loop over each row (point)
for (i in seq_len(nrow(loc_snap))) {
  # Extract a single-point sf object
  single_point <- loc_snap[i, ]
  
  # Try running discover_nhdplus_id()
  # Use tryCatch to prevent errors from stopping the loop
  comids[i] <- tryCatch({
    discover_nhdplus_id(single_point)
  }, error = function(e) {
    NA  # return NA if there's an error
  })
}

# Add the COMIDs back to the sf object
loc_snap$comid <- comids

#View(loc_snap)

```

### Join NHD attributes to sampled locations, based on COMID

see NHDv2Plus User Guide: <https://nhdplus.com/NHDPlus/NHDPlusV2_documentation.php>

'totdasqkm' = Total Upstream Cumulative Drainage Area, in square kilometers, at the downstream end of the NHDFlowline feature

```{r}
nrow(loc_snap)

loc_NHD<-merge(loc_snap, NHD_HUC4 %>% dplyr::select(comid, gnis_name, lengthkm, ftype, fcode, streamleve, streamorde, totdasqkm) %>% st_drop_geometry(), by=c("comid"))

nrow(loc_NHD)
#View(loc_NHD)

```

# National Wetlands Inventory (NWI)

See access to NWI data at: <https://www.fws.gov/program/national-wetlands-inventory/data-download>

Download data by state: <https://www.fws.gov/program/national-wetlands-inventory/download-state-wetlands-data>

### Load in wetlands data:

```{r}

#US states where samples occur (corresponding to study area):
states<-list(States_sub$STUSPS)
states


##To download NWI files for every state in the your study area of interest, loop GET url for multiple states:

# Base URL where NWI data is available
base_url <- "https://documentst.ecosphere.fws.gov/wetlands/data/State-Downloads"

# Loop through state abbreviations and download corresponding zip files

for (i in states) {
  zip_name <- paste0(i, "_geodatabase_wetlands.zip")
  url <- paste0(base_url, "/", zip_name)
  destfile <- file.path("./spatial_data", zip_name)
  
  if (!file.exists(destfile)) {
    message("Downloading: ", url)
    tryCatch({
      response <- GET(url, write_disk(destfile, overwrite = TRUE),
                      user_agent("Mozilla/5.0 (Windows NT 10.0; Win64; x64)"))
      if (response$status_code != 200) {
        warning("Download failed for ", states, " with status ", response$status_code)
      }
    }, error = function(e) {
      warning("Failed to download ", states, ": ", e$message)
    })
  } else {
    message("Already downloaded: ", states)
  }
}

# Unzip geodatabase(s)

#List of zipped files:
zip_wetlands <- list.files("./spatial_data", pattern = "wetlands.*\\.zip$", full.names = TRUE)
zip_wetlands

#Unzip files to spatial data directory:
ldply(.data = zip_wetlands, .fun = unzip, exdir = "./spatial_data")


##Read in wetlands layer from every geodatabase (from each different state) in study region
  

# Initialize a list to hold each state's wetlands layer
wetlands_list <- list()
wetlands_list
# Loop over state abbreviations and read each layer into the list

for (i in states) {
  gdb_path <- file.path("spatial_data", paste0(i, "_geodatabase_wetlands.gdb"))
  layer_name <- paste0(i, "_Wetlands")

  if (dir.exists(gdb_path)) {
    message("Reading layer for state: ", i)
    tryCatch({
      wetlands_list[[i]] <- st_read(dsn = gdb_path, layer = layer_name)
    }, error = function(e) {
      warning("Failed to read ", i, ": ", e$message)
    })
  } else {
    warning("GDB file does not exist for ", i, " at path: ", gdb_path)
  }
}


# Combine all shapefiles into one
combined_wetlands <- do.call(rbind, wetlands_list)


# Check spatial extent of wetlands data loaded in (the bounding box)
tmap_mode("view")
tm_shape(st_as_sfc(st_bbox(combined_wetlands)))+
  tm_borders()+
tm_shape(States_AOI)+
  tm_borders()

#Remove zipped geodatabase files
zip_wetlands
for (i in zip_wetlands) {
  file.remove(i)
  cat("Removed:", i, "\n")
}

```

### Join NWI attributes to sample locations:

```{r}

#Project locations into NAD83 Albers to match wetlands layer:
st_crs(combined_wetlands)
loc_NHDproj<-st_transform(loc_NHD, crs=st_crs(combined_wetlands))

Sys.time()
#Join selected wetland attributes to sample locations:
loc_NWI <- st_join(loc_NHDproj, combined_wetlands %>% dplyr::select(ATTRIBUTE, WETLAND_TYPE, ACRES), join = st_intersects)
Sys.time()

names(loc_NWI)

#View(loc_NWI)

#write intermediate sf object to file, to save all NWI info:
st_write(loc_NWI, paste0(dir.spatial, "/loc_NHD_NWI.shp"), delete_dsn = T)


```

# National Land Cover Database (NLCD)

Extract the NLCD value in the 30m cell where the sample was collected.

One option is to use get_nlcd() from the FedData package. The url this draws data from, www.mrlc.gov, can sometimes go down and be non functional.

Another option is to download NLCD data directly to the 'Spatial_data' folder. You can access the NLCD by state here: <https://datagateway.nrcs.usda.gov/GDGHome_DirectDownLoad.aspx>

Accessed 7.24.25

### Load NLCD data

**Option 1: using get_nlcd()**

Note: this function frequently encounters problems with then the mlrc site goes down - recommend using Option 2

```{r}
# Download the NLCD data for the specified AOI
#Note you are using 2019 NLCD data here (may want to adjust based on when data was collected):
#Sys.time()
#nlcd_data <- get_nlcd(template = States_AOI, label = "NLCD_AOI", year = 2019, dataset = "landcover")
#head(nlcd_data)
#Sys.time()
```

**Option 2: load after manually downloading nlcd rasters for each state to your project folder in a directory called 'nlcd'**

```{r}

library(terra)
library(pbapply)
library(digest)   # for hashing file lists

# Full path to the nlcd folder
nlcd_dir <- file.path(dir.spatial, "nlcd")

# Find all .tif files in subfolders that start with "nlcd"
tif_files <- list.files(
  nlcd_dir,
  pattern = "^nlcd.*\\.tif$",
  recursive = TRUE,
  full.names = TRUE
)

tif_files

if (length(tif_files) == 0) stop("No rasters found in ", nlcd_dir)

# Create a unique cache filename based on the rasters being used
hash_val <- digest(tif_files)
cache_file <- file.path(nlcd_dir, paste0("nlcd_combined_", hash_val, ".tif"))

#If the cache file already exists (the raster was already written, use that; otherwise load rasters and merge)

if (file.exists(cache_file)) {
  # Load from cache
  message("Loading cached raster from: ", cache_file)
  
  nlcd_combined <- rast(cache_file)

} else {
  # Load rasters with progress bar
  message("Loading rasters...")
  
  nlcd_rasters <- pblapply(tif_files, rast)
  names(nlcd_rasters) <- NULL
 
#First project to same CRS
 # Target CRS - same as NWI (EPSG:6269)
  target_crs <- crs(combined_wetlands)

  # Reproject all rasters
  message("Projecting rasters to EPSG:6269...")
  
  nlcd_rasters <- pblapply(nlcd_rasters, function(r) {
    if (crs(r) != target_crs) {
      project(r, target_crs)
    } else {
      r
    }
  })
  
  # Merge if needed
  message("Combining rasters...")
  
  if (length(nlcd_rasters) == 1) {
    nlcd_combined <- nlcd_rasters[[1]]
    message("Only one raster found; skipping merge.")
  } else {
    nlcd_combined <- do.call(terra::merge, nlcd_rasters)
    message("Merged ", length(nlcd_rasters), " rasters.")
  }
 
  # Save to cache
  message("Saving merged raster to cache...")
  writeRaster(nlcd_combined, cache_file, overwrite = TRUE)
  message("Raster cached at: ", cache_file)
}


#check reference system
st_crs(nlcd_combined)
       

#Check out on a map
names(nlcd_combined)
View(head(nlcd_combined))


#Check map to make sure it looks OK
tmap_mode("view")

tm_shape(nlcd_combined)+
  tm_raster(col="NLCD_Land")+
  tm_shape(loc_NWI)+
  tm_dots(col="black")

```

### Join NLCD attributes to sample locations

```{r}

# Extract the raster value at the point location
loc_NLCD <- extract(nlcd_combined, loc_NWI)

#Bind NLCD attributes to other attributes:
loc_NLCD<-cbind(loc_NWI, loc_NLCD)

#Save intermediate attribute file: 
st_write(loc_NLCD, "./spatial_data/loc_NHD_NWI_NLCD.shp", append=FALSE)

#Look at attribute table containing NHD, NWI and NLCD attributes
#View(loc_NLCD)
```

# SSURGO dataset

Need to identify map unit key ('mukey') for each soil sample location. Will need ssurgo raster data for all states in study area.

Download and unzip state gSSURGO files from <https://www.nrcs.usda.gov/resources/data-and-reports/gridded-soil-survey-geographic-gssurgo-database>

see: "State Databases - soils - powered by Box"

Put unzipped geodatabase folders in "spatial_data/ssurgo" folder

If you want to look in more detail at soil series, NRCS has a query table by state where you can look at map unit soil series definitions:

<https://www.nrcs.usda.gov/publications/query-by-state.html>

### Load SSURGO data

**Load** **after manually downloading gSSURGO geodatabases for each state to your project folder in a directory called 'ssurgo'**

```{r}

#test for multiple states
#states<-c('IA','IL')

#otherwise assign states to study region
states<-list(States_sub$STUSPS)

# Read in ssurgo layer from every geodatabase (from each different state) in study region
  

# Full path to the ssurgo folder
ssurgo_dir <- file.path(dir.spatial, "ssurgo")


# Initialize a list to hold each state's ssurgo map unit layer
ssurgo_list <- list()

# Loop over state abbreviations and read each layer into the list

for (i in states) {
  gdb_path <- file.path(ssurgo_dir, paste0("gSSURGO_",i, ".gdb"))
  layer_name <- "MUPOLYGON"

  if (dir.exists(gdb_path)) {
    message("Reading layer for state: ", i)
    tryCatch({
      ssurgo_list[[i]] <- st_read(dsn = gdb_path, layer = layer_name)
    }, error = function(e) {
      warning("Failed to read ", i, ": ", e$message)
    })
  } else {
    warning("GDB file does not exist for ", i, " at path: ", gdb_path)
  }
}

head(ssurgo_list)

# Combine all shapefiles into one
combined_ssurgo <- do.call(rbind, ssurgo_list)

names(combined_ssurgo)

# Check spatial extent of ssurgo data loaded in (the bounding box)
tmap_mode("view")
tm_shape(st_as_sfc(st_bbox(combined_ssurgo)))+
  tm_borders()+
tm_shape(States_AOI)+
  tm_borders()

```

### Join SSURGO attributes to sample locations

```{r}

#Load in sample locations (if not already loaded)
#Read in table as needed: 
#Note that column names get abbreviated when you write the shapefile
loc_NLCD<-st_read("./spatial_data/loc_NHD_NWI_NLCD.shp")

names(loc_NLCD)

#Note slightly different CRS
#Get CRS to match exactly

loc_NLCD_NAD83ConusAlbers<-st_transform(loc_NLCD, crs=st_crs(combined_ssurgo))
crs(loc_NLCD_NAD83ConusAlbers)

#Join gSSURO mapunit attributes to soil sampl locations:
loc_ssurgo <- st_join(loc_NLCD_NAD83ConusAlbers, combined_ssurgo %>% dplyr::select(MUKEY, MUSYM), join = st_intersects)

nrow(loc_ssurgo)

View(loc_ssurgo)

#Write spatial data with all attribute keys to file:
st_write(loc_ssurgo, paste0(dir.spatial,"/loc_NHD_NWI_NLCD_gSSURGO.shp"), append=FALSE)

#Just pull out attributes:
loc_ssurgo_data<-loc_ssurgo %>% 
  st_drop_geometry() %>% 
  data.frame() 

head(loc_ssurgo_data)
write.csv(loc_ssurgo_data, paste0(dir.output, "./sample_attributes.csv"), row.names=FALSE)

```

# Add additional StreamCat and SSURGO attributes based on unique IDs (COMID and MUKEY)

Now that attribute keys for NHD, NWI, NLCD, gSSURGO and StreamCat have been added, you can follow the modules from Dolph et al 2023 to do add additional attributes. This entails following modules 2 and 4 from the github repository: <https://github.com/cldolph/SoilP/>

Load sample points with all attributes attached so far:

```{r}

P.att<-st_read(paste0(dir.output, "/sample_attributes.csv"))

head(P.att)
names(P.att)

```

## Get StreamCat attributes by state using StreamCatTools

The `StreamCatTools` package was designed to simplify the use of [StreamCat](https://www.epa.gov/national-aquatic-resource-surveys/streamcat-dataset) data in R, leveraging the new [API for StreamCat](https://api.epa.gov/StreamCat/streams/metrics).

There are different ways to request data. One way is by state. The code below pulls StreamCat attributes from the API by state. If you have a smaller study area, it will be faster to pull for a different spatial unit (i.e., a watershed) or for a list of COMIDs, but the options in StreamCatTools are somewhat limited for larger study areas. So going with this for now.

Note: StreamCat attributes apply to two different scales: 1) catchment scale (called 'Cat') and 2) watershed scales (called 'Ws'). Each attribute will have 'Cat' or 'Ws' in the column name depending on which scale it applies to.

```{r}

#Request StreamCat data by state: 

#test for multiple states
#states<-c('IA','IL')
#otherwise assign states to study region
states<-list(States_sub$STUSPS)

#Check if StreamCat data has already been downloaded; 
#if yes, then load it 
#if not, then download from server

streamcat<-paste0(dir.output, "StreamCat.csv")

if (file.exists(streamcat)) {
  # Load from cache
  message("Loading StreamCat data from local folder: ", streamcat)
  
  sc_data_all <- read.csv(streamcat)

} else {

#Request attributes at the catchment scale (can take a long time, ~26 min for 2 states): 
#read all attributes from streamcat data into a list, where each item in the list is all the data for an individual state from the "states" list

  message("Loading catchment-scale data from StreamCat API...")
sc_data <- map(set_names(states), ~ sc_get_data(
  aoi = "cat",
  metric = "all",
  state = .x
))

# Combine into one data frame, keeping state abbreviations
sc_data_combined <- bind_rows(sc_data, .id = "State")

#Request attributes at the watershed scale:  

  message("Loading watershed-scale data from StreamCat API...")
sc_data_WS <- map(set_names(states), ~ sc_get_data(
  aoi = "ws",
  metric = "all",
  state = .x
))

# Combine into one data frame, keeping state abbreviations
sc_data_combined_WS <- bind_rows(sc_data_WS, .id = "State")

  message("Combining catchment and watershed scale attributes")
#Merge catchment and watershed scale attributes
sc_data_all<-merge(sc_data_combined, sc_data_combined_WS %>% dplyr::select(!State), by=c("comid"))

#write to file so don't have to reload every time:
message("Writing StreamCat data to file...")
write.csv(sc_data_all, paste0(dir.output, "/StreamCat.csv"), row.names=FALSE)
}

# Check for redundant columns - 
# find column names containing ".x" or ".y"
all_col_names<-colnames(sc_data_all)
matching_col_names <- all_col_names[grepl("\\.x|\\.y", all_col_names)]

print(matching_col_names)

```

Merge StreamCat attributes by COMID to sampling location:

```{r}
P.streamcat<-merge(P.att, sc_data_all, by=c("comid"))
#View(P.streamcat)

#write to file:

write.csv(P.streamcat, paste0(dir.output, "/sample_attributes_streamcat.csv"))
```

## Get SSURGO attributes by MUKEY

This part of the script uses ssurgo geodatabases downloaded to the project folder as specified above. Note downloading gSSURGO data first and then working with it will likely be faster than trying to pull data in from servers.

The following code is broken down by processing required for individual attribute tables in the gSSURGO dataset. At the end, attributes are subsetted and combined at the map unit scale.

### General Set Up

```{r}
#can ignore warnings about simple feature geometries not being present

options(warn=-1)

#Set directory for ssurgo data
ssurgo_dir<-paste0(dir.spatial, "/ssurgo")
ssurgo_dir

#test for multiple states
#states<-c('IA','IL')
#otherwise assign states to study region
states<-list(States_sub$STUSPS)

```

### Read in all the SSURGO layers you will need:

Includes the following layers: "component", "chorizon", "chtexture", "chtexturegrp", "copm", "copmgrp", "Valu1"

```{r}

#Function to read one layer across all states
read_layer <- function(layer) {
  # Use map to create a list, then row-bind
  res_list <- map(states, ~ {
    gdb_path <- file.path(ssurgo_dir, paste0("gSSURGO_", .x, ".gdb"))
    message("Reading ", layer, " for state: ", .x)
    st_read(dsn = gdb_path, layer = layer, quiet = TRUE) %>%
      mutate(State = .x)
  })
  
  # Combine list into one data frame
  list_rbind(res_list)
}

# Read multiple layers into a named list
layers <- c("component", "chorizon", "chtexture", "chtexturegrp", "chtexturegrp", "copm", "copmgrp", "Valu1")

layer_data <- map(set_names(layers), read_layer)

```

### Further processing of ssurgo attributes at the map unit (mukey) scale

Some SSURGO attributes are already summarized at the map unit scale. Others are summarized at scales smaller than the mapunit. To make everything consistent, attributes that occur at sub-map unit scales are combined (as weighted averages etc) to match the map unit scale.

#### Component table: different soil types in a map unit

'cokey' is the unique unit identifier for the component layer.

```{r}


#Check if component layers are unique
layer_data$component %>% 
  nrow()

layer_data$component %>% 
  dplyr::select(cokey) %>% 
  distinct() %>% 
  nrow()

Component<-layer_data$component

```

#### Chorizon table (Soil horizon info)

Chorizon does not have unique cokey. The multiple subcomponents within each component for Chorizon are multiple soil horizons (at different depths).

You can select by hzname, and/or by horizon depth (hzdept_r - top depth and hzdepb-r - bottom depth) - from gSSURGO Table Column report: hzdepb_r - the distance from the top of the soil to the base of the soil horizon. Note: table report notes depth units as cm

You need one set of attributes per map unit (mukey), where the attributes are weighted averages of the components in each map unit. You can't have multiple attributes at different depths per cokey, especially since the depths sampled in different cokeys aren't uniform

**Want to revisit this for this analysis:** Strategy used here: pick the lowest soil horizon corresponding to hzdepthb_r \< 31 (ie, in plowable layer) corresponding to \<12 inches or \< 30.48 cm. **That might be too deep of a layer?** **Or may want to ask the research team.**

**Original approach:**

-eliminate samples from the very shallow surface that are likely to be plant material? nrow(Chorizon.surface) Chorizon.soil\<-Chorizon.surface %\>% filter(hzdepb_r\>3

-choose uppermost surface layer, to avoid component attributes that include plant material at the surface rather than soil

**New approach:**

-taking uppermost layer in top 5 inches

-Question: does this leave enough data to work with??

The assumption here is that gSSURGO attributes for P samples that are collected below the surface layer will closely mirror those at the top layer. This is borne out when you examine the soil texture for multiple depths within a component (a cokey); Texture attributes are typically similar across multiple depths.

```{r}


#check if chorizon layers map 1:1 with mukey 
layer_data$chorizon %>% 
  nrow()

layer_data$chorizon %>% 
  dplyr::select(cokey) %>% 
  distinct() %>% 
  nrow()

#-->does not map 1:1

names(layer_data$chorizon)


Chorizon.surface<-layer_data$chorizon %>% 
  filter(hzdepb_r<12.7) #restrict to top 5 inches?


#does this still leave you with duplicates per site?
#(ie, multiple depths above 30cm)
length(Chorizon.surface$cokey)
length(unique(Chorizon.surface$cokey))

#are still multiple measures per site; multiple depths

#choose uppermost surface layer, 
# convert to data.table with key=cokey
Chorizon.table <- data.table(Chorizon.surface, key="cokey")

#get the subset of data that matches this criterion
Chorizon.upper<-unique(Chorizon.table[, .SD[hzdepb_r %in% min(hzdepb_r)], by=cokey])
head(Chorizon.upper)

#check out depths at which samples are collected
summary(Chorizon.upper$hzdepb_r)

#check for duplicates
length(Chorizon.upper$cokey)
length(unique(Chorizon.upper$cokey))


#Now only one record per cokey (the top most surface measurement in the top 12 cm (ie 5 inches))
###
```

#### Chtexture & chtexturegrp table (Soil texture)

Notes:

-   chkey - primary key in horizon table - unique identifier for each soil horizon

-   chtexture - contains info on soil texture (silt, clay, sand etc) for a given soil horizon

    -   there are different textures per horizon; for the different subcomponents - see <https://sdmdataaccess.nrcs.usda.gov/documents/ReturningSoilTextureRelatedAttributes.pdf>

-   chtkey - links soil texture (in chtexture) to soil horizon

-   chtexturegrp - contains complex texture group names

-   'rvindicator' column in chtexturegrp - from Table Column Descriptions: 'A yes/no field that indicates if a value is representative for the component'

-   Select rvindicator = Yes - leaves you with unique value, this is the 'representative' component

```{r}
#'Chtexture' Table (texture/grain size)
#Can be merged to Chorizon, & with Chtexturegrp


Texture<-merge(layer_data$chtexture, layer_data$chtexturegrp %>% dplyr::select(!State), by=c('chtgkey'))
nrow(Texture)


#Join to Texture to Chorizon, uisng chkey
names(Chorizon.upper)
Horizon<-merge(Chorizon.upper, Texture[Texture$rvindicator=="Yes",], by=c('chkey'))
head(Horizon)
nrow(Horizon)
names(Horizon)

#do I have duplicate measurements in Chtexture or Chtexturegrp?? (ie, subcomponets?)
length(Horizon$cokey)
length(unique(Horizon$cokey))
#still have duplicates


#find duplicate rows
names(Horizon)

#Duplicates have same texdesc
#duplicates have same cokey and chkey and chtgkey, different chtkey
#slight differences in the different texcl per chtkey, e.g., Loam vs Silt loam

#Reduce to one record per depth component - 
#Pick first record for each component
#since all are listed as 'representative'; have to pick one 
Horizon2 <- Horizon[!duplicated(Horizon$chkey),]
nrow(Horizon)
nrow(Horizon2)
length(Horizon2$cokey)
length(unique(Horizon2$cokey))
#Ok now are unique

```

#### Copm and copmgrp (Soil Parent Origin)

```{r}

#turn warnings back on, now that all attribute tables are loaded
options(warn=0)

#Aggregate parent material from the start 

Parent<-merge(layer_data$copm, layer_data$copmgrp %>% dplyr::select(!State), by=c('copmgrpkey'))
head(Parent)
nrow(Parent)
#same number of rows as Copm; so was a 1:1 join

#check for duplicate cokeys
length(Parent$cokey)
length(unique(Parent$cokey))
#yes are duplicates, not 1:1 with cokey

#As for Texture, select rvindicator=Yes to avoid duplicates
names(Parent)
Parent2<-Parent[Parent$rvindicator=="Yes",]

#now check for duplicate cokeys
length(Parent2$cokey)
length(unique(Parent2$cokey))

#still have duplicates -
#find duplicates:
head(Parent2)
nrow(Parent2[duplicated(Parent2$cokey)|duplicated(Parent2$cokey, fromLast=TRUE),])
duplicates.Parent<-Parent2[duplicated(Parent2$cokey)|duplicated(Parent2$cokey, fromLast=TRUE),]
duplicates.Parent[1:20,]

#pmorder - the sequence in which the parent material occurs, when more than one parent material exists for one soil profile
#Note: will select pmorder ==1; assume this is the material closes to the surface in the soil profile
#Note that when there is only one parent material, no entry is required for pmorder, so also allow NA

Parent3<-Parent2[Parent2$pmorder=="1"|is.na(Parent2$pmorder),]
nrow(Parent3)

#now check for duplicate cokeys
length(Parent3$cokey)
length(unique(Parent3$cokey))
#still a smaller number of duplicates
duplicates.Parent3<-Parent3[duplicated(Parent3$cokey)|duplicated(Parent3$cokey, fromLast=TRUE),]

#View(duplicates.Parent3[1:20,])
#are multiple pmkind in the same order/pmgroupname
#pmgroupname seems to specify all possible combinations for each component
levels(factor(Parent3$pmgroupname)) #over 5000 levels

#Pick one pmkind for each horizon - 
#since they are all 'representative'
#this isn't perfect, but subsets 1 type to each horizon
Parent4 <- Parent3[!duplicated(Parent3$cokey),]
nrow(Parent3)
nrow(Parent4)
length(Parent4$cokey)
length(unique(Parent4$cokey))
#Now there are no duplicates for cokey


```

#### Valu1 Table

These are commonly used attributes already aggregated in SSURGO at the map unit scale

```{r}
valu1<-layer_data$Valu1
```

### Subset to most useful soil attributes

Subset to desired ssurgo attributes - ie, those likely to be related to P content - as per Records et al 2016 <https://doi.org/10.1016/j.earscirev.2016.04.010>

Select attributes ending in "\_r" as these are considered "representative" (according to SSURGO manual)

Note that this list does include total P values

```{r}
#COMPLETE LIST OF SSURGO TABLES NEEDED:
names(Component) #component unit (cokey)
names(Horizon2) #component unit (cokey)
names(Parent4) #component unit (cokey)
names(valu1) #map unit (mukey)

##
##Subset to desired ssurgo attributes - ie, those likely to be related to P content (as per Records et al 2016)

#Subset Component
names(Component)
Component.list<-c('comppct_r', 'slope_r', 'slopelenusle_r','runoff',
                  'erocl', 'hydricon', 'hydricrating', 'drainagecl', 'elev_r', 'geomdesc',
                  'airtempa_r', 'reannualprecip_r', 'ffd_r', 'taxceactcl', 'taxreaction', 'taxpartsize',
                  'mukey', 'cokey')
Component.sub<-Component[,names(Component) %in% Component.list]
head(Component.sub)
names(Component.sub)

#Subset Horizon2
names(Horizon2)
Horizon.list<-c('cokey', 'hzdept_r', 'hzdepb_r','sieveno4_r','sieveno10_r','sieveno40_r','sieveno200_r',
                'sandtotal_r','silttotal_r','claytotal_r','claysizedcarb_r','om_r','ksat_r',
                'kffact','caco3_r','gypsum_r', 'freeiron_r','feoxalate_r','extral_r','aloxalate_r','pbray1_r', 'poxalate_r', 
                'partdensity', 'ph2osoluble_r', 'ptotal_r', 'texcl','texture', 'texdesc')

Horizon2<-as.data.frame(Horizon2)
Horizon.sub<-Horizon2[ , names(Horizon2) %in% Horizon.list] 
names(Horizon.sub)
#View(Horizon.sub[,c(1,2,3,26,27,28)])
nrow(Horizon.sub)

length(Horizon.sub$cokey)
length(unique(Horizon.sub$cokey))

#Subset Parent
names(Parent4)
head(Parent4)
nrow(Parent4)
Parent4.list<-c('cokey', 'pmkind', 'pmorigin')
Parent.sub<-Parent4[,names(Parent4) %in% Parent4.list]
head(Parent.sub)
nrow(Parent.sub)

#Merge everything to mukey; using cokey
#need comppct from component table for % contribution of each component to each mukey (see below)

#Merge Component and Horizon
Comp1<-merge(Component.sub, Horizon.sub, by=c('cokey'))
head(Comp1)
nrow(Comp1)

#Some mukeys are lost because of restriction to surface samples with depth <12.7 cm

#so some mukeys only have attributes for bottom soil depths below 31 cm


#check for duplicates
length(Comp1$cokey)
length(unique(Comp1$cokey))
#No duplicates


#Merge in Parent table
Comp2<-merge(Comp1, Parent.sub, by=c('cokey'))
head(Comp2)
nrow(Comp2)
#check for duplicates
length(Comp2$cokey)
length(unique(Comp2$cokey))
#No duplicates


```

### Summarize/combine gSSURGO attributes at the map unit scale

Summarize component scale attributes at the map unit scale.

For numeric variables, take weighted average for each map unit.

For categorical variable, take component with highest % of map unit.

Weights are 'comppct_r' #i.e., the percent of each component in the map unit (the mukey)

For each numerical attribute at component scale, need to calculate weighted average #for each mukey (each group)

Check column types to make sure each is designated as correct type \#(either numeric or factor)

```{r}

sapply(Comp2, class)
head(Comp2)
names(Comp2)

#use the transform method to change the in built type of each feature.
#transform numeric variables
Comp2 <- transform(
  Comp2,
  slope_r=as.numeric(slope_r),
  slopelenusle_r=as.numeric(slopelenusle_r),  
  elev_r=as.numeric(elev_r),
  airtempa_r=as.numeric(airtempa_r),
  reannualprecip_r=as.numeric(reannualprecip_r),
  ffd_r=as.numeric(ffd_r),
  sieveno4_r=as.numeric(sieveno4_r),
  sieveno10_r=as.numeric(sieveno10_r),
  sieveno40_r=as.numeric(sieveno40_r),
  sieveno200_r=as.numeric(sieveno200_r),
  sandtotal_r=as.numeric(sandtotal_r),
  silttotal_r=as.numeric(silttotal_r),
  claytotal_r=as.numeric(claytotal_r),
  claysizedcarb_r=as.numeric(claysizedcarb_r),
  om_r=as.numeric(om_r),
  partdensity=as.numeric(partdensity),
  ksat_r=as.numeric(ksat_r),
  kffact=as.numeric(kffact),
  caco3_r=as.numeric(caco3_r),
  gypsum_r=as.numeric(gypsum_r),
  freeiron_r=as.numeric(freeiron_r),
  feoxalate_r=as.numeric(feoxalate_r),
  extral_r=as.numeric(extral_r),
  aloxalate_r=as.numeric(aloxalate_r),
  pbray1_r=as.numeric(pbray1_r),
  poxalate_r=as.numeric(poxalate_r),
  ph2osoluble_r=as.numeric(ph2osoluble_r),
  ptotal_r=as.numeric(ptotal_r),
  comppct_r=as.numeric(comppct_r))

#check levels for each categorical variables
#note - random forest models can only handle categorical variables with a certain number of levels (53)

levels(factor(Comp2$runoff)) # 6 levels
levels(factor(Comp2$erocl)) # 5 levels
levels(factor(Comp2$hydricon)) # 3 levels
levels(factor(Comp2$hydricrating)) # 3 levels
levels(factor(Comp2$drainagecl)) # 8 levels
levels(factor(Comp2$geomdesc)) #4061 levels!! -->Exclude
levels(factor(Comp2$taxpartsize)) #38 levels --> Combine? Or Exclude;
#^^might be represented by other variables, e.g., % clay
levels(factor(Comp2$taxceactcl)) #5 levels
levels(factor(Comp2$taxreaction)) #7 levels
levels(factor(Comp2$texcl)) #21 levels, could substitute for taxpartsize maybe
levels(factor(Comp2$pmkind)) #65 levels -->could aggregate some levels
levels(factor(Comp2$pmorigin)) #54 levels

#Variables to exclude or modify (eg, aggregate or code dummy variables):
#geomdesc, taxpartsize

#for pmorigin, combine similar categories:
#combine all organic materials
Comp2$pmkind<-ifelse(Comp2$pmkind=="Grassy organic material", "Organic material", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Herbaceous organic material", "Organic material", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Mossy organic material", "Organic material", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Woody organic material", "Organic material", Comp2$pmkind)
#combine all types of till
Comp2$pmkind<-ifelse(Comp2$pmkind=="Basal till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Flow till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Lodgement till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Melt-out till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Subglacial till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Supraglacial till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Ablation till", "Till", Comp2$pmkind)
#combine mine spoil
Comp2$pmkind<-ifelse(Comp2$pmkind=="Coal extraction mine spoil", "Mine spoil-coal or metal", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Metal ore extraction mine spoil", "Mine spoil-coal or metal", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Mine spoil or earthy fill", "Mine spoil-coal or metal", Comp2$pmkind)

levels(factor(Comp2$pmkind))

#transform categorical variables
Comp2 <- transform(
  Comp2,
  runoff=as.factor(runoff),
  erocl=as.factor(erocl),
  hydricon=as.factor(hydricon),
  hydricrating=as.factor(hydricrating),
  drainagecl=as.factor(drainagecl),
  taxpartsize=as.factor(taxpartsize),
  taxceactcl=as.factor(taxceactcl),
  taxreaction=as.factor(taxreaction),
  texcl=as.factor(texcl),
  pmkind=as.factor(pmkind),  
  mukey=as.factor(mukey)
)

sapply(Comp2, class)

#Check data:
#check that numeric variables are numeric summaries
#check that categorical variables are counts
#check for missing data ('?')
summary(Comp2)


###
#LIST OF NUMERIC & CATEGORICAL ATTRIBUTES

Numeric.list<-c('slope_r', 'slopelenusle_r', 'elev_r', 
                'airtempa_r', 'reannualprecip_r', 'ffd_r','sieveno4_r', 'sieveno10_r', 'sieveno40_r', 'sieveno200_r', 
                'sandtotal_r', 'silttotal_r', 'claytotal_r', 'claysizedcarb_r', 'om_r', 'partdensity',
                'ksat_r', 'kffact', 'caco3_r', 'gypsum_r', 'freeiron_r', 'feoxalate_r', 'extral_r', 
'aloxalate_r', 'pbray1_r', 'poxalate_r', 'ph2osoluble_r', 'ptotal_r')

Categorical.list<-c('mukey', 'comppct_r', 'runoff', 'erocl', 'hydricon', 'hydricrating', 'drainagecl', 'taxpartsize',
                    'taxceactcl', 'taxreaction', 'texcl', 'pmkind')


#Variables to exclude: 
#geomdesc - too many levels



##SUMMARIZE ATTRIBUTES:

#NUMERIC
#calculate weighted mean for each mukey (weights are % contribution to mukey area) 
#calculate for multiple attributes at once
#note: for weighted mean approach attributes must be numeric

names(Comp2)

Mukey.mean<-Comp2 %>% group_by(mukey) %>% 
  summarise_at(Numeric.list, 
               list(~weighted.mean(., comppct_r)))

head(Mukey.mean)

#make into a regular table
Mukey.mean<-as.data.frame(Mukey.mean)
head(Mukey.mean)

#check for duplicates
length(Mukey.mean$mukey)
length(unique(Mukey.mean$mukey))
#no duplicates

#CATEGORICAL
#select component level with the highest percentage area in the map unit
#if two component levels are tied for area, use tie breaker (see below)
#highest compcct_r for each attribute in each mutkey

#group by mukey
#apply just across columns in Categorical.list
#select attribute levels in each group based on max compcct_r

# convert to data.table with key=mukey
Comp.table <- data.table(Comp2[,Categorical.list], key="mukey")
# get the subset of data that matches this criterion
Mukey.top2<-unique(Comp.table[, .SD[comppct_r %in% max(comppct_r)], by=mukey])
head(Mukey.top2)

#check for duplicates
length(Mukey.top2$mukey)
length(unique(Mukey.top2$mukey))

#Are some duplicates here - for some map units, multiple components have same area
#Need to further winnow down - see below


###
##So now you have:
#weighted average for numeric variables for each map unit
#component with highest percentage area in map unit for categorical variables

#Merge the two;
SSURGO.att<-merge(Mukey.mean, Mukey.top2, by=c('mukey'))
head(SSURGO.att)
#check for duplicates
length(SSURGO.att$mukey)
length(unique(SSURGO.att$mukey))

#Add in attributes from Valu1 table
SSURGO.all<-merge(SSURGO.att, valu1, by=c('mukey'))
head(SSURGO.all)

#check for duplicates
length(SSURGO.all$mukey)
length(unique(SSURGO.all$mukey))

#are some duplicates - due to multiple components of same area in some map units
#Choose higher slope as the tie breaker, because surface P runoff increases for soils with greater slope
#select rows with highest slope in each mukey
names(SSURGO.all)

SSURGO.top<-as.data.frame(
  SSURGO.all %>% group_by(mukey) %>% top_n(1, slope_r))
head(SSURGO.top)

#check for duplicates
length(SSURGO.top$mukey)
length(unique(SSURGO.top$mukey))

#Still some duplicates above -- Choose a row at random, as a second tie breaker
SSURGO.unique <- SSURGO.top[!duplicated(SSURGO.top$mukey),]
head(SSURGO.unique)

#check for duplicates
length(SSURGO.unique$mukey)
length(unique(SSURGO.unique$mukey))
#Now have unique records

#write to table;
#these are unique attributes for each map unit, from multiple SSURGO tables

write.table(SSURGO.unique, paste0(dir.output, '/SSURGO_attributes_mapunit.csv'), sep=",", row.names=FALSE)
```

### Merge ssurgo attributes to sample locations

```{r}

#Load in Soil P locations if not already loaded:
soilP<-read.csv(paste0(dir.output, '/sample_attributes_streamcat.csv'),header=TRUE)
names(soilP[1:25])

#Load in SSURGO attributes if needed
SSURGO.unique<-read.csv(paste0(dir.output, '/SSURGO_attributes_mapunit.csv'))

#make matching MUKEY column
SSURGO.unique<-SSURGO.unique %>% 
  dplyr::rename(MUKEY=mukey)
names(SSURGO.unique)


soilP.att<-merge(soilP, SSURGO.unique, by=c('MUKEY'))
head(soilP.att)
names(soilP.att)
nrow(soilP.att)

#View(soilP.att)

#Write to file
write.table(soilP.att, paste0(dir.output, '/soil_sample_attributes_data_for_modelling.csv'), sep=",", row.names=FALSE)

```

# Clean up all attributes - Might not need this for now? LEFT OFF 8.7.25

Load sampling locations w/spatial attributes and merge with SSURGO & StreamCat attributes

Sample datasets have had the following pre-processing:

-   NHD stream type from NHDv2Plus

-   NWI wetland type, if within wetland polygon

-   NLCD 2019 Land Use category in 30m cell where sample occurs

-   EPA StreamCat attributes

-   gSSURGO attributes
