---
title: "Assigning spatial attributes to sample locations"
author: "Christy Dolph"
date: "2024-07-17"
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Overview: Generating spatial attributes for sites of interest 

This script generates spatial attributes associated with specific point locations.

Attributes are from the following sources:

-    the NHDPlusv2 stream network (NHD)

-   the National Wetlands Inventory (NWI),

-   National Land Cover Dataset (NLCD)

-   SSURGO soils data

-   EPA StreamCat

This approach is updated and adapted from the scripts available at: <https://github.com/cldolph/SoilP>

## Steps:

1.  Merge locations to NHD, NLCD, and NWI attributes + SSURGO map unit key (mukey)
2.  Compile and process SSURGO attributes and merge
3.  Compile and process StreamCat attributes and merge

# National Hydrography Dataset (NHD)

Use nhdplusTools package to link sample locations to NHD COMID (unique identifier for each stream reach of the NHD)

The COMID can then be used to link the location to NHD attributes (stream type, stream order, etc) as well as EPA StreamCat attributes.

Note that 'featureid' of an NHD Catchment is equal to COMID of an NHD flowline; see p. 47 of the NHDPlusV2 User Guide:

[https://www.epa.gov/system/files/documents/2023-04/NHDPlusV2_User_Guide.pdf](https://www.epa.gov/system/files/documents/2023-04/NHDPlusV2_User_Guide.pdf,)

### Download flowlines for NHD stream network

Note: it might be faster to save this to a temp file - need to figure out how to do that

```{r}

#Get NHD data for HUC4s of interest via web service, using nhdplusTools:

NHD_HUC4<-get_nhdplus(
  AOI = HUC4_AOI)

names(NHD_HUC4)

st_crs(NHD_HUC4) #check reference system

#plot NHD flowlines (note this can take several minutes)
#tm_shape(NHD_HUC4)+
 # tm_lines()

View(head(NHD_HUC4))

```

### Identify NHD COMIDs associated with sample locations

```{r}

#first transform pour point and flowlines to same projection:
#continental US (Conus Albers NAD83)

loc_trans <- st_transform(loc_sf, 5070)
NHD_trans<-st_transform(NHD_HUC4, 5070)

Sys.time()
loc_snap <- st_snap(loc_trans, NHD_trans, tolerance = 100)
Sys.time() #4 min for IA sites

#Quick plot check
plot(loc_snap$geometry)


#The function discover_nhdplus_id() can pull comids for single point locations; to get comids for all points you must loop this function over every point

# Initialize a vector to store COMIDs
comids <- numeric(nrow(loc_snap))

# Loop over each row (point)
for (i in seq_len(nrow(loc_snap))) {
  # Extract a single-point sf object
  single_point <- loc_snap[i, ]
  
  # Try running discover_nhdplus_id()
  # Use tryCatch to prevent errors from stopping the loop
  comids[i] <- tryCatch({
    discover_nhdplus_id(single_point)
  }, error = function(e) {
    NA  # return NA if there's an error
  })
}

# Add the COMIDs back to the sf object
loc_snap$comid <- comids

#View(loc_snap)

```

### Join NHD attributes to sampled locations, based on COMID

see NHDv2Plus User Guide: <https://nhdplus.com/NHDPlus/NHDPlusV2_documentation.php>

'totdasqkm' = Total Upstream Cumulative Drainage Area, in square kilometers, at the downstream end of the NHDFlowline feature

```{r}
nrow(loc_snap)

loc_NHD<-merge(loc_snap, NHD_HUC4 %>% dplyr::select(comid, gnis_name, lengthkm, ftype, fcode, streamleve, streamorde, totdasqkm) %>% st_drop_geometry(), by=c("comid"))

nrow(loc_NHD)
View(loc_NHD)
```

# National Wetlands Inventory (NWI)

See access to NWI data at: <https://www.fws.gov/program/national-wetlands-inventory/data-download>

Download data by state: <https://www.fws.gov/program/national-wetlands-inventory/download-state-wetlands-data>

### Load in wetlands data:

```{r}

#US states where samples occur (corresponding to study area):
states<-list(States_sub$STUSPS)
states


##To download NWI files for every state in the your study area of interest, loop GET url for multiple states:

# Base URL where NWI data is available
base_url <- "https://documentst.ecosphere.fws.gov/wetlands/data/State-Downloads"

# Loop through state abbreviations and download corresponding zip files

for (i in states) {
  zip_name <- paste0(i, "_geodatabase_wetlands.zip")
  url <- paste0(base_url, "/", zip_name)
  destfile <- file.path("./spatial_data", zip_name)
  
  if (!file.exists(destfile)) {
    message("Downloading: ", url)
    tryCatch({
      response <- GET(url, write_disk(destfile, overwrite = TRUE),
                      user_agent("Mozilla/5.0 (Windows NT 10.0; Win64; x64)"))
      if (response$status_code != 200) {
        warning("Download failed for ", states, " with status ", response$status_code)
      }
    }, error = function(e) {
      warning("Failed to download ", states, ": ", e$message)
    })
  } else {
    message("Already downloaded: ", states)
  }
}
--------------------------------------------------------

# Unzip geodatabase(s)

#List of zipped files:
zip_wetlands <- list.files("./spatial_data", pattern = "wetlands.*\\.zip$", full.names = TRUE)
zip_wetlands

#Unzip files to spatial data directory:
ldply(.data = zip_wetlands, .fun = unzip, exdir = "./spatial_data")

--------------------------------------------------------
##Read in wetlands layer from every geodatabase (from each different state) in study region
  

# Initialize a list to hold each state's wetlands layer
wetlands_list <- list()

# Loop over state abbreviations and read each layer into the list

for (i in states) {
  gdb_path <- file.path("spatial_data", paste0(i, "_geodatabase_wetlands.gdb"))
  layer_name <- paste0(i, "_Wetlands")

  if (dir.exists(gdb_path)) {
    message("Reading layer for state: ", i)
    tryCatch({
      wetlands_list[[i]] <- st_read(dsn = gdb_path, layer = layer_name)
    }, error = function(e) {
      warning("Failed to read ", i, ": ", e$message)
    })
  } else {
    warning("GDB file does not exist for ", i, " at path: ", gdb_path)
  }
}


# Combine all shapefiles into one
combined_wetlands <- do.call(rbind, wetlands_list)


# Check spatial extent of wetlands data loaded in (the bounding box)
tmap_mode("view")
tm_shape(st_as_sfc(st_bbox(combined_wetlands)))+
  tm_borders()+
tm_shape(States_AOI)+
  tm_borders()

#Remove zipped geodatabase files

# Get list of all zip files in the directory
zip_files <- list.files(gdb_dir, pattern = "\\.zip$", full.names = TRUE)

# Remove each zip file
for (zip_file in zip_files) {
  file.remove(zip_file)
  cat("Removed:", zip_file, "\n")
}

```

### Join NWI attributes to sample locations:

```{r}

#Project locations into NAD83 Albers to match wetlands layer:
st_crs(combined_wetlands)
loc_NHDproj<-st_transform(loc_NHD, crs=st_crs(combined_wetlands))

Sys.time()
#Join selected wetland attributes to sample locations:
loc_NWI <- st_join(loc_NHDproj, combined_wetlands %>% dplyr::select(ATTRIBUTE, WETLAND_TYPE, ACRES), join = st_intersects)
Sys.time()

names(loc_NWI)

#View(loc_NWI)

#write intermediate sf object to file, to save all NWI info:
st_write(loc_NWI, paste0(dir.spatial, "/loc_NHD_NWI.shp"), append=TRUE)


```

# National Land Cover Database (NLCD)

Extract the NLCD value in the 30m cell where the sample was collected.

One option is to use get_nlcd() from the FedData package. The url this draws data from, www.mrlc.gov, can sometimes go down and be non functional.

Another option is to download NLCD data directly to the 'Spatial_data' folder. You can access the NLCD by state here: <https://datagateway.nrcs.usda.gov/GDGHome_DirectDownLoad.aspx>

Accessed 7.24.25

Load in NLCD data:

LEFT OFF 7.24.25- need to rework this section, may need to download NLCD data directly (get_nlcd not working)

--hmm, might be that mlrc site is down?? (can't access it on internet either)

### Load NLCD data

**Option 1: using get_nlcd()**

```{r}
# Download the NLCD data for the specified AOI
#Note you are using 2019 NLCD data here (may want to adjust based on when data was collected):
Sys.time()
nlcd_data <- get_nlcd(template = States_AOI, label = "NLCD_AOI", year = 2019, dataset = "landcover")
head(nlcd_data)
Sys.time()
```

**Option 2: load after manually downloading nlcd rasters for each state to your project folder in a directory called 'nlcd'**

```{r}

# Load all rasters for each state in study region


# Full path to the nlcd folder
nlcd_dir <- file.path(dir.spatial, "nlcd")

# Find all .tif files in subfolders that start with "nlcd"
tif_files <- list.files(nlcd_dir,
                        pattern = "^nlcd.*\\.tif$",
                        recursive = TRUE,
                        full.names = TRUE)
tif_files

# Read each TIFF file into a SpatRaster object
nlcd_rasters <- lapply(tif_files, rast)


# Strip names from rasters
names(nlcd_rasters) <- NULL

# Combine using terra's merge (for adjacent, non-overlapping rasters)
#Note - this is taking a long time!
Sys.time()
nlcd_combined <- do.call(terra::merge, nlcd_rasters)
Sys.time()


#check reference system
st_crs(nlcd_combined)
       
#Check out on a map
names(nlcd_combined)
#tmap_mode("view")

tm_shape(nlcd_combined)+
  tm_raster(col="Class")+
  tm_shape(loc_NWI)+
  tm_dots(col="black")

```

### Join NLCD attributes to sample locations

```{r}
#Project land cover raster to NAD83 Albers:
Sys.time()
target_crs<-crs(combined_wetlands)
nlcd_projAlbers <- project(nlcd_combined, target_crs)
Sys.time()

# Extract the raster value at the point location
loc_NLCD <- extract(nlcd_projAlbers, loc_NWI)


#Bind NLCD attributes to other attributes:
loc_NLCD<-cbind(loc_NWI, loc_NLCD)

#Rename NLCD attribute
loc_NLCD<-loc_NLCD %>% 
  rename(NLCDCat=Class)

#Save intermediate attribute file: 
st_write(loc_NLCD, "./spatial_data/loc_NHD_NWI_NLCD.shp", append=FALSE)


#Look at attribute table containing NHD, NWI and NLCD attributes:
View(loc_NLCD)
```

# SSURGO dataset

Need to identify map unit key ('mukey') for each soil sample location. Will need ssurgo raster data for all states in study area.

Download and unzip state gSSURGO files from <https://www.nrcs.usda.gov/resources/data-and-reports/gridded-soil-survey-geographic-gssurgo-database>

see: "State Databases - soils - powered by Box"

Put unzipped geodatabase folders in "spatial_data/ssurgo" folder

If you want to look in more detail at soil series, NRCS has a query table by state where you can look at map unit soil series definitions:

<https://www.nrcs.usda.gov/publications/query-by-state.html>

### Load SSURGO data

**Load** **after manually downloading nlcd rasters for each state to your project folder in a directory called 'nlcd'**

```{r}

#test for multiple states
states<-c('IA','IL')
#otherwise assign states to study region
#states<-list(States_sub$STUSPS)

# Read in ssurgo layer from every geodatabase (from each different state) in study region
  

# Full path to the ssurgo folder
ssurgo_dir <- file.path(dir.spatial, "ssurgo")


# Initialize a list to hold each state's ssurgo map unit layer
ssurgo_list <- list()

# Loop over state abbreviations and read each layer into the list

for (i in states) {
  gdb_path <- file.path(ssurgo_dir, paste0("gSSURGO_",i, ".gdb"))
  layer_name <- "MUPOLYGON"

  if (dir.exists(gdb_path)) {
    message("Reading layer for state: ", i)
    tryCatch({
      ssurgo_list[[i]] <- st_read(dsn = gdb_path, layer = layer_name)
    }, error = function(e) {
      warning("Failed to read ", i, ": ", e$message)
    })
  } else {
    warning("GDB file does not exist for ", i, " at path: ", gdb_path)
  }
}

head(ssurgo_list)

# Combine all shapefiles into one
combined_ssurgo <- do.call(rbind, ssurgo_list)

names(combined_ssurgo)

# Check spatial extent of ssurgo data loaded in (the bounding box)
tmap_mode("view")
tm_shape(st_as_sfc(st_bbox(combined_ssurgo)))+
  tm_borders()+
tm_shape(States_AOI)+
  tm_borders()

```

### Join SSURGO attributes to sample locations

```{r}

#Load in sample locations (if not already loaded)
#Read in table as needed: 
loc_NLCD<-st_read("./spatial_data/loc_NHD_NWI_NLCD.shp")

#Note slightly different CRS
#Get CRS to match exactly

loc_NLCD_NAD83ConusAlbers<-st_transform(loc_NLCD, crs=st_crs(combined_ssurgo))
crs(loc_NLCD_NAD83ConusAlbers)

#Join gSSURO mapunit attributes to soil sampl locations:
loc_ssurgo <- st_join(loc_NLCD_NAD83ConusAlbers, combined_ssurgo %>% dplyr::select(MUKEY, MUSYM), join = st_intersects)

nrow(loc_ssurgo)

View(loc_ssurgo)

#Write data with all attribute keys to file:
st_write(loc_ssurgo, "./spatial_data/loc4_NHD_NWI_NLCD_gSSURGO.shp", append=FALSE)

#Just pull out attributes:
loc_ssurgo_data<-st_drop_geometry(loc_ssurgo)
head(loc_ssurgo_data)
write.csv(loc_ssurgo_data, "sample_attributes.csv", sep=",", row.names=FALSE)

```

### 

# Load Soil Samples with all attribute keys attached:

```{r}

soil_loc<-st_read("./spatial_data/loc4_NHD_NWI_NLCD_gSSURGO.shp")
class(soil_loc)

plot(soil_loc)


```

Extract just the data frame of attributes for further manipulation:

```{r}

P.att <- soil_loc %>%
  mutate(x = st_coordinates(.)[,1], y = st_coordinates(.)[,2]) %>%
  st_drop_geometry() %>% 
  data.frame() 

#Drop unneeded columns
drops <- c("MUKEY","ACRES")
P.att2<-P.att[ , !(names(P.att) %in% drops)]

View(P.att2)

#write to file:
write.csv(P.att, "./Attributes_for_soilP_model.csv", row.names=FALSE)
```

# Get StreamCat Attributes

NOTE, 2.5.25: this is now defunct, as FTP servers for StreamCat have been retired

Download StreamCat attributes for states of interest (MS and AR):

```{r}
#Download StreamCat tables for states of interest (MS and AR) from the web
#Note that this will download more attributes than you need for soil P model, but can subset them later using soil P model attribute list


# URL of the webpage
url <- "https://gaftp.epa.gov/epadatacommons/ORD/NHDPlusLandscapeAttributes/StreamCat/States/"

# Read the webpage content
webpage <- read_html(url)

# Extract all links from the webpage
links <- webpage %>% html_nodes("a") %>% html_attr("href")

# Filter links that contain "MS"
ms_links <- links[grep("MS.zip|AR.zip", links)]

# Base URL for downloading files
base_url <- "https://gaftp.epa.gov/epadatacommons/ORD/NHDPlusLandscapeAttributes/StreamCat/States/"

# Directory to save the downloaded files
download_dir <- "StreamCat"
if (!dir.exists(download_dir)) {
  dir.create(download_dir)
}

# Download each file
for (link in ms_links) {
  file_url <- paste0(base_url, link)
  file_name <- paste0(download_dir, "/", basename(link))
  GET(file_url, write_disk(file_name, overwrite = TRUE))
  cat("Downloaded:", file_name, "\n")
}

cat("All files downloaded.\n")


```

Unzip StreamCat files:

```{r}


# Directory where the zip files are located
zip_dir <- "./StreamCat"

# Get list of all zip files in the directory
zip_files <- list.files(zip_dir, pattern = "\\.zip$", full.names = TRUE)


# Unzip files
for (zip_file in zip_files) {
    # Define the directory where the unzipped files will be saved
    unzip_dir <- file_path_sans_ext(zip_file)
    if (!dir.exists(unzip_dir)) {
      dir.create(unzip_dir)
    }
    
    # Unzip the file
    unzip(zip_file, exdir = unzip_dir)
    cat("Unzipped:", zip_file, "\n")
}

cat("All matching files have been unzipped.\n")
```

Remove zipped StreamCat files to save space:

```{r}
#Remove zipped files to save space


# Directory where the zip files are located
zip_dir <- "StreamCat"

# Get list of all zip files in the directory
zip_files <- list.files(zip_dir, pattern = "\\.zip$", full.names = TRUE)

# Remove each zip file
for (zip_file in zip_files) {
  file.remove(zip_file)
  cat("Removed:", zip_file, "\n")
}

cat("All zip files have been removed.\n")
```

Read in StreamCat csv files and merge into one file:

```{r}
#Read in StreamCat files
#Have to read in separately by each state at first 



# Directory where the CSV files are located
base_dir <- "./StreamCat"

# Partial strings to match in filenames
partial_string_1 <- "MS.csv"
partial_string_2 <- "AR.csv"

# Function to list all CSV files in subdirectories
list_csv_files <- function(base_dir) {
  list.files(path = base_dir, pattern = "\\.csv$", recursive = TRUE, full.names = TRUE)
}

# Function to read and process each CSV file
read_and_process_csv <- function(file) {
  read_csv(file, show_col_types = FALSE) %>%
    mutate(source = basename(file))
}

# Get list of all CSV files in the base directory and its subdirectories
csv_files <- list_csv_files(base_dir)

# Separate CSV files into two lists based on partial string matches
csv_files_1 <- csv_files[grepl(partial_string_1, csv_files)]
csv_files_2 <- csv_files[grepl(partial_string_2, csv_files)]

# Read and process each list of CSV files
list_of_dataframes_1 <- map(csv_files_1, read_and_process_csv)
list_of_dataframes_2 <- map(csv_files_2, read_and_process_csv)

(list_of_dataframes_1)

#Exclude redundant/unneeded columns (will add WsArea and CatArea back in later)
area.att<-c('CatAreaSqKm', 'WsAreaSqKm', 'CatPctFull', 'WsPctFull', 'CatAreaSqKmRp100', 'WsAreaSqKmRp100','CatPctFullRp100','WsPctFullRp100','source', 'CatAreaSqKmSlp20', 'WsAreaSqKmSlp20',"CatPctFullSlp20","WsPctFullSlp20","CatAreaSqKmSlp10","WsAreaSqKmSlp10","CatPctFullSlp10","WsPctFullSlp10")

list_of_data_frames_1b<-lapply(list_of_dataframes_1, function(x) x[!(names(x) %in% area.att)])

list_of_data_frames_2b<-lapply(list_of_dataframes_2, function(x) x[!(names(x) %in% area.att)])

#First merge all dataframes for each state, by COMID
merged_data1 <- reduce(list_of_data_frames_1b, full_join, by =('COMID'))

merged_data2 <- reduce(list_of_data_frames_2b, full_join, by =('COMID'))

#Pull out area attributes for each state:
Area1<-as.data.frame(list_of_dataframes_1[1]) %>% dplyr::select(c('COMID', 'CatAreaSqKm', 'WsAreaSqKm'))
Area2<-as.data.frame(list_of_dataframes_2[1]) %>% dplyr::select(c('COMID', 'CatAreaSqKm', 'WsAreaSqKm'))

#Add area attributes back in:
merged_data1.area<-merge(merged_data1, Area1, by=c("COMID"))
merged_data2.area<-merge(merged_data_2, Area2, by=c("COMID"))


#Remove redundant columns (occuring in more than one StreamCat table)
#remove attributes that contain ".y" or multiple ".x"
#remove ".x" from any column names

merged_data_1simp<-merged_data1.area %>% 
  dplyr::select(!contains(c(".y",".x.x")))
colnames(merged_data_1simp) <- gsub("\\.x$", "", colnames(merged_data_1simp))

merged_data_2simp<-merged_data2.area %>% 
 dplyr::select(!contains(c(".y",".x.x")))
colnames(merged_data_2simp) <- gsub("\\.x$", "", colnames(merged_data_2simp))


#Merge all StreamCat attributes from all states:
streamcat.all<-rbind(merged_data_1simp, merged_data_2simp)
nrow(streamcat.all)

# Optional: Save the merged streamcat data to a new CSV file
write_csv(streamcat.all, "./StreamCat/StreamCat_merged_data.csv")

#Option to remove all unzipped StreamCat files to save space:
# Directory where the subdirectories are located
base_dir <- "./StreamCat"

# Define the partial strings to match
patterns_to_remove <- c("MS", "AR")

# Function to list all subdirectories
list_subdirs <- function(base_dir) {
  dir_ls(base_dir, type = "directory", recurse = TRUE)
}

# Get list of all subdirectories in the base directory
subdirs <- list_subdirs(base_dir)

# Identify subdirectories to remove based on partial string matches
subdirs_to_remove <- subdirs[grepl(paste(patterns_to_remove, collapse = "|"), subdirs)]

# Remove the identified subdirectories
for (subdir in subdirs_to_remove) {
  dir_delete(subdir)
  cat("Removed:", subdir, "\n")
}

```

# Use Attribute Keys to Add and Process Geospatial Attributes

Now that attribute keys for NHD, NWI, NLCD, gSSURGO and StreamCat have been added, you can follow the modules from Dolph et al 2023 to do additional processing of attributes necessary for running the soil P model. This entails following modules 2 and 4 from the github repository: <https://github.com/cldolph/SoilP/>

### Module 2: Preprocess SSURGO attributes

```{r}
#Module for pre-processing SSURGO attributes for use as covariates in the random forest model

#Updated 7/23/2024

#Read in ArcGIS dbf attribute tables 

#Download and unzip state gSSURGO files from https://nrcs.app.box.com/v/soils/folder/148414960239
#to the project directory 'gSSURGO'
#states needed: AR, IA, IL, IN, KS, MN, MO, ND, NE, OH, SD, WI

#load relevant tables from state geodatabases:


#can ignore warnings about simple feature geometries not being present
options(warn=-1)

#Set directory for ssurgo data
ssurgo_dir<-"./spatial_data"

#'Component' Table

AR <- sf::st_read(dsn = "./spatial_data/gSSURGO_AR.gdb", layer = "component")
MS<-sf::st_read(dsn = "./spatial_data/gSSURGO_MS.gdb", layer = "component")

#Append all tables using rbind
Component<-rbind(AR, MS)
nrow(Component)
head(Component)

#look for example mukey
Component %>% filter(mukey=="MS083")


#check if cokeys are unique
length(Component$cokey)
length(unique(Component$cokey))
#Ok same length - so are unique

#'Chorizon' Table
AR2 <- sf::st_read(dsn = "./spatial_data/gSSURGO_AR.gdb", layer = "chorizon")
MS2 <- sf::st_read(dsn = "./spatial_data/gSSURGO_MS.gdb", layer = "chorizon")

#Append all tables using rbind
Chorizon<-rbind(AR2, MS2)
nrow(Chorizon)
head(Chorizon)
names(Chorizon)

#check for unique cokey
length(Chorizon$cokey)
length(unique(Chorizon$cokey))

#Chorizon does not have unique cokey; 
#Note multiple subcomponents within component layer for Chorizon are multiple soil horizons (at different depths)
#you can select by hzname, and/or by horizon depth (hzdept_r - top depth and hzdepb-r - bottom depth)
#from gSSURGO Table Column report: hzdepb_r - the distance from the top of the soil to the base of the soil horizon
#note: table report notes depth units as cm

#you need one set of attributes per map unit (mukey), where the attributes are weighted averages
#of the components; so can't have multiple attributes at different depths per cokey
#especially since the depths sampled in different cokeys aren't uniform 

#pick all hzdepthb_r < 31 (ie, in plowable layer) 
#corresponding to <12 inches or < 30.48 cm
#assumption here is that gSSURGO attributes for P samples that are collected below the surface layer
#will closely mirror those at the top layer
#this is borne out when you examine the soil texture for multiple depths within a component (a cokey);
#texture attributes are typically similar across multiple depths

Chorizon[1:5, c(1,7, 10, 170, 171)]

Chorizon.surface<-Chorizon %>% 
  filter(hzdepb_r<31)#restrict to plowable layer

#does this still leave you with duplicates per site?
#(ie, multiple depths above 30cm)
length(Chorizon.surface$cokey)
length(unique(Chorizon.surface$cokey))

#are still multiple measures per site; multiple depths

#eliminate samples from the very shallow surface that are likely to be plant material
nrow(Chorizon.surface)
Chorizon.soil<-Chorizon.surface %>% 
  filter(hzdepb_r>3)

#choose lowermost surface layer, to avoid component attributes that include plant material at the surface rather than soil
# convert to data.table with key=cokey
Chorizon.table <- data.table(Chorizon.soil, key="cokey")

#get the subset of data that matches this criterion
Chorizon.bottom<-unique(Chorizon.table[, .SD[hzdepb_r %in% max(hzdepb_r)], by=cokey])
head(Chorizon.bottom)

#check out depths at which samples are collected
summary(Chorizon.bottom$hzdepb_r)
names(Chorizon.bottom)
#View(Chorizon.bottom[, c(1,8, 11, 171)])


#check for duplicates
length(Chorizon.bottom$cokey)
length(unique(Chorizon.bottom$cokey))
###
#OK! Now only one record per cokey (the lowermost surface measurement in the top 30cm)
###


#'Chtexture' Table (grain size)
#Can be merged to Chorizon, with Chtexturegrp
AR5 <- sf::st_read(dsn = "./spatial_data/gSSURGO_AR.gdb", layer = "chtexture")
MS5 <- sf::st_read(dsn = "./spatial_data/gSSURGO_MS.gdb", layer = "chtexture")

#Append all tables using rbind
Chtexture<-rbind(AR5, MS5)
nrow(Chtexture)
head(Chtexture)
names(Chtexture)

#check for unique subunits
length(Chtexture$chtkey)
length(unique(Chtexture$chtkey))
#Chtkey is unique

#'Chtexturegrp' Table (links chtexture to chorizon table)
#Can be merged to Chorizon, with Chtexturegrp
AR6 <- sf::st_read(dsn = "./spatial_data/gSSURGO_AR.gdb", layer = "chtexturegrp")
MS6 <- sf::st_read(dsn = "./spatial_data/gSSURGO_MS.gdb", layer = "chtexturegrp")

#Append all tables using rbind
Chtexturegrp<-rbind(AR6, MS6)
nrow(Chtexturegrp)
head(Chtexturegrp)
names(Chtexturegrp)

Texture<-merge(Chtexture, Chtexturegrp, by=c('chtgkey'))
nrow(Texture)

#check for duplicates at highest level of organization (chkey)
#-->yes, are multiple chkeys; chtkeys appear unique
#can be different textures per horizon
#what are the different subcomponents - see https://sdmdataaccess.nrcs.usda.gov/documents/ReturningSoilTextureRelatedAttributes.pdf

#there is a column 'rvindicator' in chtexturegrp - from Table Column Descriptions:
#'A yes/no field that indicates if a value is representative for the component'
#so select rvindicator = Yes - leaves you with unique value
#this is the 'representative' component

#Join to Texture to Chorizon, uisng chkey
names(Chorizon.bottom)
Horizon<-merge(Chorizon.bottom, Texture[Texture$rvindicator=="Yes",], by=c('chkey'))
head(Horizon)
nrow(Horizon)

#do I have duplicate measurements in Chtexture or Chtexturegrp?? (ie, subcomponets?)
length(Horizon$cokey)
length(unique(Horizon$cokey))
#still have duplicates
#with rvindicator=Yes for Texture, leaves 113 duplicate cokeys for merged Horizon table

#find duplicate rows
names(Horizon)

#View(Horizon[duplicated(Horizon$chkey)|duplicated(Horizon$chkey, fromLast=TRUE),c(1,2,3,9,12,172,173,175,176,179)])

#Duplicates have same texdesc
#duplicates have same cokey and chkey and chtgkey, different chtkey
#slight differences in the different texcl per chtkey, e.g., Loam vs Silt loam

#Reduce to one record per depth component - 
#Pick first record for each component
#since all are listed as 'representative'; have to pick one 
Horizon2 <- Horizon[!duplicated(Horizon$chkey),]
nrow(Horizon)
nrow(Horizon2)
length(Horizon2$cokey)
length(unique(Horizon2$cokey))
#Ok now are unique

#'Copm' Table (Soil Parent Origin)
AR3 <- sf::st_read(dsn = "./spatial_data/gSSURGO_AR.gdb", layer = "copm")
MS3 <- sf::st_read(dsn = "./spatial_data/gSSURGO_MS.gdb", layer = "copm")

#Append all tables using rbind
Copm<-rbind(AR3, MS3)
nrow(Copm)
head(Copm)
names(Copm)



#'Copmgrp Table' (Needed to link Copm - parent material - table to component)
AR4 <- sf::st_read(dsn = "./spatial_data/gSSURGO_AR.gdb", layer = "copmgrp")
MS4 <- sf::st_read(dsn = "./spatial_data/gSSURGO_MS.gdb", layer = "copmgrp")

#turn warnings back on, now that all attribute tables are loaded
options(warn=0)

#Append all tables using rbind
Copmgrp<-rbind(AR4, MS4)
nrow(Copmgrp)
head(Copmgrp)
names(Copmgrp)

#Aggregate parent material from the start 
#(unless there are additional subcomponents within components??)

Parent<-merge(Copm, Copmgrp, by=c('copmgrpkey'))
head(Parent)
nrow(Parent)
#same number of rows as Copm; so was a 1:1 join
#check for duplicate cokeys
length(Parent$cokey)
length(unique(Parent$cokey))
#yes are duplicates, not 1:1 with cokey
#As for Texture, select rvindicator=Yes to avoid duplicates
names(Parent)
Parent2<-Parent[Parent$rvindicator=="Yes",]
#now check for duplicate cokeys
length(Parent2$cokey)
length(unique(Parent2$cokey))
#still have duplicates -
#find duplicates:
head(Parent2)
nrow(Parent2[duplicated(Parent2$cokey)|duplicated(Parent2$cokey, fromLast=TRUE),])
duplicates.Parent<-Parent2[duplicated(Parent2$cokey)|duplicated(Parent2$cokey, fromLast=TRUE),]
duplicates.Parent[1:20,]
#many duplicates: 200,000+
#out of ~430,000 rows
#pmorder - the sequence in which the parent material occurs, when more than one parent material exists for one soil profile
#Note: will select pmorder ==1; assume this is the material closes to the surface in the soil profile
#Note that when there is only one parent material, no entry is required for pmorder, so also allow NA

Parent3<-Parent2[Parent2$pmorder=="1"|is.na(Parent2$pmorder),]
nrow(Parent3)

#now check for duplicate cokeys
length(Parent3$cokey)
length(unique(Parent3$cokey))
#still a smaller number of duplicates - ~2000
duplicates.Parent3<-Parent3[duplicated(Parent3$cokey)|duplicated(Parent3$cokey, fromLast=TRUE),]
#View(duplicates.Parent3[1:20,])
#are multiple pmkind in the same order
#pmgroupname seems to specify all possible combinations for each component
levels(factor(Parent3$pmgroupname)) #over 5000 levels
#Pick one pmkind for each horizon - 
#since they are all 'representative'
#this isn't perfect, but subsets 1 type to each horizon
Parent4 <- Parent3[!duplicated(Parent3$cokey),]
nrow(Parent3)
nrow(Parent4)
length(Parent4$cokey)
length(unique(Parent4$cokey))
#Now there are no duplicates for cokey

#'Valu1' Table:

AR_valu <- sf::st_read(dsn = "./spatial_data/gSSURGO_AR.gdb", layer = "Valu1")
MS_valu <- sf::st_read(dsn = "./spatial_data/gSSURGO_MS.gdb", layer = "Valu1")

head(AR_valu)

valu1<-rbind(MS_valu, AR_valu)
head(valu1)
valu1$MUTKEY<-valu1$mukey

#COMPLETE LIST OF SSURGO TABLES NEEDED:
names(Component) #component unit (cokey)
names(Horizon2) #component unit (cokey)
names(Parent4) #component unit (cokey)
names(valu1) #map unit (mukey)

##
##Subset to desired ssurgo attributes - ie, those likely to be related to P content (as per Records et al 2016)

#Subset Component
names(Component)
Component.list<-c('comppct_r', 'slope_r', 'slopelenusle_r','runoff',
                  'erocl', 'hydricon', 'hydricrating', 'drainagecl', 'elev_r', 'geomdesc',
                  'airtempa_r', 'reannualprecip_r', 'ffd_r', 'taxceactcl', 'taxreaction', 'taxpartsize',
                  'mukey', 'cokey')
Component.sub<-Component[,names(Component) %in% Component.list]
head(Component.sub)
names(Component.sub)

#Subset Horizon2
names(Horizon2)
Horizon.list<-c('cokey', 'hzdept_r', 'hzdepb_r','sieveno4_r','sieveno10_r','sieveno40_r','sieveno200_r',
                'sandtotal_r','silttotal_r','claytotal_r','claysizedcarb_r','om_r','ksat_r',
                'kffact','caco3_r','gypsum_r', 'freeiron_r','feoxalate_r','extral_r','aloxalate_r','pbray1_r', 'poxalate_r', 
                'partdensity', 'ph2osoluble_r', 'ptotal_r', 'texcl','texture', 'texdesc')

Horizon2<-as.data.frame(Horizon2)
Horizon.sub<-Horizon2[ , names(Horizon2) %in% Horizon.list] 
names(Horizon.sub)
#View(Horizon.sub[,c(1,2,3,26,27,28)])
nrow(Horizon.sub)

length(Horizon.sub$cokey)
length(unique(Horizon.sub$cokey))

#Subset Parent
names(Parent4)
head(Parent4)
nrow(Parent4)
Parent4.list<-c('cokey', 'pmkind', 'pmorigin')
Parent.sub<-Parent4[,names(Parent4) %in% Parent4.list]
head(Parent.sub)
nrow(Parent.sub)

#Merge everything to mukey; using cokey
#need comppct from component table for % contribution of each component to each mukey (see below)

#Merge Component and Horizon
Comp1<-merge(Component.sub, Horizon.sub, by=c('cokey'))
head(Comp1)
nrow(Comp1)

#Some mukeys are lost because of restriction to surface samples with depth <31 cm
#e.g., 
filter(Comp1, mukey==424124) #value is not present
filter(Component.sub, mukey==424124) #it is present in the original Component table
#so some mukeys only have attributes for bottom soil depths below 31 cm

#how many mukeys are lost?

#~10,000 mukeys lost, region wide

#check for duplicates
length(Comp1$cokey)
length(unique(Comp1$cokey))
#No duplicates


#Merge in Parent table
Comp2<-merge(Comp1, Parent.sub, by=c('cokey'))
head(Comp2)
nrow(Comp2)
#check for duplicates
length(Comp2$cokey)
length(unique(Comp2$cokey))
#No duplicates


##################################################################
#Summarize component scale attributes at the map unit scale
#For numeric variables, take weighted average for each map unit; 
#For categorical variable, take component with highest % of map unit;

#weights are 'comppct_r'
#i.e., the percent of each component in the map unit (the mukey)

#For each numerical attribute at component scale, need to calculate weighted average
#for each mukey (each group)

#check column types to make sure each is designated as correct type
#(either numeric or factor)

sapply(Comp2, class)
head(Comp2)
names(Comp2)

#use the transform method to change the in built type of each feature.
#transform numeric variables
Comp2 <- transform(
  Comp2,
  slope_r=as.numeric(slope_r),
  slopelenusle_r=as.numeric(slopelenusle_r),  
  elev_r=as.numeric(elev_r),
  airtempa_r=as.numeric(airtempa_r),
  reannualprecip_r=as.numeric(reannualprecip_r),
  ffd_r=as.numeric(ffd_r),
  sieveno4_r=as.numeric(sieveno4_r),
  sieveno10_r=as.numeric(sieveno10_r),
  sieveno40_r=as.numeric(sieveno40_r),
  sieveno200_r=as.numeric(sieveno200_r),
  sandtotal_r=as.numeric(sandtotal_r),
  silttotal_r=as.numeric(silttotal_r),
  claytotal_r=as.numeric(claytotal_r),
  claysizedcarb_r=as.numeric(claysizedcarb_r),
  om_r=as.numeric(om_r),
  partdensity=as.numeric(partdensity),
  ksat_r=as.numeric(ksat_r),
  kffact=as.numeric(kffact),
  caco3_r=as.numeric(caco3_r),
  gypsum_r=as.numeric(gypsum_r),
  freeiron_r=as.numeric(freeiron_r),
  feoxalate_r=as.numeric(feoxalate_r),
  extral_r=as.numeric(extral_r),
  aloxalate_r=as.numeric(aloxalate_r),
  pbray1_r=as.numeric(pbray1_r),
  poxalate_r=as.numeric(poxalate_r),
  ph2osoluble_r=as.numeric(ph2osoluble_r),
  ptotal_r=as.numeric(ptotal_r),
  comppct_r=as.numeric(comppct_r))

#check levels for each categorical variables
#note - random forest models can only handle categorical variables with a certain number of levels (53)

levels(factor(Comp2$runoff)) # 6 levels
levels(factor(Comp2$erocl)) # 5 levels
levels(factor(Comp2$hydricon)) # 3 levels
levels(factor(Comp2$hydricrating)) # 3 levels
levels(factor(Comp2$drainagecl)) # 8 levels
levels(factor(Comp2$geomdesc)) #4061 levels!! -->Exclude
levels(factor(Comp2$taxpartsize)) #38 levels --> Combine? Or Exclude;
#^^might be represented by other variables, e.g., % clay
levels(factor(Comp2$taxceactcl)) #5 levels
levels(factor(Comp2$taxreaction)) #7 levels
levels(factor(Comp2$texcl)) #21 levels, could substitute for taxpartsize maybe
levels(factor(Comp2$pmkind)) #65 levels -->could aggregate some levels
levels(factor(Comp2$pmorigin)) #54 levels

#Variables to exclude or modify (eg, aggregate or code dummy variables):
#geomdesc, taxpartsize

#for pmorigin, combine similar categories:
#combine all organic materials
Comp2$pmkind<-ifelse(Comp2$pmkind=="Grassy organic material", "Organic material", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Herbaceous organic material", "Organic material", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Mossy organic material", "Organic material", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Woody organic material", "Organic material", Comp2$pmkind)
#combine all types of till
Comp2$pmkind<-ifelse(Comp2$pmkind=="Basal till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Flow till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Lodgement till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Melt-out till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Subglacial till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Supraglacial till", "Till", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Ablation till", "Till", Comp2$pmkind)
#combine mine spoil
Comp2$pmkind<-ifelse(Comp2$pmkind=="Coal extraction mine spoil", "Mine spoil-coal or metal", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Metal ore extraction mine spoil", "Mine spoil-coal or metal", Comp2$pmkind)
Comp2$pmkind<-ifelse(Comp2$pmkind=="Mine spoil or earthy fill", "Mine spoil-coal or metal", Comp2$pmkind)

levels(factor(Comp2$pmkind))

#transform categorical variables
Comp2 <- transform(
  Comp2,
  runoff=as.factor(runoff),
  erocl=as.factor(erocl),
  hydricon=as.factor(hydricon),
  hydricrating=as.factor(hydricrating),
  drainagecl=as.factor(drainagecl),
  taxpartsize=as.factor(taxpartsize),
  taxceactcl=as.factor(taxceactcl),
  taxreaction=as.factor(taxreaction),
  texcl=as.factor(texcl),
  pmkind=as.factor(pmkind),  
  mukey=as.factor(mukey)
)

sapply(Comp2, class)


#Note: i'm guessing you will also lose some levels when you merge to soil P data (subset of the country)

#Check data:
#check that numeric variables are numeric summaries
#check that categorical variables are counts
#check for missing data ('?')
summary(Comp2)


###
#LIST OF NUMERIC & CATEGORICAL ATTRIBUTES

Numeric.list<-c('slope_r', 'slopelenusle_r', 'elev_r', 
                'airtempa_r', 'reannualprecip_r', 'ffd_r','sieveno4_r', 'sieveno10_r', 'sieveno40_r', 'sieveno200_r', 
                'sandtotal_r', 'silttotal_r', 'claytotal_r', 'claysizedcarb_r', 'om_r', 'partdensity',
                'ksat_r', 'kffact', 'caco3_r', 'gypsum_r')

Categorical.list<-c('mukey', 'comppct_r', 'runoff', 'erocl', 'hydricon', 'hydricrating', 'drainagecl', 'taxpartsize',
                    'taxceactcl', 'taxreaction', 'texcl', 'pmkind')


#Variables to exclude: 
#geomdesc - too many levels
#almost no or no data; 'freeiron_r', 'feoxalate_r', 'extral_r', 
#'aloxalate_r', 'pbray1_r', 'poxalate_r', 'ph2osoluble_r', 'ptotal_r'

##SUMMARIZE ATTRIBUTES:

#NUMERIC
#calculate weighted mean for each mukey (weights are % contribution to mukey area) 
#calculate for multiple attributes at once
#note: for weighted mean approach attributes must be numeric

#library(dplyr)
names(Comp2)

Mukey.mean<-Comp2 %>% group_by(mukey) %>% 
  summarise_at(Numeric.list, 
               list(~weighted.mean(., comppct_r)))

head(Mukey.mean)

#make into a regular table
Mukey.mean<-as.data.frame(Mukey.mean)
head(Mukey.mean)

#check for duplicates
length(Mukey.mean$mukey)
length(unique(Mukey.mean$mukey))
#no duplicates

#CATEGORICAL
#select component level with the highest percentage area in the map unit
#if two component levels are tied for area, use tie breaker (see below)
#highest compcct_r for each attribute in each mutkey

#group by mukey
#apply just across columns in Categorical.list
#select attribute levels in each group based on max compcct_r

# convert to data.table with key=mukey
Comp.table <- data.table(Comp2[,Categorical.list], key="mukey")
# get the subset of data that matches this criterion
Mukey.top2<-unique(Comp.table[, .SD[comppct_r %in% max(comppct_r)], by=mukey])
head(Mukey.top2)

#check for duplicates
length(Mukey.top2$mukey)
length(unique(Mukey.top2$mukey))

#Are some duplicates here - for some map units, multiple components have same area
#Need to further winnow down - see below


###
##So now you have:
#weighted average for numeric variables for each map unit
#component with highest percentage area in map unit for categorical variables

#Merge the two;
SSURGO.att<-merge(Mukey.mean, Mukey.top2, by=c('mukey'))
head(SSURGO.att)
#check for duplicates
length(SSURGO.att$mukey)
length(unique(SSURGO.att$mukey))

#Add in attributes from Valu1 table
SSURGO.all<-merge(SSURGO.att, valu1, by=c('mukey'))
head(SSURGO.all)

#check for duplicates
length(SSURGO.all$mukey)
length(unique(SSURGO.all$mukey))

#are some duplicates - due to multiple components of same area in some map units
#Choose higher slope as the tie breaker, because surface P runoff increases for soils with greater slope
#select rows with highest slope in each mukey
names(SSURGO.all)

SSURGO.top<-as.data.frame(
  SSURGO.all %>% group_by(mukey) %>% top_n(1, slope_r))
head(SSURGO.top)

#check for duplicates
length(SSURGO.top$mukey)
length(unique(SSURGO.top$mukey))

#Still some duplicates above -- Choose a row at random, as a second tie breaker
SSURGO.unique <- SSURGO.top[!duplicated(SSURGO.top$mukey),]
head(SSURGO.unique)

#check for duplicates
length(SSURGO.unique$mukey)
length(unique(SSURGO.unique$mukey))
#Now have unique records

#write to table;
#these are unique attributes for each map unit, from multiple SSURGO tables

write.table(SSURGO.unique, './SSURGO_attributes_unique_for_mapunit_Delta.csv', sep=",", row.names=FALSE)

```

### Module 4: Merge and clean up all attributes

```{r}
# STEP 4: LOAD IN SOIL P Locations w/SPATIAL ATTRIBUTES #
#& MERGE WITH SSURGO & STREAMCAT ATTRIBUTES #

#soil P datasets have had the following pre-processing:

#NHD stream type from NHDv2Plus, if within 100 m buffer of stream network
#NWI wetland type, if within wetland polygon
#NLCD 2019 Land Use category in 30m cell where sample occurs
#SSURGO mutkey ID (for merging with ssurgo attributes - see below)
#NHDv2Plus Catchment COMID (for merging with StreamCat attributes - see below)

#Load in Soil P locations
soilP<-read.delim('./Attributes_for_soilP_model.csv', sep=',',header=TRUE)
head(soilP)
names(soilP)

#Format column headers to match existing soil P script:
soilP<-soilP %>% 
  rename(FTYPE=ftype) %>% 
  rename(FCODE=fcode) %>% 
  rename(NLCD19=NLCDCat) %>% 
  rename(LONGITUDE=x) %>% 
  rename(LATITUDE=y) %>% 
  rename(WETLAND_TYPE=WETLAND_TY) %>% 
  rename(MUTKEY=MUKEY)

#Check # of unique sample locations
nrow(unique(soilP[c("LATITUDE","LONGITUDE")]))


#Attributes to keep:
Attributes.Keep<-c('LATITUDE', 'LONGITUDE', 'COMID', 'MUTKEY', 'FTYPE', 'FCODE', 'WETLAND_TYPE', 
                      'NLCD19')

soilP2<-soilP[,names(soilP) %in% Attributes.Keep]
names(soilP2)

###########################################################
##ADDITIONAL PREDICTOR VARIABLES

#load in additional SSURGO (soil) attributes
ssurgo<-read.delim('./SSURGO_attributes_unique_for_mapunit_Delta.csv', sep=",", header=TRUE)
head(ssurgo)
names(ssurgo)


###############################################################

##MERGE SOIL P DATA TO PREDICTOR VARIABLES


#For USGS GEOCHEM DATA + NCSS TP DATA:
soilP.att<-merge(soilP2, ssurgo, by=c('MUTKEY'))
head(soilP.att)
names(soilP.att)
nrow(soilP.att)


####
#Create Levels for Categorical Attribute data

#set NA to 0 for FCODE
soilP.att$FCODE[is.na(soilP.att$FCODE)] <- 0
head(soilP.att)

#Check levels for streams layer
levels(factor(soilP.att$FCODE))

#Create levels for streams layer;
#levels are: 
soilP.att$StreamType<-factor(soilP.att$FCODE, levels=c('0', '33600', '46003', '46006', '55800'),
                             labels=c('NonStream', 'Canal/Ditch', 'Intermittent_Stream', 'Perennial_Stream', 'Artificial_Path'))


#Create categorical variable for NWI categories
names(soilP.att)
levels(factor(soilP.att$WETLAND_TYPE))
soilP.att$WetlandType<-factor(soilP.att$WETLAND_TYPE, levels=c( 
                                                               'Freshwater Forested/Shrub Wetland'),
                              labels=c('Freshwater Forested/Shrub Wetland'))
head(soilP.att)
levels(soilP.att$WetlandType)


#NLCD 2019 Land Use category - note this category has some overlap with NWI
#check number of levels
levels(factor(soilP.att$NLCD19))
#15 levels should be OK
#create categorical variable
#1 - 21 - Developed, Open Space
#2 - 22 - Developed, Low Intensity
#3 - 23 - Developed, Medium Intensity
#12 - 82 - Cultivated Crops
#13 - 90 - Woody Wetlands
#14 - 95 - Emergent Herbaceous Wetlands


######################################################
#SUBSET COLUMNS to bind to StreamCat dataset:
names(soilP.att)
head(soilP.att)

Numeric.list<-c('P_mgkg',
                'p_ox', 'p_bray1', 'pbray2', 'p_olsn', 'p_nz', 'p_h2o', 'p_citx', 'p_mehlich3', 'n_no3',
                'slope_r', 'slopelenusle_r', 'elev_r', 
                'airtempa_r', 'reannualprecip_r', 'ffd_r','sieveno4_r', 'sieveno10_r', 'sieveno40_r', 'sieveno200_r', 
                'sandtotal_r', 'silttotal_r', 'claytotal_r', 'claysizedcarb_r', 'om_r', 'partdensity',
                'ksat_r', 'kffact', 'caco3_r', 'gypsum_r',
                'aws0_5', 'aws5_20', 'aws20_50', 'aws50_100', 'aws100_150', 'aws150_999', 'aws0_20', 'aws0_30',
                'aws0_100', 'aws0_150', 'aws0_999','soc0_5' , 'soc5_20' ,'soc20_50'                   
                , 'soc50_100' ,'soc100_150', 'soc150_999' ,'soc0_20'                    
                , 'soc0_30' ,'soc0_100' , 'soc0_150','soc0_999'                                                 
                , 'nccpi3corn' ,'nccpi3soy' , 'nccpi3cot','nccpi3sg'                   
                , 'nccpi3all','pctearthmc' , 'rootznemc','rootznaws'                  
                , 'droughty','pwsl1pomu', 'Depth_cm')

Categorical.list<-c('mukey', 'comppct_r', 'runoff', 'erocl', 'hydricon', 'hydricrating', 'drainagecl', 'taxpartsize',
                    'taxceactcl', 'taxreaction', 'texcl', 'pmkind', 'Wetland_Attribute', 'StreamType',
                    'WetlandType', 'NLCD19')

Admin.list<-c('SiteID', 'mukey', 'site_key', 'latitude_decimal_degrees', 'longitude_decimal_degrees', 'state_name', 
              'county_code', 'county_name', 'labsampnum', 
              'hzn_top', 'hzn_bot',
              'hzn_desgn', 'hzn_prime', 'COMID','FEATUREID', 'musumcpct','COMID.x', 'COMID.y', 'COMID')

#musumcpct - the sum of comppct_r values for all listed components in the map unit

Location.list<-c('LATITUDE', 'LONGITUDE')

Big.List<-c(Numeric.list, Categorical.list, Admin.list, Location.list)


#Subset data to the attributes specified above:
data<-soilP.att[,names(soilP.att) %in% Big.List]
head(data)
names(data)
nrow(data)


################################################
##ADD IN STREAM CAT VARIABLES 
streamcat<-fread('./StreamCat/StreamCat_merged_data.csv', sep=",", header=TRUE)
head(streamcat)
names(streamcat)

#MERGE STREAMCAT TO ALL SOIL P DATA
All.Att<-merge(data, streamcat, by=c('COMID'))
head(All.Att)
nrow(All.Att)

```

#### Matching NLCD land cover years

-   The original soil P model for the UMRB was built with 2006 NLCD data.

-   It probably makes more sense to use the closest matching NLCD to when soil samples were collected (assuming 2019 for Mississippi Delta soil samples)

    -   Drop 2006 attributes from attribute list

    -   Label 2019 attributes as "2006" even though they are actually 2019!

-   See file "streamcat_variable_info_downloaded_072424.csv" for more info on StreamCat attributes at the time this script was written

-   Alternatively, you could use 2006 NLCD variables for Mississippi, because they match (not taking this approach for now)

```{r}

#Drop 2006 attributes from the attribute list, EXCEPT "PctAg2006Slp10Ws, as 2006 appears to be the only year for which this attribute exists (even tho streamcat variable list says it should exist for multiple NLCD series)

ncol(All.Att)

All.Att.cor <- All.Att[, !(grepl("2006", names(All.Att)) & names(All.Att) != "PctAg2006Slp10Ws")]

ncol(All.Att.cor) #check number of remaining columns


#Replace 2019 attributes with "2006"
colnames(All.Att.cor) <- gsub("2019", "2006", colnames(All.Att.cor))

#Similarly, rename the NLCD19 column to match dataset used to build model:
All.Att.cor<-All.Att.cor %>% 
  rename(NLCD06Cat=NLCD19) %>% 
  mutate(CatPctFull=100) %>% 
  mutate(WsPctFull=100)

#Create CatPctFull and WsPctFull columns (these typically don't contain useful information anyway)


#################################################################
###EXCLUDE VARIABLES:

#For this dataset, we need the same attributes that were used to build the original soil P model for the UMRB. 

#Load in attribute data for the original soil P model in the UMRB:
UMRB<-read.delim("./_UMRB_Soil_P_data_for_modelling.txt", sep=",", header=TRUE)
head(UMRB)
names(UMRB)

Orig.List<-colnames(UMRB)
Orig.List

#Find columns missing from MS Delta dataset that are in original UMRB dataset:

Orig.List[!Orig.List %in% colnames(All.Att.cor)]

#Temperature and Precip columns appear to have been updated since UMRB model was built
#Based on the streamcat variable list, "Precip8110" are the 30 year average values 1981-2010, Precip9120 are 30 year average balues 1991-2020

#In previous model, used mean values from 2008
#used 30 year averages from 1981-2010

#For this model, use mean values from 2008
#Use 30 year averages from 1981-2010

#Looks like "PctAg2006Slp10Ws" was renamed to "PctAgSlpMid[Year][AOI]" in StreamCat dataset


All.Att.cor %>% 
  dplyr::select(contains(c("Precip","Tmean","Tmax","Tmin"))) %>% names()

All.Att.cor2<-All.Att.cor %>% 
  rename(PrecipCat=Precip08Cat) %>% 
  rename(PrecipWs=Precip08Ws) %>% 
  rename(TmaxCat=Tmax8110Cat) %>% 
  rename(TminCat=Tmin8110Cat) %>% 
  rename(TmeanCat=Tmean08Cat) %>% 
  rename(TmaxWs=Tmax8110Ws) %>% 
  rename(TminWs=Tmin8110Ws) %>% 
  rename(TmeanWs=Tmean08Ws) 

#Check column matching again

Orig.List[!Orig.List %in% colnames(All.Att.cor2)]
#P_mgkg and Depth_cm are missing and need to be added by Fanyu


#Select only columns that were in original Soil P UMRB dataset:
All.Att.select<-All.Att.cor2 %>% 
  dplyr::select(c(LATITUDE, LONGITUDE, any_of(Orig.List)))



names(All.Att.select)


```

```{r}
#Write the final attributes data to file 

write.table(All.Att.select, "./soil_sample_attributes_data_for_modelling_Delta.csv", sep=",", row.names=FALSE)

```
