---
title: "module_tune_models"
author: "Christy Dolph"
date: "2025-08-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

# 1 Install and load the required libraries

```{r}

knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace 
rm(list=ls())


#list of necessary packages 
want <- c('tidyverse', 'sf', 'tmap', 'nhdplusTools', 'data.table', 'xgboost', 'caret', 'dplyr', 'ggplot2', 'scales', 'tidymodels', 'missRanger', 'doParallel', 'DALEXtra', 'vip', 'grid', 'stringr', 'readxl', 'writexl', 'shapviz', 'readr', 'yardstick', 'glmnet', 'janitor')

need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)

#Main directory
dir.root<-getwd()

#Input data
dir.data<-paste(dir.root, "/TP_data", sep="")

#Spatial data
dir.spatial<-paste(dir.root, "/Spatial_data", sep="")

#Output data (produced by script)
dir.output<-paste0(dir.root, "/output_data")

dir.figures<-paste0(dir.root, '/Maps_figures')

#set directories for final saved models + data

dir.final.model.data<-paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets")

dir.final.model<-paste0(dir.output, "/Final_model_comparisons")

```

# 2 Load the dataset(s)

```{r}

#Select the input data you'd like to use: 

#Margenot et al IL streambank P data, with limited ssurgo attributes to retain all data, for independent streambank model or for running through existing soil TP model (predictors are matched below): 
model.input<-"/Margenot_sample_attributes_for_modelling_wSSURGO_LIMITED"


#Modern upland Soil P database:
#model.input<-'/USGS_NCSS_soils_database_attributes_alldepths_for_modelling'

#Modern upland soil P database but with limited ssurgo attributes that should encompass water areas?
#model.input<-'/USGS_NCSS_soils_database_attributes_alldepths_for_modelling_wSSURGO_LIMITED'

P_dataset <-read.csv(paste0(dir.output, model.input, '.csv'))
                    
names(P_dataset) 
nrow(P_dataset)
#View(P_dataset)

#Check class of all attributes
sapply(P_dataset, class)

#Remove any invalid P measurements
P_dataset<-P_dataset %>% filter(P_mgkg>0)

#Convert characters to factors:
P_dataset <- P_dataset %>% 
  type.convert(as.is=FALSE)

#Convert integer attributes to numeric:
integer_cols <- sapply(P_dataset, is.integer)
P_dataset[integer_cols] <- lapply(P_dataset[integer_cols], as.numeric)

# Check the class of columns
sapply(P_dataset, class)

#Make name of P_mgkg column consistent
#NOTE: need to fix this later! (not working)
P_dataset<-P_dataset %>%
  { if ("meanP_mgkg" %in% names(.)) rename(., P_mgkg = meanP_mgkg) else . }

#Create group for siteID (multiple depth samples collected at same coordinate location)
#(only need to do this once)

P_dataset<-P_dataset %>%  
  mutate(y_rd = round(y, 5), x_rd = round(x, 5)) %>% 
  dplyr::group_by(y_rd, x_rd) %>%
  dplyr::mutate(siteID = cur_group_id()) %>%
  ungroup() %>% 
  #remove unneeded columns
  dplyr::select(-c(y_rd, x_rd))

names(P_dataset)

levels(factor(P_dataset$siteID))
```

Note that samples were collected in clusters along select river reaches in IL. As a result, samples in the same NHD catchment, will have the same values for features from EPA Streamcat (that are associated with NHD). Add in COMID if needed (NHD catchment identifier) so you can group-split sites in same COMID (as well as siteID) for train/test.

Update: in the updated script, comid was retained in module 2, so does not need to be added here.

```{r}



#create shapefile of sample locations:

P_sf<-st_as_sf(P_dataset, coords=c("x", "y"), crs=4269, remove=FALSE)


#creating bounding box for sample locations

bb = st_as_sfc(st_bbox(P_sf))

#Get NHD COMID info for catchments in the bounding box

NHD_catchment<-get_nhdplus(
  AOI = bb,
  comid = NULL,
  nwis = NULL,
  realization = "catchment",
  streamorder = NULL,
  t_srs = NULL
)

names(NHD_catchment)

#View(NHD_catchment)

#Check map:
tmap_mode("view")

#tm_shape(NHD_catchment)+
#  tm_borders()+
#  tm_shape(P_sf)+
#  tm_dots()


#Create COMID column
NHD_catchment$comid<-NHD_catchment$featureid

#Add COMID to sample sites

P_dataset<-st_join(P_sf, NHD_catchment %>% dplyr::select(comid)) %>% sf::st_drop_geometry()

#Get additional NHD attributes by comid

unique_comids <- P_dataset %>%
  pull(comid) %>%
  unique() %>%
  na.omit()

comid_attributes <- get_nhdplus(
  comid = unique_comids, 
  realization = "flowline"
) %>% st_drop_geometry()

#Add selected NHD attributes:
nrow(P_dataset)

P_dataset<-merge(P_dataset, comid_attributes %>% dplyr::select(comid, streamorde, totdasqkm, areasqkm, slope), by=c("comid"))

nrow(P_dataset)


#Write to file to skip this step next time:
write_xlsx(P_dataset, paste0(dir.output, model.input, "_comid.xlsx"))

```

## 2.1 Option to add pH and other site-specific data for streambank samples

```{r}

ph<-readxl::read_excel(paste0(dir.data, "/Margenot streambank soil data to Christy Dolph_01202026_wpH.xlsx"))
nrow(ph)
#match pH to existing attributes based on lat long

names(P_dataset)
names(ph)

#Clean up columns
ph.clean<-ph %>% 
  separate_wider_delim(cols = 'Depth (cm)', delim = "-", names = c("Depth_min_cm", "Depth_max_cm"),cols_remove = FALSE, too_few="debug") %>%
  dplyr::rename(x=Long, y=Lat, pct.sand=`%Sand`, pct.silt=`%Silt`, pct.clay=`%Clay`, Depth_cm=Depth_max_cm) %>% 
   dplyr::select(x,y,pH,pct.silt, pct.sand, pct.clay, Depth_cm) %>% 
  na.omit() #drop samples with missing data
names(ph.clean)

#View(ph.clean)
nrow(ph.clean)

#View(P_dataset %>% dplyr::select(x,y,P_mgkg, Depth_cm))

class(ph$x)
class(P_dataset$x)

P_dataset.ph<-merge(P_dataset, ph.clean, by=c('x', 'y', 'Depth_cm'))

nrow(P_dataset.ph) #Note 34 sites were dropped after merge 

#To use pH data:
P_dataset<-P_dataset.ph
```

# 3 Explore the data

## 3.1 Distribution of streambank P values

```{r}

nrow(P_dataset)

#Histogram of P
P.hist<-
  ggplot(P_dataset, aes(P_mgkg))+
  #geom_histogram(binwidth = 1,colour = "lightgray", fill = "darkgrey") +
  geom_histogram(binwidth=10, colour = "black", fill ="black") +
  #geom_vline(aes(xintercept=mean(P_mgkg)), linetype="dashed", color="blue")+
  geom_vline(aes(xintercept=mean(P_mgkg)), linetype="dashed", color="purple", linewidth=1.5)+
  #geom_vline(aes(xintercept=1000), linetype="dotdash", color="black")+
  #geom_vline(aes(xintercept=1000), linetype="dotdash", #color="darkgray", linewidth=1.5)+
  xlab("P (mg/kg)") +
  ylab ("Frequency") +
  #ggtitle("Near Channel Soil P distribution")+
  #ggtitle("Soil P Samples distribution") +
  theme(axis.title = element_text(size=20))+
  theme(axis.text=element_text(size=20))+
  #theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  #theme(plot.title = element_text(size = 26, face = "bold", hjust=0.5))
P.hist

summary(P_dataset$P_mgkg)

#Percent of values over 1000 mg/kg
nrow(P_dataset)

P_dataset %>% filter(P_mgkg>1000) %>% nrow()

#Max and mean of samples >1000 mg/kg
P_dataset %>% filter(P_mgkg>1000) %>% summarise(maxP=max(P_mgkg), mean=mean(P_mgkg))

```

Mean P value

```{r}

#Including and excluding samples with P >1000 mg/kg

P_dataset %>% 
  filter(P_mgkg<1000) %>% 
  summarise(meanP=mean(P_mgkg))

```

P values by comid (NHD catchment)

```{r}

ggplot(P_dataset)+
  geom_boxplot(aes(factor(comid), P_mgkg))

```

Plot streambank P vs bulk density

```{r}

ggplot(P_dataset) +
  geom_point(aes(meanBD_gcm3, P_mgkg))

```

## 3.2 Check class of all variables

```{r}
summary(P_dataset)
class(P_dataset)

```

Create correct/matching levels for NLCD attribute

```{r}
#Set NLCD_Land to categorical

levels(factor(P_dataset$NLCD_Land))

#For Margenot Streambank samples, get levels to match soil TP:
P_dataset$NLCD_Land<-factor(P_dataset$NLCD_Land,
                       labels=c("Barren Land", "Cultivated Crops", "Deciduous Forest", "Developed, Low Intensity", "Developed, Open Space", "Emergent Herbaceous Wetlands", "Pasture/Hay", "Mixed Forest", "Open Water", "Woody Wetlands"))

#For big soil TP dataset, set factor levels:
P_dataset$NLCD_Land<-factor(P_dataset$NLCD_Land, 
                            levels=c('11','21','22','23','24','31','41','42','43','52', '71','81','82','90','95'),
                            labels=c('Open Water', 'Developed, Open Space', 'Developed, Low Intensity','Developed, Medium Intensity', 'Developed, High Intensity', 'Barren Land','Deciduous Forest','Evergreen Forest','Mixed Forest','Shrub/Scrub','Grassland/Herbaceous', 'Pasture/Hay','Cultivated Crops','Woody Wetlands','Emergent Herbaceous Wetlands'))

levels(P_dataset$NLCD_Land)
```

Assign correct variable type to remaining variables:

```{r}
#Set remaining categorical attributes to factor
cat_vars <- names(P_dataset)[sapply(P_dataset, function(x) is.factor(x) | is.character(x))]
cat_vars

P_dataset<-P_dataset %>% 
  mutate(across(cat_vars, as.factor))


#Set remaining numeric attributes to numeric
dep_var="P_mgkg"
num_vars <- names(P_dataset)[sapply(P_dataset, is.numeric) & names(P_dataset) != dep_var]
num_vars

P_dataset <- P_dataset %>%
  mutate(across(num_vars, as.numeric))

#Check all columns
sapply(P_dataset, class)
```

## 3.3 Relationship to other variables

### a. Categorical attributes

```{r}

#For categorical attributes, plot boxplots:  

#full dataset:
df<-P_dataset

dep_var="P_mgkg"
cat_vars <- names(df)[sapply(df, function(x) is.factor(x) | is.character(x))]


# Open a PDF device
pdf(paste0(dir.output, model.input, "_boxplots_by_category.pdf"))

# Loop through categorical variables and plot to PDF (one per page)
walk(cat_vars, function(cat_var) {
  p <- ggplot(df, aes_string(x = cat_var, y = dep_var)) +
    geom_boxplot(fill = "skyblue", color = "black") +
    labs(title = paste("Boxplot of", dep_var, "by", cat_var)) +
    theme_minimal()+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 
  
  print(p)  # important to print inside loop when saving
})

# Close PDF device
dev.off()

```

### b. Numeric attributes

```{r}

#full dataset:
df<-P_dataset

num_vars <- names(df)[sapply(df, is.numeric) & names(df) != dep_var]

# Open a PDF device (all plots go here, one per page)
pdf(paste0(dir.output, model.input, "_scatterplots_", dep_var, ".pdf"))

# Loop through numeric variables and plot
walk(num_vars, function(num_var) {
  
  # Fit linear model
  formula <- as.formula(paste(dep_var, "~", num_var))
  model <- lm(formula, data = df)
  r2 <- summary(model)$r.squared
  
  # Create scatterplot with R² annotation
  p <- ggplot(df, aes_string(x = num_var, y = dep_var)) +
    geom_point(color = "steelblue", alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = paste("Scatterplot of", dep_var, "vs", num_var)) +
    annotate("text", 
             x = Inf, y = Inf, 
             label = paste0("R² = ", round(r2, 3)), 
             hjust = 1.1, vjust = 1.5, size = 5, color = "black") +
    theme_minimal()
  
  print(p)  # each plot on a new page
})

# Close PDF device
dev.off()
```

### c. Manuscript figure (Appendix): Examine relationships among parent material: glacial till vs loess

"pcteolfinews" = % of AOI area classified as lithology type: eolian sediment, fine-textured (glacial loess)

```{r}
P_dataset %>% order() %>% names()

P_dataset %>% 
 select(order(colnames(P_dataset))) %>% names()

#Parent material:

parent<-c("pctglactilclayws",
"pctglaclakefinews",
"pctglaclakecrsws",
"pcteolcrsws",
"pcteolfinews",
"pctglactilloamws",
"pctglactilcrsws",
"pctalluvcoastws",
"pctcolluvsedws",
"clayws",
"pctglactilcrscat",
"pctglactilclaycat")

soils<-P_dataset %>% dplyr::select(all_of(parent))

ggplot(soils)+
  geom_point(aes(pctglactilloamws, pcteolfinews))

ggplot(soils)+
  geom_point(aes(pctglactilclayws, pctglactilloamws))

reg<-with(soils, lm(pctglactilloamws ~ pctglactilclayws))
summary(reg)
```

# 4 Data Preparation

### 4.1 Remove any remaining administrative attributes:

(Note this is outdated and no longer needed)

```{r}
P_dataset<-P_dataset %>% dplyr::select(-c(comppct_r.x, comppct_r.y))

```

### 4.2 Option to remove x and y as predictors

```{r}
P_dataset<-P_dataset %>% dplyr::select(-x, -y)
```

### 4.3 Option to restrict to samples with TP \<1000 mg/kg

```{r}
P_dataset<-P_dataset %>% filter(P_mgkg<1000)

#Number of remaining samples and unique sites
P_dataset %>% nrow()
```

### 4.4 Option to remove outlier(s)

Note we did not remove outliers for the streambank model.

```{r}
# Calculate the 0.5th and 99.5th percentiles 
data_vector<-P_dataset$P_mgkg 

lower_bound <- quantile(data_vector, 0.005) 
lower_bound  

upper_bound <- quantile(data_vector, 0.995) 
upper_bound  

# Remove outliers 
data_without_outliers <- P_dataset %>%    
  filter(P_mgkg>lower_bound & P_mgkg<upper_bound) 
nrow(data_without_outliers)

#Option to set data as full dataset, or without outliers
#without outliers:
P_dataset<-data_without_outliers

summary(P_dataset$P_mgkg)
```

### 4.5 Omit attributes with mostly NA values or with no information (standard deviation of 0)

Machine learning models can't handle NA, so need to either drop attributes that have NA or impute missing values (below)

BUG: this may not be dropping columns that have blank rather than NA??

```{r}


#Identify columns containing NA values:
cols_with_na <- P_dataset %>%
  dplyr::select_if(~any(is.na(.))) %>%
  colnames()

# Print the names of columns with NA values
print(cols_with_na)

#Identify columns that are >50% NA values

cols_to_remove<-P_dataset %>%
  dplyr::select_if(~ mean(is.na(.)) > 0.5) %>% 
  names()

print(cols_to_remove)

#Remove columns that are >50% NA

P_dataset<-P_dataset %>% 
  dplyr::select(-all_of(cols_to_remove))

names(P_dataset)

#ID columns with no information (numeric)
cols_no_info<-P_dataset %>% 
   dplyr::select(where(is.numeric)) %>% 
dplyr::select(where(~ sd(.x, na.rm = TRUE) == 0)) %>% names()

print(cols_no_info)

#Omit predictors with no information (standard deviation of 0)
P_dataset <-P_dataset %>% 
  dplyr::select(-all_of(cols_no_info))

```

### 4.6 Drop or reclassify factor levels with only one representative (one data point):

ID factors that have a level with only one data point. If these factors are included ML models typically won't run.

```{r}
cols_with_single_point_levels <- P_dataset %>%
  dplyr::select(where(is.factor)) %>% # Select only factor columns
  sapply(function(x) any(table(x) == 1)) %>% # Check if any level has count 1
  names(.)[.] # Get the names of columns where the condition is TRUE

cols_with_single_point_levels
```

Reclassify or drop factors:

```{r}


#Drop remaining attributes
cols_with_single_point_levels <- P_dataset %>%
  dplyr::select(where(is.factor)) %>% # Select only factor columns
  sapply(function(x) any(table(x) == 1)) %>% # Check if any level has count 1
  names(.)[.] # Get the names of columns where the condition is TRUE

cols_with_single_point_levels

P_dataset<-P_dataset %>% 
  dplyr::select(-any_of(cols_with_single_point_levels))

names(P_dataset)
```

### 4.7 Identify & remove highly correlated predictors

While xgboost and other ML models are typically robust to multi-collinearity, reducing the number of features in the model by eliminating redundant/correlated variables can help with 1) model training time, 2) interpretation of the model in terms of variable importance and 3) may offer some performance gains.

Note: you can only do this for numeric variables where data isn't missing! So will need to exclude NA rows and categorical variables

Use the findCorrelation() function from the caret package to remove correlated variables while keeping the one of a correlated pair:

- From the function documentation: "If two variables have a high correlation, the function looks at the mean absolute correlation of each variable and removes the variable with the largest mean absolute correlation." In other words, it considers both negative and positive correlations by using the absolute value

- the function chooses one of the two variables from a pair based on how correlated it is with all the other variables

- Note: this correlation matrix includes the dependent variable, so need to think about if you want to include that

```{r}
# Calculate the correlation matrix

set.seed(123)
cor_matrix <- P_dataset %>% 
  dplyr::select(where(is.numeric)) %>% #ignore comid and siteid
  dplyr::select(-c(comid, siteID)) %>% 
  na.omit() %>% 
  dplyr::select(where(~ sd(.) > 0)) %>% 
  cor(use = "pairwise.complete.obs") # 'use' argument handles remaining NAs safely

# View highly correlated pairs

cor_matrix_view <- cor_matrix 

# Set upper triangle to NA in the copy only
cor_matrix_view[upper.tri(cor_matrix_view, diag = TRUE)] <- NA

# Flatten and filter the copy
high_correlations <- cor_matrix_view %>%
  as.table() %>%
  as.data.frame() %>%
  filter(!is.na(Freq)) %>%          # Remove the NAs we just made
  mutate(Freq_abs = abs(Freq)) %>% 
  filter(Freq_abs > 0.8) %>% 
  arrange(desc(Freq_abs))

#View(high_correlations)

View(high_correlations %>% filter(Var2=="pctglactilloamws"))

#Use findCorrelation() to drop highly correlated variables - 
#first ID indices of highly correlated predictors

highly_correlated_indices <- findCorrelation(cor_matrix, cutoff = 0.80, exact = FALSE)

# Extract names to remove
cor_cols_to_remove <- colnames(cor_matrix)[highly_correlated_indices]
print(cor_cols_to_remove)

#Remove highly correlated variabes
P_dataset <- P_dataset %>%
  dplyr::select(-all_of(cor_cols_to_remove))

P_dataset %>% 
    dplyr::select(order(colnames(.))) %>% 
  names()

```

# 5 Split data into training and testing

For remaining attributes, impute missing values. But first, must split data into training and testing datasets (to prevent data leakage).

There are repeat samples collected from locations (at multiple depths) and multiple samples collected from the same NHD catchments (comid). This is a hierarchical (nested) structure. To prevent data leakage (ie, data in train and test datasets from same location and/or catchment), we need to split data by catchment (which will also account for repeat samples collected at depths).

We are also stratifying into train and test by mean P of samples in each catchment, so that values will end up in train and test that reflect the distribution of all P values

```{r}

#View(P_dataset)

#Summarize and manually create 3 distinct strata bins (e.g., Low, Med, High P_mgkg)

P_dataset$comid<-as.factor(P_dataset$comid)

catchment_summary <- P_dataset %>%
  dplyr::group_by(comid) %>%
  dplyr::summarize(strata_p = median(P_mgkg, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(p_bin = as.factor(ntile(strata_p, 3))) 

#Split data using the manual factor bins
set.seed(123)
summary_split <- group_initial_split(
  catchment_summary, 
  group  = comid, 
  strata = p_bin, 
  prop   = 0.75
)

#Pull the matched data rows from each comid back into train and test sets
train_catchments <- training(summary_split) %>% pull(comid)
test_catchments  <- testing(summary_split)  %>% pull(comid)

#Designate train and test datasets
P_train <- P_dataset %>% filter(comid %in% train_catchments)
P_test  <- P_dataset %>% filter(comid %in% test_catchments)

#Verify the distribution balance across your split
cat("Train Median P_mgkg:", median(P_train$P_mgkg, na.rm = TRUE), "\n")
cat("Test Median P_mgkg :", median(P_test$P_mgkg, na.rm = TRUE), "\n")

#check comids in train vs test

levels(factor(P_train$comid))
levels(factor(P_test$comid))

#check for overlap (shouldn't be any)
any(P_test$comid %in% P_train$comid)

any(test.prep$comid %in% P.Predictors$comid)

###########################
##OLDER METHOD
#without group splitting

set.seed(123)
#P_split <- group_initial_split(P_dataset, strata = P_mgkg)
#Note - Stratification ensures that the distribution of a variable is preserved across training and testing sets.

#P_train <- training(P_split)
nrow(P_train)
names(P_train)

#P_test <- testing(P_split)
nrow(P_test)
names(P_test)
```

# 6 Impute missing values:

```{r}
#Separate out predictor variables from dependent variable (P_mgkg):
#Also dropping bulk density for now
#Drop siteID and comid

Predictors<-P_train %>% 
  dplyr::select(-(c("P_mgkg")))

#R with parameters set to run faster

start_time <- Sys.time()
start_time
dataImputed <- missRanger(Predictors,
                          formula = . ~ . ,
                          num.trees = 100, 
                          verbose = 1, seed = 111)

end_time <- Sys.time()
end_time - start_time

nrow(dataImputed)
names(dataImputed)

#View(P_train)
#View(dataImputed)

#rematch to soil P
P.Predictors<-cbind(P_train$P_mgkg, dataImputed) 
P.Predictors<-P.Predictors %>% 
  dplyr::rename(P_mgkg = `P_train$P_mgkg`)

#View(P.Predictors)

###
#Impute *testing data* separately (for use in model testing later)
#Combine with predictor attributes to have more data to impute missing values

dataImputed<-P.Predictors %>% 
  dplyr::select(-P_mgkg)
names(dataImputed)


Test.impute<-rbind(
  (P_test %>% 
    dplyr::select(-(c("P_mgkg")))),
dataImputed)
nrow(Test.impute)

start_time <- Sys.time()
start_time
data.Test.Imputed <- missRanger(Test.impute,
                                formula = . ~ . ,
                                num.trees = 100, 
                                verbose = 1, seed = 111)

end_time <- Sys.time()
end_time - start_time

#select out test data; and attach to soil P
test<-data.Test.Imputed[c(1:nrow(P_test)),]
P.test<-P_test %>% dplyr::select(P_mgkg)
P.test
test.prep<-cbind(P.test, test)
nrow(test.prep)

View(P_test)
View(test.prep)

names(P_test)
names(test.prep)
```

## 6.1 Remove factors that only have 1 data point in any level

This is to ensure the model does not crash because there is a level in the test dataset that does not occur in the training dataset.

We also need to check if all levels of factors in the test data occur in the training dataset.

```{r}
#Do all factor levels in test dataset have an example in train dataset?


#eg, all factor levels in test data should occur in train data:
levels(factor(test.prep$NLCD_Land))
levels(factor(P.Predictors$NLCD_Land))

#Check if all levels of all factors in test data occur in train data: 
df1<-test.prep
df2<-P.Predictors

# Identify factor columns in df1
factor_cols_df1 <- names(df1)[sapply(df1, is.factor)]
factor_cols_df1

#ignore comid
factor_cols_df1 <- setdiff(names(df1)[sapply(df1, is.factor)], "comid")
factor_cols_df1

results <- lapply(factor_cols_df1, function(col_name) {
  if (col_name %in% names(df2) && is.factor(df2[[col_name]])) {
    levels_not_in_df2 <- setdiff(levels(df1[[col_name]]), levels(df2[[col_name]]))
    if (length(levels_not_in_df2) > 0) {
      return(paste0("Levels in '", col_name, "' of df1 not found in df2: ",
                    paste(levels_not_in_df2, collapse = ", ")))
    } else {
      return(paste0("All levels of '", col_name, "' in test dataset are present in predictor dataset."))
    }
  } else {
    return(paste0("Column '", col_name, "' is not a factor in predictor dataset or does not exist in predictor dataset."))
  }
})

# Print the results
for (res in results) {
  print(res)
}


#ID factors in training data with only a single data point in a given level: 

train_with_single_point_levels <- P.Predictors %>%
  dplyr::select(where(is.factor)) %>% # Select only factor columns
  dplyr::select(-comid) %>% #ignore comid
  sapply(function(x) any(table(x) == 1)) %>% # Check if any level has count 1
  names(.)[.] # Get the names of columns where the condition is TRUE
train_with_single_point_levels

#Drop factors from train and test with only a single data point in a given level 

P.Predictors <- P.Predictors %>%
  dplyr::select(-any_of(train_with_single_point_levels))

names(P.Predictors)

#Drop same factors from test data:

test.prep<-test.prep %>%
  dplyr::select(-any_of(train_with_single_point_levels))

names(test.prep)    

#check column match - should be 0 columns in test that are not in Predictors
check.col <- names(test.prep)[!names(test.prep) %in% names(P.Predictors)]
check.col
```

## 6.2 Write imputed and cleaned data to file

```{r}
#WRITE IMPUTED DATA TO FILE
#(note this is training data only)
write.table(P.Predictors, paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets", model.input, "_train_imputed_groupsplit_UPDATED_REDUCEDCOR_less1000_09142026.csv"), sep=",", row.names=FALSE)

write.table(test.prep, paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets", model.input, "_TEST_imputed_groupsplit_UPDATED_REDUCEDCOR_less1000_09142026.csv"), sep=",", row.names=FALSE)

```

## 6.3 Load saved input data (as needed)

For consistency, read in this data when you run model on same dataset

```{r}

#READ IMPUTED DATA
dir.final.model.data<-paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets")
#(note this is training data only)
P.Predictors<-read.csv(paste0(dir.final.model.data, model.input, "_train_imputed_SOIL_08132026.csv"))

P.Predictors$NLCD_Land<-factor(P.Predictors$NLCD_Land)

names(P.Predictors)

#View(P.Predictors)

#(test data)
test.prep<-read.csv(paste0(dir.final.model.data, model.input, "_TEST_imputed_SOIL_08132026.csv"))

test.prep$NLCD_Land<-factor(test.prep$NLCD_Land)

names(test.prep)
#View(test.prep)

#SET ATTRIBUTE TYPES
#Set categorical attributes to factor
#and set ID attributes to factor
cat_vars <- names(P.Predictors)[sapply(P.Predictors, function(x) is.factor(x) | is.character(x))]
cat_vars

P.Predictors<-P.Predictors %>% 
  dplyr::mutate(across(cat_vars, as.factor)) |> 
  dplyr::mutate(across(c(siteID, comid), as.factor))
test.prep<-test.prep %>% 
  dplyr::mutate(across(cat_vars, as.factor)) |> 
  dplyr::mutate(across(c(siteID, comid), as.factor))


#Set remaining numeric and integer attributes to numeric
P.Predictors<-P.Predictors |> 
dplyr::mutate(across(where(is.numeric), as.numeric))

test.prep<-test.prep |> 
  dplyr::mutate(across(where(is.numeric), as.numeric))


sapply(P.Predictors, class)

sapply(test.prep, class)

#Check to make sure columns compare
library(janitor)
compare_df_cols(P.Predictors, test.prep, return = "mismatch")
```

## 6.4 Option to combine streambank and upland data into training dataset

Need to assign attributes to indicate streambank vs upland

Need to weight streambank samples if predicting streambank TP

In order to get predictors to match

Steps would be:

-add imputed upland predictors

-add raw streambank

-fix nlcd factor issues

-fix wetland issues

\- split streambank into train and test by comid

-combine streambank train and upland

-impute streambank train and upland

-impute streambank test

```{r}

#READ IMPUTED STREAMBANK AND UPLAND DATA
dir.final.model.data<-paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets")
#(note this is training data only)

P.Predictors.streambank<-read.csv(paste0(dir.output, "/Margenot_sample_attributes_for_modelling_wSSURGO_LIMITED_comid.csv"))

P.Predictors.upland<-read.csv(paste0(dir.final.model.data, "/USGS_NCSS_soils_database_attributes_alldepths_for_modelling_wSSURGO_LIMITED_comid_train_imputed_less1000_groupsplit_08252026.csv"))

#For Margenot Streambank samples, get NLCD levels to match soil TP dataset:

#Match categorical levels where needed
levels(factor(P.Predictors.upland$NLCD_Land))
levels(factor(P.Predictors.streambank$NLCD_Land))


P.Predictors.streambank$NLCD_Land<-factor(P.Predictors.streambank$NLCD_Land,
                                          labels=c("Barren Land", "Cultivated Crops", "Deciduous Forest", "Developed, Low Intensity", "Developed, Open Space", "Emergent Herbaious Wetlands", "Mixed Forest", "Open Water", "Pasture/Hay", "Woody Wetlands"))


levels(factor(P.Predictors.upland$WETLAND_TYPE))
levels(factor(P.Predictors.streambank$WETLAND_TYPE))

#Set non wetland values to match soil TP dataset:

P.Predictors.upland$WETLAND_TYPE<-factor(P.Predictors.upland$WETLAND_TYPE, levels=c(
  "",                                               "Freshwater Emergent Wetland",                      "Freshwater Forested/Shrub Wetland",
  "Freshwater Pond",                                   "Lake",                                            "Riverine"),  labels=c(
    
    "Non wetland",                                 "Freshwater Emergent Wetland",                       "Freshwater Forested/Shrub Wetland", 
    "Freshwater Pond",                                  "Lake",                                             "Riverine"))


#Reformat siteID for upland dataset so doesn't overlap with streambank dataset
add.siteID<-max(P.Predictors.streambank$siteID)

P.Predictors.upland<-P.Predictors.upland %>% 
  mutate(siteID=siteID+add.siteID)

#Create sample ID column for streambank

sample.id<-max(P.Predictors.upland$X)

P.Predictors.streambank<-P.Predictors.streambank %>% 
  mutate(X=1:nrow(P.Predictors.streambank)+sample.id) 

#Make sure predictors match across datasets 
P.Predictors.streambank<-P.Predictors.streambank %>%
  { if ("meanP_mgkg" %in% names(.)) rename(., P_mgkg = meanP_mgkg) else . } %>% #rename P column if took average of subsamples
  dplyr::select(any_of(names(P.Predictors.upland))) #grab all columns that match P.Predictors 

#Do all columns match?
all(sort(names(P.Predictors.streambank)) == sort(names(P.Predictors.upland))) # should be TRUE


setdiff(names(P.Predictors.streambank), names(P.Predictors.upland))

#Make a factor for streambank sample or not:

P.Predictors.upland$is_streambank <- 0
P.Predictors.streambank$is_streambank <- 1



#Select out test dataset used to test other models

test.prep.orig<-test.prep<-read.csv(paste0(dir.final.model.data, "/Margenot_sample_attributes_for_modelling_wSSURGO_LIMITED_TEST_imputed_nosoil_08242026.csv"))

#Siphon off test data:
test.prep<-P.Predictors.streambank %>% filter(siteID %in% test.prep.orig$siteID)

nrow(test.prep)

streambank.train<-P.Predictors.streambank %>% filter(!siteID %in% test.prep.orig$siteID)

nrow(streambank.train)


# Combine into single training dataset
P.Predictors <- bind_rows(P.Predictors.upland, streambank.train)



#Set categorical attributes to factor
cat_vars <- names(P.Predictors)[sapply(P.Predictors, function(x) is.factor(x) | is.character(x))]
cat_vars

P.Predictors<-P.Predictors %>% 
  mutate(across(cat_vars, as.factor))
test.prep<-test.prep %>% 
  mutate(across(cat_vars, as.factor))

#Set remaining numeric attributes to numeric
dep_var="P_mgkg"
num_vars <- names(P.Predictors)[sapply(P.Predictors, is.numeric) & names(P.Predictors) != dep_var]
num_vars

P.Predictors <- P.Predictors %>%
  mutate(across(num_vars, as.numeric))
test.prep<-test.prep %>% 
  mutate(across(num_vars, as.numeric))

sapply(P.Predictors, class)

sapply(test.prep, class)

#Check to make sure columns compare
library(janitor)
compare_df_cols(P.Predictors, test.prep, return = "mismatch")

#---------------------------------------------
#Impute missing values from streambank dataset
#---------------------------------------------
P.impute<-P.Predictors %>% 
  dplyr::select(-any_of(c("P_mgkg", "BD_gcm3")))

#R with parameters set to run faster

start_time <- Sys.time()
start_time
dataImputed <- missRanger(P.impute,
                          formula = . ~ . ,
                          num.trees = 100, 
                          verbose = 1, seed = 111)

end_time <- Sys.time()
end_time - start_time

nrow(dataImputed)
names(dataImputed)

#rematch to soil P
P.Predictors.final<-cbind(P.Predictors$P_mgkg, dataImputed) 
P.Predictors.final<-P.Predictors.final %>% 
  dplyr::rename(P_mgkg = `P.Predictors$P_mgkg`)

#set back to P.Predictors
P.Predictors<-P.Predictors.final
#View(P.Predictors)

###
#Impute *testing data* separately (for use in model testing later)
#Combine with predictor attributes to have more data to impute missing values

dataImputed<-P.Predictors %>% 
  dplyr::select(-P_mgkg)
names(dataImputed)


Test.impute<-bind_rows(
  (test.prep %>% 
     dplyr::select(-any_of(c("P_mgkg", "BD_gcm3")))),
  dataImputed)
nrow(Test.impute)

start_time <- Sys.time()
start_time
data.Test.Imputed <- missRanger(Test.impute,
                                formula = . ~ . ,
                                num.trees = 100, 
                                verbose = 1, seed = 111)

end_time <- Sys.time()
end_time - start_time

#select out test data; and attach to soil P
test<-data.Test.Imputed[c(1:nrow(test.prep)),]
P.test<-test.prep %>% dplyr::select(P_mgkg)
P.test
test.prep<-cbind(P.test, test)
nrow(test.prep)

View(P_test)
View(test.prep)

names(P.Predictors)
names(test.prep)

#------------------------------------------
# Compute weight ratio: boost streambank so total influence is equal
n_upland <- sum(P.Predictors$is_streambank == 0)
n_streambank <- sum(P.Predictors$is_streambank == 1)
weight_multiplier <- n_upland / n_streambank

P.Predictors <- P.Predictors %>%
  mutate(
    # Create importance weights column recognized by tidymodels
    case_wts = importance_weights(
      ifelse(is_streambank == 1, weight_multiplier, 1.0)
    ),
    is_streambank = as.factor(is_streambank)
  )


#--------------------------------------------
# Create recipe specifying case weights
P_recipe <- P.Predictors %>% 
  recipe(P_mgkg ~ .) %>%
  # Register case_wts as sample weights, not a predictor
  update_role(comid, siteID, X, new_role = "ID") %>%
  # Convert factor domain indicator to numeric (0/1) for tree splitting/multiplication
  step_mutate(is_sb_num = as.numeric(is_streambank) - 1) %>%
  # Generate interaction between domain indicator and ALL numeric predictors
  # (step_interact automatically scales across your >200 predictors)
  step_interact(terms = ~ all_numeric_predictors():is_sb_num) %>%
  # Remove the temporary numeric dummy if needed, keep factors standard
  step_dummy(all_nominal_predictors(), -all_outcomes()) %>%
  step_zv(all_predictors()) # Remove zero-variance terms if any interaction is static


# -------------------------------------------------------------------
# 3. Define XGBoost Model Specification
# -------------------------------------------------------------------
xgb_spec <- boost_tree(
  trees = 500,
  tree_depth = 4,             # Keep shallow (3-5) with >200 features to avoid overfitting
  learn_rate = 0.03,
  mtry = 120,           # Subsample 30% of features per split (critical for 200+ features)
  min_n = 10,
  loss_reduction = 0.01
) %>%
  set_engine("xgboost", objective="reg:tweedie") %>%
  set_mode("regression")

# -------------------------------------------------------------------
# 4. Create Workflow & Fit
# -------------------------------------------------------------------
xgb_workflow <- workflow() %>%
  add_recipe(P_recipe) %>%
  add_model(xgb_spec) %>%
  # Explicitly instruct tidymodels to pass weights to XGBoost
  add_case_weights(case_wts)

# Train model
fitted_xgb <- fit(xgb_workflow, data = P.Predictors)

# -------------------------------------------------------------------
# 5. Evaluate on Independent Streambank Test Set
# -------------------------------------------------------------------
# Independent test set (Streambank only)
test.prep <- test.prep %>%
  mutate(
    is_streambank = as.factor(1)
  )

results <- test.prep %>%
  bind_cols(predict(fitted_xgb, new_data = test.prep)) %>%
  metrics(truth = P_mgkg, estimate = .pred)

print(results)
```

# 7 Alternative Models

## 7.0 Mean value as benchmark

Use mean value of training dataset as prediction for test dataset, evaluate against test data.

```{r}

#Mean value of train dataset as prediction for test dataset


Model_MEAN<-test.prep %>% dplyr::select(P_mgkg) %>% 
  mutate(Predicted_P=mean(P.Predictors$P_mgkg))

head(Model_MEAN)

write.csv(Model_MEAN %>% dplyr::select(P_mgkg, Predicted_P), paste0(dir.final.model, "/Model_predictions_test_data/MidwetModelMEAN_groupsplit_less1000.csv"), row.names=FALSE)


rmse <- with(Model_MEAN, sqrt(mean((P_mgkg - Predicted_P)^2)))
print(rmse)

```

## 7.1 Random Forest

#### a. Run Version of Original Model from Dolph et al 2023

```{r}
#Define recipe to use in both model types 

P_recipe_RF <- 
  # which consists of the formula (outcome ~ predictors)
  recipe(P_mgkg ~ ., data = P.Predictors) %>%
  step_center(all_numeric_predictors()) %>%
  step_scale(all_numeric_predictors()) %>% 
  update_role(comid, siteID, X, new_role = "ID")
  #Other options excluded for now, eg:
  #%>%
  #step_dummy(all_nominal()) #%>%
  #step_impute_bag(all_predictors(), trees=100) #option to impute missing values, using all other predictors
P_recipe_RF


P_train_preprocessed <- P_recipe_RF %>%
  # apply the recipe to the training data
  prep(P.Predictors) %>%
  # extract the pre-processed training dataset
  juice()
P_train_preprocessed
names(P_train_preprocessed)

#So need to bake() and not juice
P_test_preprocessed <- P_recipe_RF %>%
  # apply the recipe to the training data
  prep(P.Predictors) %>%
  # extract the pre-processed testing dataset
  bake(new_data=test.prep)
nrow(P_test_preprocessed)
names(P_test_preprocessed)

short_model_RF<-
  rand_forest(
    mtry = 50, 
    min_n=3, 
    trees = 1000) %>%
  set_mode("regression") %>%
  set_engine("randomForest")

#put into workflow
short_wf_RF<-workflow()%>%
  add_recipe(P_recipe_RF) %>%
  add_model(short_model_RF)


#run models
RF_fit<-short_wf_RF %>%
  fit(P_train_preprocessed)
RF_fit
summary(RF_fit)


#Check performance
names(P_test_preprocessed)

randfor_pred<-predict(RF_fit, new_data=P_test_preprocessed) 
#add predictions to test data:
RF.pred<-as.data.frame(cbind(P_test_preprocessed, randfor_pred))

names(RF.pred)

#check R2 value for actual vs predicted
reg.AO<- with(RF.pred,lm(P_mgkg~.pred))
summary(reg.AO)


RMSE = mean((RF.pred$P_mgkg - RF.pred$.pred)^2) %>% sqrt()
RMSE

#Variable Importance, previous method:
##RUN CONDITIONAL PERMUTATION IMPORTANCE:
#read in saved model if needed;


#Need to rerun model using randomforest() package

start_time <- Sys.time()
start_time

set.seed(542863)
rfP_model <- randomForest(P_mgkg ~ ., data = as.data.frame(P_train_preprocessed), mtry = 50, ntree=1000, replace = FALSE, 
                          nodesize = 3
                          , keep.forest = TRUE, keep.inbag = TRUE)
end_time <- Sys.time()
end_time - start_time

summary(rfP_model)

#Note: permimp does require training data to run!

library(permimp)

start_time <- Sys.time()
start_time

CPI_permimpRF <- permimp(rfP_model, 
conditional = TRUE, progressBar = TRUE, do_check=FALSE)

CPI_permimpRF
end_time <- Sys.time()
end_time - start_time

#Rank importance values and plot a subset:
#Make a dot plot:
library(dplyr)
library(ggplot2)
VarImp<-as.data.frame(sort(CPI_permimpRF$values, decreasing=TRUE), optional=T)
colnames(VarImp)[1]<-"Imp"
rownames(VarImp)
VarImp2 <- cbind(rownames(VarImp), data.frame(VarImp, row.names=NULL))
colnames(VarImp2)[1]<-"Var"
head(VarImp2)
View(VarImp)
VarImp2[1:75,]


#save permimp object!
saveRDS(CPI_permimpRF, paste0(dir.output, model.input, "_permimp_results_less1000.rds"))


```

#### b. Tune New Random Forest Models

```{r}

library(foreach)
library(ranger)


P_recipe_RF <- 
  # which consists of the formula (outcome ~ predictors)
  recipe(P_mgkg ~ ., data = P.Predictors) %>%
  #step_center(all_numeric_predictors()) %>%
  #step_scale(all_numeric_predictors()) %>% 
  #Other options excluded for now, eg:
  step_dummy(all_nominal()) %>% 
  update_role(comid, siteID, new_role = "ID")

doParallel::registerDoParallel()

tune_spec <- rand_forest(
  mtry = tune(),
  trees = 1000,
  min_n = tune()
) %>%
  set_mode("regression") %>%
  set_engine("ranger")

#specify cross validation for the training dataset, note that default is k=10 folds
#split groups by comid to prevent data leakage (will also cover repeat measures at a specific location)
#set.seed(234)
#P_folds <- vfold_cv(P.Predictors) #original folds, included minor data leakage for repeat measurements at same location (checked and did not fundamentally alter findings, but is technically incorrect)

set.seed(234)
P_folds <- group_vfold_cv(P.Predictors, group = comid, v = 10)

###
#PUT EVERYTHING TOGETHER IN A WORKFLOW

tune_wf <- workflow() %>%
  add_recipe(P_recipe_RF) %>%
  add_model(tune_spec)


###
#TUNE THE PARAMETERS

#Check start time for initial tuning
start_time <- Sys.time()
start_time

set.seed(345)
tune_res <- tune_grid(
  tune_wf,
  resamples = P_folds,
  grid = 20,
  control = control_grid(save_pred = TRUE, verbose=TRUE)
)
tune_res


#Check end time for initial tuning: 
end_time <- Sys.time()
end_time - start_time #took ~3 min


#Check out initial tuning results
tune_res %>%
  collect_metrics() %>%
  filter(.metric == "rsq") %>%
  select(mean, min_n, mtry) %>%
  pivot_longer(min_n:mtry,
               values_to = "value",
               names_to = "parameter") %>%
  arrange(mean) %>%
  print(n=40) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "rsq")

foreach::registerDoSEQ()

#find the best set:
best_rmse1<-select_best(tune_res, metric="rmse")
best_rmse1


####
#FINALIZE THE WORKFLOW

final_rf <- finalize_model(
  tune_spec,
  best_rmse1
)

final_rf

final_wf <- workflow() %>%
  add_recipe(recipe(P_mgkg ~ ., data = P.Predictors)) %>%
  add_model(final_rf)

###
#FIT THE MODEL BASED ON TRAINING DATA
rf_fit <- final_wf %>%
  fit(P.Predictors)
rf_fit

#Save the final model 
# Save the trained model object
saveRDS(rf_fit, file = paste0(dir.final.model, model.input, "_RF_model_LIMITED_stepdummy_group_split_083126.rds"))

```

#### c. Check RF Model Performance

```{r}
#Predict on the independent test set
pred_test <- predict(rf_fit, new_data = test.prep)
pred_test 
                    
Test_predictions<-cbind(test.prep, pred_test)%>% rename(Predicted_P=`.pred`)
names(Test_predictions)

#Write predictions to file
write.csv(Test_predictions %>% dplyr::select(P_mgkg, Predicted_P), paste0(dir.final.model, "/Model_predictions_test_data/STREAMBANK_RF_groupsplit_full_nostrata_nosoil.csv"), row.names=FALSE)

#Calculate R-squared
rsq_test <- Test_predictions %>%
  rsq(truth = P_mgkg, estimate = Predicted_P)

print(rsq_test)

rmse <- with(Test_predictions, sqrt(mean((P_mgkg - Predicted_P)^2)))
print(rmse)

#Scatter plot: actual vs predicted
test.plot<-ggplot(Test_predictions, aes(x = P_mgkg, y = Predicted_P)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    x = "Observed soil TP (mg/kg)",
    y = "Predicted soil TP (mg/kg)",
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test$.estimate, 3), ")")
  ) +
  theme_minimal()+
  xlim(0, 1500)+
  ylim(0,1500)
test.plot 

#Save to plot to file
png(paste0(file.name=dir.figures, "/Fig_pred_vs_obs_xgboost_reduced_with_WS_included.png"), width = 4, height = 4, units="in", res = 300)    

grid.newpage() 
print(test.plot) 

dev.off() 


#Check for samples with P<1000
rsq_test.less1000 <- Test_predictions %>% filter(P_mgkg<1000) %>%
  rsq(truth = P_mgkg, estimate = .pred)
rsq_test.less1000

rmse.less1000<-Test_predictions %>% 
    filter(P_mgkg<1000) %>% 
    mutate(rmse=sqrt(mean((P_mgkg - .pred)^2))) %>% 
    dplyr::select(rmse) %>% distinct()
rmse.less1000


#Plot Test data vs Predicted for samples <1000 mg/kg TP
ggplot(Test_predictions %>% filter(P_mgkg<1000), aes(x = P_mgkg, y = .pred)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    x = "Actual P_mgkg",
    y = "Predicted P_mgkg",
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test.less1000$.estimate, 3), ")")
  ) +
  theme_minimal()+
  xlim(0,1000)+
  ylim(0,1000)
```

#### d. Variable Importance

```{r}
#Run Model in randomForest
library(randomForest)

start_time <- Sys.time()
start_time

#check parameter values
best_rmse1

set.seed(542863)
rfP_model <- randomForest(mean.SRP ~ ., data = as.data.frame(P_train_preprocessed), 
                          mtry = best_rmse1$mtry, ntree=1000, replace = FALSE, #specify mtry and nodesize(same as min_n)
                          nodesize = best_rmse1$min_n,
                          keep.forest = TRUE, keep.inbag = TRUE)
end_time <- Sys.time()
end_time - start_time

summary(rfP_model)

#SAVE FINAL RANDOM FOREST MODEL!
saveRDS(rfP_model, "./FINALMODEL_randomForest_latesummer.rds")

#LOAD IN MODEL (if needed)
rfP_model<-readRDS("./FINALMODEL_randomForest_latesummer.rds")

print(rfP_model)
names(rfP_model)

rfP_model$mtry

#Check performance of the model (should match performance of the random_Forest() model tuned above, using tidymodels?)
rf.pred<-predict(rfP_model, newdata=P_test_preprocessed)
head(rf.pred)

#Match actual to predicted data:
predicted.P<-cbind(as.data.frame(P_test_preprocessed), rf.pred)
head(predicted.P)

#actual vs predicted
reg.check<-with(predicted.P, lm(mean.SRP~rf.pred))
summary(reg.check)
#calculate rmse
RMSE = mean((predicted.P$mean.SRP - predicted.P$rf.pred)^2) %>% sqrt()
RMSE

names(predicted.P)
names(P.Predictors)


##RUN CONDITIONAL PERMUTATION IMPORTANCE:
#read in saved model if needed;
setwd(output_dir)
rfP_model<-readRDS("./FINALMODEL_randomforest_latesummer.rds")
rfP_model

#Note: permimp does require training data to run!

start_time <- Sys.time()
start_time

CPI_permimpRF <- permimp(rfP_model, 
                         conditional = TRUE, progressBar = TRUE, do_check=FALSE)

CPI_permimpRF
end_time <- Sys.time()
end_time - start_time #took 14 minutes

#save permimp object!
saveRDS(CPI_permimpRF, "./permimp_results_latesummer.rds")


#read in permimp results if needed
CPI_permimpRF<-readRDS("./permimp_results_latesummer.rds")


#Rank importance values and plot a subset:
#Make a dot plot:
#library(dplyr)
#library(ggplot2)
VarImp<-as.data.frame(sort(CPI_permimpRF$values, decreasing=TRUE), optional=T)
colnames(VarImp)[1]<-"Imp"
rownames(VarImp)
VarImp2 <- cbind(rownames(VarImp), data.frame(VarImp, row.names=NULL))
colnames(VarImp2)[1]<-"Var"
head(VarImp2)
View(VarImp)
VarImp2[1:75,]
```

## 7.2 XGBoost

### a. Tune XGBoost Model

Helpful resources:

<https://juliasilge.com/blog/xgboost-tune-volleyball/>

Note that grid_space_filling() is the newer preferred way to create space-filling hyperparameter grids, it uses a latin hypercube by default

XGBoost model is based on trees - don’t need much preprocessing for our data, ie don’t need to worry about factors or centering or scaling our data. Can just go straight to setting up our model specification.

Note that in tidy models, "xgboost does not have a means to translate factor predictors to grouped slplits. Factor/categorical predictors need to be converted to numeric values (dummy or indicator variables) for this engine. When using the formula via fit.model_spec(), parsnip will convert factor columns to indicators using a one-hot encoding"

Citation:<https://parsnip.tidymodels.org/reference/details_boost_tree_xgboost.html>

```{r}

#Specify model type and hyperparameters to tune 

xgb_spec <- boost_tree(
  trees = 1000,
  tree_depth = tune(), min_n = tune(),
  loss_reduction = tune(), ## first three: model complexity
  sample_size = tune(), mtry = tune(),  ## randomness
  learn_rate = tune()                   ## step size
) %>%
  set_engine("xgboost", objective="reg:tweedie") %>%
  set_mode("regression")

xgb_spec
```

Because there are a lot of hyperparameters to tune, reduce computational complexity by setting up possible values for hyperparameters that cover the hyperparameter space.

Note on setting the size of the search space: \~ 10 observations per parameter being tuned are needed to adequately cover the search space - **need a citation for this!**

Here we are tuning 6 parameters (`tree_depth`, `min_n`, `loss_reduction`, `sample_prop`, `mtry`, `learn_rate`)

Recommended Size: 6 x 10 = 60.

```{r}

#NEW WAY 
#Set up hyperparameter space to search 

set.seed(123)
xgb_grid <- grid_space_filling(
  parameters(
    tree_depth(),
    min_n(),
    loss_reduction(),
    sample_prop(),
    finalize(mtry(), P.Predictors),
    learn_rate()
  ),
  size = 60
)

xgb_grid
```

#Option to both stratify by P and split by group

```{r}
#Create stratification bins - median P values by comid 

catchment_lookup <- P.Predictors %>%
  group_by(comid) %>%
  # Use na.rm = TRUE to completely ignore missing data during the calculation
  summarize(catchment_p_med = median(P_mgkg, na.rm = TRUE), .groups = "drop") %>%
  # Create the categorical bins explicitly on clean, summarized data
  mutate(comid_strata_bin = as.factor(ntile(catchment_p_med, 3))) %>%
  # Keep only the ID and the clean stratum bin column
  select(comid, comid_strata_bin)

# 2. Join the clean, non-variable bin column back into your main training data
train_data_prepared <- P.Predictors %>%
  inner_join(catchment_lookup, by = "comid")

#View(train_data_prepared %>% dplyr::select(P_mgkg, comid, siteID, comid_strata_bin))
```

Create cross validation resamples for tuning, while account for repeat samples in same catchment - see explanation above for splitting data into training and test

```{r}

#Create cross-validation folds that do group splitting by comid (note: does not currently account for stratification in P_mgkg)
set.seed(123)
cv_folds <- group_vfold_cv(
  data   = P.Predictors, 
  group  = comid,            
  #strata = is_streambank,
  #strata = P_mgkg, 
  v      = 10                 
)

cv_folds

###OLDER METHOD
#does not split by group, but did stratify outcome uniformly into folds across values of P
#set.seed(123)
#P_folds <- vfold_cv(P.Predictors, strata = P_mgkg)

#P_folds

```

Note that I need to account for repeat samples by catchment in cross validation, as well as stratification of P values

Put model into a workflow

```{r}

#Original:
#xgb_wf <- workflow() %>%
#  add_formula(P_mgkg ~ .) %>%
#  add_model(xgb_spec)

#xgb_wf


#With recipe (for use with shapviz later): 
#specify comid and siteID as ID variables

P_recipe <- P.Predictors %>%
 recipe(P_mgkg ~ .) %>% 
  #update_role(comid, siteID, X, new_role = "ID") %>% #if X included as sample ID
  update_role(comid, siteID, new_role = "ID") %>% 
  #option to log transform outcome
  #step_log(P_mgkg, base = exp(1), skip = TRUE) %>%
 #step_integer(all_nominal()) #integer encode factors
 step_dummy(all_nominal_predictors(), one_hot = TRUE) 
#If pooling streambank and midwest datasets:
#P_recipe <- P.Predictors %>% 
#  recipe(P_mgkg ~ .) %>%
  # Register case_wts as sample weights, not a predictor
 # update_role(comid, siteID, X, new_role = "ID") %>%
  # Convert factor domain indicator to numeric (0/1) for tree splitting/multiplication
#  step_mutate(is_sb_num = as.numeric(is_streambank) - 1) %>%
  # Generate interaction between domain indicator and ALL numeric predictors
  # (step_interact automatically scales across your >200 predictors)
#  step_interact(terms = ~ all_numeric_predictors():is_sb_num) %>%
  # Remove the temporary numeric dummy if needed, keep factors standard
#  step_dummy(all_nominal_predictors(), -all_outcomes()) %>%
 # step_zv(all_predictors()) # Remove zero-variance terms if any interaction is static


#Model workflow:
xgb_wf <- workflow() %>%
  add_recipe(P_recipe) %>%
  add_model(xgb_spec) 
  #add_case_weights(case_wts) #only needed if you are adding weights by input dataset
```

Tune model:

-Update: takes \~1-2 hours for full USGS/USDA soil P database, 30 min for IL streambank samples, for a search space of 30 and full variables

-Update 2: for a reduced dataset (dropping co-linear variables), and a larger search space (n=80), takes about 13-24 minutes for IL streambank samples

Note for small datasets: The following warning might arise if a dataset is so small that trees can't split according to the hyperparameter values in parameter space: 'A correlation computation is required, but 'estimate' is constant and has 0 standard deviation, resulting in a divide by 0 error. 'NA' will be returned'

See below for a possible alternative solution for small datasets (Bayesian tuning)

```{r}


#Tune model: 
start.time<-Sys.time()
start.time

doParallel::registerDoParallel()

set.seed(234)
xgb_res <- tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid = xgb_grid,
  control = control_grid(extract = identity, save_pred = TRUE, verbose=TRUE)
)

xgb_res

end.time<-Sys.time()
end.time

total.time<-end.time-start.time
total.time

# Run this immediately after your model finishes tuning/fitting
foreach::registerDoSEQ()
```

Explore XGBoost model results:

```{r}
collect_metrics(xgb_res)
```

Visualize tuning results:

```{r}
names(collect_metrics(xgb_res))

View(collect_metrics(xgb_res))

xgb_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  dplyr::select(mean, mtry:sample_size) %>%
  pivot_longer(mtry:sample_size,
               values_to = "value",
               names_to = "parameter"
  ) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "RMSE")
```

Find best set of parameters:

```{r}
best_rmse<-show_best(xgb_res, metric="rmse", n=1)
best_rmse


best_rsq<-show_best(xgb_res, metric="rsq", n=1)
best_rsq
```

Finalize tunable workflow with the best parameters:

```{r}
final_xgb <- finalize_workflow(
  xgb_wf,
  best_rmse
)

final_xgb
```

### b. Save final model object

```{r}
#set directories for final saved models + data

dir.final.model.data<-paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets")

dir.final.model<-paste0(dir.output, "/Final_model_comparisons")

#Run the final model:
set.seed(123)
final_fit <- final_xgb %>%
  fit(data = P.Predictors)

# Save the trained model object
saveRDS(final_fit, file = paste0(dir.final.model, model.input, "_xgb_model_LIMITED_stepdummy_full_SOIL_TWEEDIE_groupsplit_nostrat_RERUN_09212026.rds"))

```

### c. Load Saved Model (as needed)

```{r}

model.input

# Load saved model
#final_fit <- readRDS(paste0(dir.output, model.input, "_xgb_model_LIMITED_less1000_corReduced_stepdummy_021926.rds"))

final_fit<-readRDS(paste0(dir.final.model, model.input, "_xgb_model_LIMITED_stepdummy_full_SOIL_TWEEDIE_groupsplit_nostrat_08132026.rds"))

summary(final_fit)
#summary(final_fit_all)

```

### d. Check model hyperparameters

```{r}

#Different approaches for examing hyperparameter values:

print(final_fit) 

xgb_engine_fit <- extract_fit_engine(final_fit)

print(xgb_engine_fit$params)

```

### e. Specify XGB Model directly using explicit parameters

Based on previously tuned models (using above script)

```{r}


#Check to make sure train and test datasets are split by comid

any(test.prep$comid %in% P.Predictors$comid)


# Set seed for reproducibility
set.seed(123)

#Cross-Validation Setup (10-Fold Group CV by 'comid')
cv_folds <- group_vfold_cv(
  data  = P.Predictors,
  group = comid,
  v     = 10
)


#Model Specification - use previously tuned parameter values

#specs: best soil model
#119	1000	16	1	0.012	0.634	1.37
#specs: best no soil model 
#216	1000	30	11	0.00855	0.512	0.00000000569

xgb_spec <- boost_tree(
  #mtry = 119,
  #trees = 1000,
  #min_n = 16,
  #tree_depth = 1,
  #learn_rate = 0.012,
  #sample_size = 0.634,
  #loss_reduction = 1.37
  
  mtry = 216,
  trees = 1000,
  min_n = 30,
  tree_depth = 11,
  learn_rate = 0.00855,
  sample_size = 0.512,
  loss_reduction = 0.00000000569
) %>%
  set_engine(
    "xgboost",
    objective = "reg:tweedie"
  ) %>%
  set_mode("regression")

#Recipe Setup
xgb_recipe <- recipe(P_mgkg ~ ., data = P.Predictors) |> 
  update_role(siteID, comid, new_role = "ID") |>  # Keep ID variables out of predictors
  step_novel(all_nominal_predictors(), new_level = "new") |> 
  step_dummy(all_nominal_predictors())
 

#Workflow Assembly
xgb_workflow <- workflow() %>%
  add_recipe(xgb_recipe) %>%
  add_model(xgb_spec)

#Evaluate Out-of-Fold Performance via Group Cross-Validation
set.seed(123)
cv_resampling_results <- fit_resamples(
  xgb_workflow,
  resamples = cv_folds,
  metrics   = metric_set(yardstick::rmse, yardstick::rsq, yardstick::mae)
)

# Display cross-validation performance metrics across folds
cat("--- 10-Fold Group Cross-Validation Results ---\n")
print(collect_metrics(cv_resampling_results))

#Fit Final Model on Full Training Dataset
final_fit <- fit(xgb_workflow, data = P.Predictors)

#Predict on Independent Test Data using augment()
Test_predictions <- augment(final_fit, new_data = test.prep) %>%
  dplyr::rename(Predicted_P = .pred)

# Verify column names
names(Test_predictions)


#Calculate Independent Test Metrics
rsq_test <- Test_predictions %>%
  rsq(truth = P_mgkg, estimate = Predicted_P)

rmse_test <- Test_predictions %>%
  rmse(truth = P_mgkg, estimate = Predicted_P)

cat("\n--- Independent Test Set Results ---\n")
print(rsq_test)
print(rmse_test)


# 8. Export Predictions to CSV
write.csv(Test_predictions %>% dplyr::select(P_mgkg, Predicted_P), paste0(dir.final.model, "/Model_predictions_test_data/ModelTWEEDIE_groupsplit_nosoil_full_nostrata.csv"), row.names=FALSE)




#Scatter plot: actual vs predicted
test.plot<-ggplot(Test_predictions, aes(x = P_mgkg, y = Predicted_P)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test$.estimate, 2), ")")
  ) +
  xlab(bquote(Observed~TP~(mg~kg^-1)))+
  ylab(bquote(Predicted~TP~(mg~kg^-1)))+
  theme_minimal()+
  xlim(0, 2100)+
  ylim(0,2100)
test.plot 

#Save to plot to file
png(paste0(file.name=dir.figures, "/Fig_STREAMBANK_pred_vs_obs_xgboost_best_NOSOIL_model.png"), width = 4, height = 4, units="in", res = 300)    

grid.newpage() 
print(test.plot) 

dev.off() 

```

### f. Variable importance: XGBoost

#### i. Built in variable importance for XGBoost

Based on "Gain" - the improvement in accuracy (loss reduction) brought by a feature to the branches it is used in.It measures how much each feature contributes to the model's predictive power, with higher values indicating more important features. Can also evaluate based on Cover or Frequency

```{r}

#use vip() to automatically plot importance:

final_fit %>%
  extract_fit_parsnip() %>%
  vip(geom = "point", num_features = 10)+
  theme_bw()+
  theme(panel.grid=element_blank())+
  #scale_y_discrete(labels = my_labels)+
  xlab("Variable Importance")


#A more adjustable plot:


#extract model first 

xgb_booster_object <- extract_fit_engine(final_fit)  

#importance measures
importance <- xgb.importance(model = xgb_booster_object) 

importance

#ID top 10 most important predictors 
top_vars <- importance$Feature[1:10] 
top_vars 

#make nicer labels
#Cat scale predictors only:
labels = list(
  "Water input",
  "% Glacial till, loamy (watershed)",
  "Soil erodibility factor",
  "Depth (cm)",
  "% Glacial till, loamy (catchment)",
  "WWTP density",
  "% Ag drainage",
  "% Hay",
  "Hydraulic conductivity",
  "% Mixed forest"
              
            #"Net anthropogenic N inputs",
             # "Annual min air temp, 1981-2010",
              #"Mean topographic wetness index",
              #"Sediment regulation index",
              #"Elevation",
              #"Slope Length - USLE",
              #"Population density",
              #"Soil organic carbon, 100-150cm"
)

#Watershed and Cat scale predictors:
labels = list("Depth",
              "Water input",
              "WWTP density",
              "Mean annual precip",
              "% ag on high slopes",
              "Elevation",
              "% Glacial till, loamy",
              "Soil erodibility",
              "N from rock weathering",
              "Water withdrawals"
              )


#select top 10 by gain and assign better labels
imp.data<-importance %>% 
  arrange(desc(as.numeric(Gain))) %>% 
   slice_max(Gain, n = 10)

imp.data$labels<-as.character(labels)
imp.data


#Make nicer plot: 
imp.plot<-ggplot(imp.data)+
  geom_point(aes(x=Gain, y = reorder(labels, Gain)), size=3)+
  theme_bw()+
  theme(panel.grid=element_blank(), axis.text=element_text(size=16), axis.title=element_text(size=18))+
  #scale_y_discrete(labels = my_labels)+
  xlab("Variable Importance")+
  ylab("Model Predictors")
imp.plot

#save plot to file

png(paste0(file.name=dir.figures, "/Fig_streambank_importance_plot_WSincluded_corReduced_less1000.png"), width = 2900, height = 1800, res = 300)    

grid.newpage() 
print(imp.plot) 

dev.off()  
```

#### ii. Shapley Importance Values

There are multiple packages/approaches for illustrating SHAP values:

1.  shapviz
2.  SHAPforxgboost
3.  xgboost has a native shap function (xgb.plot.shap)

```{r}

#####################################
#1. Using shapviz method
#install.packages("shapviz")
library(shapviz)


# The 'X_pred' argument in shapviz() is prepared predictor matrix
set.seed(123)
shap <- shapviz(extract_fit_engine(final_fit), 
                X_pred = bake(
      prep(xgb_recipe),
      has_role("predictor"),
      new_data = NULL,
      composition = "matrix"
    )
  )
shap

#View(shap)

#Importance plot for top predictors: 
sv_importance(shap, kind = "beeswarm", show_numbers = TRUE, max_display=15)

#Importance plot for all features for Appendix:
shap.plot.all<-sv_importance(shap, kind = "beeswarm", show_numbers = TRUE, max_display=Inf, number_size=5)+
  theme(axis.text=element_text(size=16), legend.text=element_text(size=16), legend.title=element_text(size=16))
shap.plot.all
#save plot to file

png(paste0(file.name=dir.figures, "/Fig_REVISED_streambank_SHAP_importance_plot_SOIL_TWEEDIE_groupslit_nostrata_09212026.png"), width = 4000, height = 6000, res = 300)    

grid.newpage() 
shap.plot.all
dev.off() 




#top 10 features with nicer labels, with site-specific soil included:
shap.plot1<-sv_importance(shap, kind = "beeswarm", show_numbers = TRUE, max_display=10)+
  scale_y_discrete(labels = c(
    "pct.sand" = "% sand",
    "pct.clay"= "% clay",
    "rocknws" = "N from rock weathering",
    "naniws" = "Anthropogenic N inputs",
    "tmax8110cat" = "Avg annual max air temp (catchment)",
    "bfiws" = "Baseflow index (watershed)",
    "pH" = "pH",
    "tmax8110ws" = "Avg annual max air temp (watershed)",
    "pwsl1pomu" = "% potential wetland soil in mapunit",
    "bficat" = "Baseflow index (catchment)"
  ))+
  theme_bw()
shap.plot1

#top 10 without site specific characteristics:
shap.plot2<-sv_importance(shap, kind = "beeswarm", show_numbers = TRUE, max_display=10)+
  scale_y_discrete(labels = c(
    "Depth_cm" = "Depth (cm)",
    "tmax8110cat" = "Avg annual max air temp (catchment)",
    "naniws" = "Anthropogenic N inputs",
    "tmax8110ws" = "Avg annual max air temp (watershed)",
    "pwsl1pomu" = "% potential wetland soil in mapunit", 
    "bficat" = "Baseflow index, catchment scale",
    "NLCD_Land_Deciduous.Forest" = "NLCD Deciduous Forest",
    "bfiws" = "Baseflow index (watershed)",
    "WETLAND_TYPE_Non.wetland" = "Non wetland",
    "waterinputcat" = "Water input"
    ))+
  theme_bw()
shap.plot2



#save plot to file

png(paste0(file.name=dir.figures, "/Fig_REVISED_streambank_SHAP_importance_plot_TWEEDIE_soil_groupsplit.png"), width = 2200, height = 1400, res = 300)    

grid.newpage() 
shap.plot1
dev.off()  

```

SHAP Dependency Plots:

```{r}
#Dependency plots using shapviz:

#For a specific variable:

sv_dependence(shap, v="pctwdwet2011ws", share_y=TRUE, color_var=NULL)+xlab("% woody wetlands")

#Original Streambank P model (no site specific features)

#plot3<-sv_dependence(shap, v="Depth_cm", share_y=TRUE, color_var=NULL)+xlab("Depth (cm)")

#plot4<-sv_dependence(shap, v="kffactws", share_y=TRUE, color_var=NULL)+xlab("Soil erodibility")

#plot5<-sv_dependence(shap, v="pctglactilloamcat", share_y=TRUE, color_var=NULL)+xlab("% Glacial till, loamy \n (catchment)")

#plot6<-sv_dependence(shap, v="hydricrating_No", share_y=TRUE, color_var=NULL)+ xlab("Not hydric soils")

#for some reason, this plot coming out incorrectly (are no negative values for wwtpmajordensws)
#plot5<-sv_dependence(shap, v="wwtpmajordensws", share_y=TRUE, color_var=NULL)+xlab("WWTP density")
summary(P.Predictors$wwtpmajordensws)
#plotting manually:
sv_dependence(shap, v="wwtpmajordensws", share_y=TRUE)

wwtp.shap.data<-as.data.frame(shap[["S"]]) %>% dplyr::select(wwtpmajordensws)
head(wwtp.shap.data)

og.data<-as.data.frame(shap[["X"]][["wwtpmajordensws"]])
head(og.data)
colnames(og.data)[1]<-"WWTP_dens"

wwtp.shap<-cbind(wwtp.shap.data, og.data)

#plot7<-ggplot(wwtp.shap)+
  geom_point(aes(WWTP_dens, wwtpmajordensws), col="#3b528b")+
  xlab("WWTP density")+
  ylab("SHAP value")
#plot7


#plot8<-sv_dependence(shap, v="precip2009ws", share_y=TRUE, color_var=NULL)+xlab("Mean Precipitation (mm)")

#plot9<-sv_dependence(shap, v="omws", share_y=TRUE, color_var=NULL)+xlab ("Organic Matter")

#plot10<-sv_dependence(shap, v = "rocknws", share_y = TRUE, color_var=NULL)+ xlab("N from rock weathering")

-----------------------
## Site Specific features included
    "pct.sand" = "% sand",
  "pct.clay"= "% clay",
  "rocknws" = "N from rock weathering",
  "naniws" = "Anthropogenic N inputs",
  "tmax8110cat" = "Avg annual max air temp (catchment)",
  "bfiws" = "Baseflow index (watershed)",
  "pH" = "pH",
  "tmax8110ws" = "Avg annual max air temp (watershed)",
  "pwsl1pomu" = "% potential wetland soil in mapunit",
  "bficat" = "Baseflow index, catchment scale"

plot1<-sv_dependence(shap, v="pct.sand", share_y=TRUE, color_var=NULL)+xlab("% Sand")
plot1

plot2<-sv_dependence(shap, v="pct.clay", share_y=TRUE, color_var=NULL)+xlab("% Clay")
plot2
  
plot3<-sv_dependence(shap, v="rocknws", share_y=TRUE, color_var=NULL)+xlab("N from rock weathering")
plot3

plot4<-sv_dependence(shap, v="naniws", share_y=TRUE, color_var=NULL)+xlab("Anthropogenic N inputs")
plot4
  
plot5<-sv_dependence(shap, v="tmax8110cat", share_y=TRUE, color_var=NULL)+xlab("Avg annual max air temp (watershed)")
plot5

plot6<-sv_dependence(shap, v="bfiws", share_y=TRUE, color_var=NULL)+xlab("Baseflow index (watershed)")
plot6

plot7<-sv_dependence(shap, v="pH", share_y=TRUE, color_var=NULL)+xlab("pH")
plot7

plot8<-sv_dependence(shap, v="tmax8110ws", share_y=TRUE, color_var=NULL)+xlab("Avg annual max air temp (watershed)")
plot8

plot9<-sv_dependence(shap, v="pwsl1pomu", share_y=TRUE, color_var=NULL)+xlab("% potential wetland soil in mapunit")
plot9

plot10<-sv_dependence(shap, v="bficat", share_y=TRUE, color_var=NULL)+ xlab("Baseflow index, catchment scale")
plot10



library(gridExtra)

grid.arrange(plot1,
plot2,
plot3,
plot4,
plot5,
plot6,
plot7,
plot8,
plot9,
plot10, nrow=5, ncol=2)

#save plot to file 
png(paste0(file.name=dir.figures, "/Fig_streambank_SHAP_dependency_tweedie_groupsplit_soil.png"), width = 1400, height = 3000, res = 300)    

grid.arrange(plot1,
plot2,
plot3,
plot4,
plot5,
plot6,
plot7,
plot8,
plot9,
plot10, nrow=5, ncol=2)

dev.off()

##
#Other plot types:

sv_force(shap, row_id = 1)
sv_waterfall(shap, row_id = 1)

#####################################
#2. Using SHAPforxgboost method
#(produces same results as above)

library(SHAPforxgboost)

game_shap <-
  shap.prep(
    xgb_model = extract_fit_engine(final_fit),
    X_train = bake(
      prep(P_recipe),
      has_role("predictor"),
      new_data = NULL,
      composition = "matrix"
    )
  )

shap.plot2<-shap.plot.summary(game_shap) #this shows SHAP values for all predictors 
shap.plot2

png(paste0(file.name=dir.figures, "/Fig_streambank_SHAP_importance_plot_2+WSincluded_corReduced_less1000.png"), width = 1800, height = 2900, res = 300)    
grid.newpage() 
shap.plot2
dev.off()  

##################################
#3. Using native xgboost method 

#One dependency plot for all top variables of interest
#**Note: this is showing slightly different variable importance order than previous two methods! **

xgb.plot.shap(data = bake(
      prep(P_recipe),
      has_role("predictor"),
      new_data = NULL,
      composition = "matrix"
    ), 
    model = extract_fit_engine(final_fit), top_n = 5)          
             
#save plot to file

png(paste0(file.name=dir.figures, "/Fig_streambank_SHAP_importance_plot_3_WSincluded_corReduced_less1000.png"), width = 1800, height = 2900, res = 300)    

grid.newpage() 
xgb.plot.shap(data = bake(
      prep(P_recipe),
      has_role("predictor"),
      new_data = NULL,
      composition = "matrix"
    ), 
    model = extract_fit_engine(final_fit), top_n = 15)  
dev.off()  


```

### f. Partial dependence plots:

Use DALEX for partial dependence plots when using tidymodels

Example: <https://juliasilge.com/blog/mario-kart/>

```{r}
#Use a model_explainer from DALEXtra to create partial dependence plots:  
model_explainer <- explain_tidymodels(   
  final_fit,   
  data = dplyr::select(P.Predictors, -P_mgkg),   
  y = P.Predictors$P_mgkg,   
  verbose = FALSE )  

#Plot PDP plots for top 10 most important variables:  
pdp_time <- model_profile(   
  model_explainer,   
  variables = top_vars,   
  N = NULL )  

plot(pdp_time)
```

Make fancier Partial Dependence Plots for manuscript:

```{r}

#Needs: variables in correct order
#be able to control plot aesthetics 

names(pdp_time)

#look at top 10 most important predictors:
top_vars

#Extract relationship between variables and predicted dependent variable:

PDP<-as.data.frame(pdp_time$agr_profiles)
names(PDP)

#Remove funky formatting and 
#Select top 10 variables:
PDP.sel<-PDP %>% 
  rename_with(~ str_remove_all(.x, "^_|_$"))%>% 
  filter(vname %in% top_vars)

#Make nicer labels for predictors and put them in right order: 
PDP.sel$PDP.label<-factor(PDP.sel$vname, levels=top_vars, labels=labels)
head(PDP.sel)

#Plot using ggplot: 

PDP.Plot<-ggplot(PDP.sel, aes(x = x, y = yhat)) +
  geom_line(color = "blue", linewidth = 1.2) + # Change line style here
  facet_wrap(~ PDP.label, scales = "free_x", ncol = 4, labeller = label_wrap_gen(width = 20)) + # 'free_x' allows different units; wrap strip text
  labs(
    y = "Predicted P (mg/kg)",
    x = "Predictor Value",
    title = "Partial Dependence Plots"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.background = element_rect(fill = "gray95"), # Box behind variable names
    strip.text = element_text(face = "bold", size = 10)
  )
PDP.Plot


#save plot to file

png(paste0(file.name=dir.figures, "/Fig_PDP_plot_streambank_WS_included_corReduced_less1000.png"), width = 2000, height = 2000, res = 300)    
grid.newpage() 
print(PDP.Plot) 

dev.off()  

```

### g. Alternative tuning method for small datasets - try a Bayesian tuning approach

Define search space with safer bounds for small dataset:

```{r}
param_set <- parameters(   
  tree_depth(range = c(2L, 6L)),   
  min_n(range = c(2L, 20L)),   
  loss_reduction(range = c(0, 5)),   # gamma, no negative   
  sample_size = sample_prop(range = c(0.6, 1.0)),   
  finalize(mtry(), P.Predictors),   
  learn_rate(range = c(0.01, 0.3)) )
```

Bayesian tuning:

Start with some random trials (`initial`) and then let Bayesian optimization adapt:

Note: this is slow, even for a small dataset

```{r}
set.seed(345) 
xgb_bayes <- tune_bayes(   
  xgb_wf,   
  resamples = P_folds,   
  param_info = param_set,   
  metrics = metric_set(rmse, mae),   
  initial = 10, # initial random search   
  iter = 20, # number of Bayesian iterations   
  control = control_bayes(verbose = TRUE, no_improve = 10, 
  save_pred=TRUE) )
```

Plot tuning improvements over iterations

```{r}
bayes_results <- xgb_bayes %>% collect_metrics() bayes_params <- xgb_bayes %>% collect_predictions()  xgb_bayes %>% 
  show_best(metric = "rmse", n = 20) 
# top-performing iterations  

# Plot RMSE over iterations 
bayes_results %>%   
  filter(.metric == "rmse") %>%   
  mutate(iter = 1:n()) %>%  # iteration order   
  
  ggplot(aes(x = iter, y = mean)) +   
  geom_point(color = "blue") +   
  geom_line(color = "darkblue") +   
  labs(     
    x = "Iteration",     
    y = "Mean RMSE",     
    title = "Bayesian Optimization Progress"   ) +   
  theme_minimal()
```

Pick best model:

```{r}
best_bayes <- select_best(xgb_bayes, metric = "rmse")
best_bayes 

final_xgb <- finalize_workflow(xgb_wf, best_bayes)  

final_fit <- fit(final_xgb, data = P.Predictors)

library(vip)  
final_fit %>%   
  extract_fit_engine() %>%   
  vip(geom = "point")
```

## 7.4 Linear model

For comparison, as a baseline. Using top variables from xgboost

(Though note there are other reasons why a linear model might not be great - if variables don't meet the assumptions, etc)

<https://rpubs.com/ajdowny_student/300663>

<https://online.stat.psu.edu/stat462/node/117/>

<https://blog.marketingdatascience.ai/linear-regression-vs-xgboost-which-predicts-sales-better-using-r-3bdf5aaab6bc>

### a. Run vanilla model

```{r}

#Load in data - see section 6.2 

names(P.Predictors)

#run vanilla linear model - tidymodels

#Create the linear model specification
lm_spec <- linear_reg() %>% 
  set_engine("lm") %>% 
  set_mode("regression")

#Create a recipe and add step_dummy for categorical predictors

#top 5 predictors, no soil model:
#"Depth_cm" = "Depth (cm)",
#"tmax8110cat" = "Avg annual max air temp (catchment)",
#"naniws" = "Anthropogenic N inputs",
#"tmax8110ws" = "Avg annual max air temp (watershed)",
#"pwsl1pomu" = "% potential wetland soil in mapunit", 

#top 5 predictors, soil model
#"pct.sand" = "% sand",
#"pct.clay"= "% clay",
#"rocknws" = "N from rock weathering",
#"naniws" = "Anthropogenic N inputs",
#"tmax8110cat" = "Avg annual max air temp (catchment)",

lm_recipe <- recipe(P_mgkg ~ pct.sand + pct.clay + rocknws + naniws + tmax8110cat, data = P.Predictors) %>% 
  step_dummy(all_nominal_predictors())

#Combine model and recipe into a workflow
lm_workflow <- workflow() %>% 
  add_model(lm_spec) %>% 
  add_recipe(lm_recipe)

#Fit the workflow to the data
lm_fit <- fit(lm_workflow, data = P.Predictors)

# View the trained model coefficients
tidy(lm_fit)


#predict using test data

lm_preds <- predict(lm_fit, new_data = test.prep)
head(lm_preds)

Test_predictionsLM<-cbind(test.prep, lm_preds) %>% dplyr::rename(Predicted_P=.pred)

names(Test_predictionsLM)

#Write predictions to file
write.csv(Test_predictionsLM %>% dplyr::select(P_mgkg, Predicted_P), paste0(dir.final.model, "/Model_predictions_test_data/ModelLM_groupsplit_SOIL_full.csv"), row.names=FALSE)

#Calculate R-squared
rsq_test <- Test_predictionsLM %>%
  rsq(truth = P_mgkg, estimate = Predicted_P)

print(rsq_test)

rmse <- with(Test_predictionsLM, sqrt(mean((P_mgkg - Predicted_P)^2)))
print(rmse)

#Scatter plot: actual vs predicted
LM.plot<-ggplot(Test_predictionsLM, aes(x = P_mgkg, y = Predicted_P)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test$.estimate, 2), ")")
  ) +
  xlab(bquote(Observed~soil~TP~(mg~kg^-1)))+
  ylab(bquote(Predicted~soil~TP~(mg~kg^-1)))+
  theme_minimal()+
  xlim(0, 2200)+
  ylim(0,2200)
LM.plot 

#Calculate all benchmarks
#Define metrics for comparison
metrics <- yardstick::metric_set(
  yardstick::rsq, 
  yardstick::rmse, 
  yardstick::mae, 
  yardstick::mape, 
  yardstick::mpe
)

#Evaluate metrics by model
evalLM <- Test_predictionsLM %>%
  metrics(truth = P_mgkg, estimate = Predicted_P)

evalLM
```

### b. Adjust and transform independent/dependent variables

```{r}

#Adjust and transform

#Log transform P_mgkg (for right skew)

lm_model2 <- lm(log(P_mgkg) ~ pct.silt+pct.clay+pctglactilloamws+pH+waterinputcat, data = P.Predictors)
summary(lm_model2)


#predict using test data

lm_preds2 <- predict(lm_model2, newdata = test.prep)
head(lm_preds2)

Test_predictionsLM2<-cbind(test.prep, lm_preds2) %>% rename(logPredicted_P=lm_preds2) %>% mutate(logP_mgkg=log(P_mgkg)) 


#Back transform predictions? note - can introduce bias


names(Test_predictionsLM2)

#Write predictions to file
write.csv(Test_predictionsLM %>% dplyr::select(P_mgkg, Predicted_P), paste0(dir.final.model, "/Model_predictions_test_data/ModelLM.csv"), row.names=FALSE)


#Calculate R-squared
rsq_test <- Test_predictionsLM2 %>%
  rsq(truth = logP_mgkg, estimate = logPredicted_P)

print(rsq_test)

rmse <- with(Test_predictionsLM2, sqrt(mean((logP_mgkg - logPredicted_P)^2)))
print(rmse)

#Scatter plot: actual vs predicted
LM.plot<-ggplot(Test_predictionsLM2, aes(x = logP_mgkg, y = logPredicted_P)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test$.estimate, 2), ")")
  ) +
  xlab(bquote(Observed~soil~TP~(mg~kg^-1)))+
  ylab(bquote(Predicted~soil~TP~(mg~kg^-1)))+
  theme_minimal()+
  xlim(5, 7.5)+
  ylim(5,7.5)
LM.plot 


#Other possible transformations:
P.Predictors_adjust<-P.Predictors %>% 
  mutate(pctglactilloamws_adjust=ifelse(pctglactilloamws==0, 0.001, pctglactilloamws)) %>% 
  mutate(pctglactilloamws_adjust=ifelse(pctglactilloamws==100, 99.999, pctglactilloamws_adjust)) %>% 
  mutate(pct.silt_adjust=ifelse(pct.silt==0, 0.001, pct.silt))

View(P.Predictors_adjust%>% dplyr::select(pct.silt_adjust, pct.clay, pctglactilloamws_adjust, pH, waterinputcat))

lm_model <- lm(P_mgkg ~ log(pct.silt_adjust)+log(pct.clay)+log(pctglactilloamws_adjust)+pH+waterinputcat, data = P.Predictors_adjust)
summary(lm_model)

```

### c. Check linear regression assumptions

```{r}

#Check assumptions

#Set model for checking assumptions

check.model<-lm_model2

#1. Non-linearity, unequal error variances, and outliers

#Residuals vs fitted plot

plot(check.model, 1)

#2. Independence of variables
corrplot(cor(Prestige[,c(1,2,4)]),method='circle')

#3. Normal distribution of residuals
plot(check.model, 2)

#4. Homoscedasticity (is error term the same across all values of the independent variable)

plot(check.model, 3)
```

## 7.5 Generalized Linear Model

for skewed, positive continuous data - use gamma family.

<https://rpubs.com/kaz_yos/glm-Gamma>

```{r}



#Create the linear model specification
glm_spec <- linear_reg() %>% 
  set_engine("glm", family = Gamma(link = "log")) %>% 
  set_mode("regression")

#Create a recipe and add step_dummy for categorical predictors


glm_recipe <- recipe(P_mgkg ~ Depth_cm + tmax8110cat + naniws + tmax8110ws + pwsl1pomu, data = P.Predictors) %>% 
  step_dummy(all_nominal_predictors())

#Combine model and recipe into a workflow
glm_workflow <- workflow() %>% 
  add_model(glm_spec) %>% 
  add_recipe(glm_recipe)

#Fit the workflow to the data
glm_fit <- fit(glm_workflow, data = P.Predictors)

# View the trained model coefficients
tidy(glm_fit)


#predict using test data

glm_preds <- predict(glm_fit, new_data = test.prep)
head(glm_preds)

Test_predictionsGLM<-cbind(test.prep, glm_preds) %>% dplyr::rename(Predicted_P=.pred)

names(Test_predictionsGLM)

#Write predictions to file
write.csv(Test_predictionsGLM %>% dplyr::select(P_mgkg, Predicted_P), paste0(dir.final.model, "/Model_predictions_test_data/ModelGLM_groupsplit_full_nosoil.csv"), row.names=FALSE)

#Calculate R-squared
rsq_test <- Test_predictionsGLM %>%
  rsq(truth = P_mgkg, estimate = Predicted_P)

print(rsq_test)

rmse <- with(Test_predictionsGLM, sqrt(mean((P_mgkg - Predicted_P)^2)))
print(rmse)



#Scatter plot: actual vs predicted
GLM.plot<-ggplot(Test_predictionsGLM, aes(x = P_mgkg, y = Predicted_P)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    title = paste0("Model Predictions (R² = ", round(rsq_test$.estimate, 2), ")")
  ) +
  xlab(bquote(Observed~soil~TP~(mg~kg^-1)))+
  ylab(bquote(Predicted~soil~TP~(mg~kg^-1)))+
  theme_minimal()+
  xlim(0,1500)+
  ylim(0,1500)
GLM.plot 

#Calculate all benchmarks
#Define metrics for comparison
metrics <- yardstick::metric_set(
  yardstick::rsq, 
  yardstick::rmse, 
  yardstick::mae, 
  yardstick::mape, 
  yardstick::mpe
)

#Evaluate metrics by model
evalLM <- Test_predictionsGLM %>%
  metrics(truth = P_mgkg, estimate = Predicted_P)

evalLM

```

## 7.6 LASSO Model

See: <https://www.publichealth.columbia.edu/research/population-health-methods/least-absolute-shrinkage-and-selection-operator-lasso>

Things to figure out:

1.  is there a group splitting function - yes
2.  is there a stratification option

```{r}
#install.packages("glmnet")
library(glmnet)


#==================================================
##Set up train and test data
#==================================================

#Note that glmnet automatically centers and scales predictors by default

#Make sure categorical features match across train and test datasets 
feature_cols <- setdiff(names(P.Predictors), c("P_mgkg", "comid", "siteID"))

for (col in feature_cols) {
  if (is.factor(P.Predictors[[col]]) || is.character(P.Predictors[[col]])) {
    # Convert both to factor with identical levels
    P.Predictors[[col]] <- factor(P.Predictors[[col]])
    test.prep[[col]] <- factor(test.prep[[col]], levels = levels(P.Predictors[[col]]))
  }
}

#Prepare train and test matrices
#exclude ID columns
x    <- model.matrix(P_mgkg ~ . - comid - siteID - X, data = P.Predictors)[, -1]
newx <- model.matrix(P_mgkg ~ . - comid - siteID - X, data = test.prep)[, -1]

#Check to make sure train and test columns align
setdiff(names(as.data.frame(x)), names(as.data.frame(newx)))

#Prepare target features
y <- P.Predictors$P_mgkg

#for log-transformed model (to address right skew)
#y<-log(P.Predictors$P_mgkg)

#==================================================
#Set Up Cross Validation
#==================================================

#Group Splitting:
#Group samples from same comid into same folds

# Extract the unique comids from the training data
unique_comids <- unique(P.Predictors$comid) 

# Randomly assign each unique comid to one of the 10 folds evenly
k_folds<-10
set.seed(123)
fold_assignments <- sample(rep(1:k_folds, length.out = length(unique_comids)))

fold_assignments

# Create a named vector to match fold IDs to comids
names(fold_assignments) <- as.character(unique_comids)

# Map the assigned folds back to the original dataset based on comid
# This ensures every row with the same comid gets the exact same fold ID
fold_ids <- fold_assignments[as.character(P.Predictors$comid)]

fold_ids



# Pass the grouped fold IDs into cv.glmnet using the foldid argument, conduct k-fold cross validation
set.seed(123)
cv_lasso <- cv.glmnet(x, y, alpha = 1, foldid = fold_ids)

# Plot cross-validation error vs log(lambda)
plot(cv_lasso)

# Extract optimal lambda
best_lambda <- cv_lasso$lambda.min
print(paste("Optimal Lambda:", best_lambda))

# ==========================================
#Final Model & Predictions
# ==========================================
set.seed(123)
final_model <- glmnet(x, y, alpha = 1, lambda = best_lambda)


# Generate predictions
pred_test<-predict(final_model, newx = newx)
# Append predictions to test data
Test_predictions <- test.prep %>% 
  mutate(Predicted_P = as.vector(pred_test))

#For Log Model - 
pred_test_log <- predict(final_model, newx = newx)
# Back-transform the predictions to the original scale for log model
pred_test_original_scale <- exp(pred_test_log)
# Append predictions to test data (log transformed)
Test_predictionsLOG <- test.prep %>% 
  mutate(Predicted_P = as.vector(pred_test_original_scale))


# Save predictions
write.csv(
  Test_predictions %>% dplyr::select(P_mgkg, Predicted_P), 
  paste0(dir.final.model, "/Model_predictions_test_data/UMRB_ModelLASSO_groupsplit_full.csv"), 
  row.names = FALSE
)

# ==========================================
#Evaluate Performance
# ==========================================
rsq_test <- Test_predictionsLOG %>%
  rsq(truth = P_mgkg, estimate = Predicted_P)

rmse_test <- with(Test_predictionsLOG, sqrt(mean((P_mgkg - Predicted_P)^2)))

print(rsq_test)
print(paste("RMSE:", rmse_test))


#Scatter plot: actual vs predicted
test.plot<-ggplot(Test_predictions, aes(x = P_mgkg, y = Predicted_P)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    title = paste0("LASSO Predictions (R² = ", round(rsq_test$.estimate, 2), ")")
  ) +
  xlab(bquote(Observed~soil~TP~(mg~kg^-1)))+
  ylab(bquote(Predicted~soil~TP~(mg~kg^-1)))+
  theme_minimal()+
  xlim(0, 1500)+
  ylim(0,1500)
test.plot 

#============================================
#Variable Importance
#============================================

#Look at coefficients for predictors
lasso_coefs <- coef(final_model, s = "lambda.1se")

important_vars <- which(lasso_coefs != 1) # drops zero coefficients
# Or view them explicitly as a data frame
coef_matrix <- as.matrix(lasso_coefs)

#Format predictors and standardized coefficient:
imp.var<-as.data.frame(subset(coef_matrix, coef_matrix[, 1] != 0))
head(imp.var)

# Appends the new column to the END of your data frame
imp.var$Predictors <- rownames(imp.var)

# Optional: Clear the old row names if desired
rownames(imp.var) <- NULL 

colnames(imp.var)[1]<-"standardized_coefficient"

imp.var<-imp.var %>% dplyr::select(Predictors, standardized_coefficient) %>% arrange(standardized_coefficient) %>% mutate(abs_sc=abs(standardized_coefficient))

View(imp.var)

ggplot(imp.var)+
  geom_col(aes(reorder(Predictors, standardized_coefficient), standardized_coefficient))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# 8 Model Performance: Independent validation data

## 8.0 Option to format test data for P stratification

```{r}
test_catchment_lookup <- test.prep %>%
  group_by(comid) %>%
  # Use na.rm = TRUE to completely ignore missing data during the calculation
  summarize(catchment_p_med = median(P_mgkg, na.rm = TRUE), .groups = "drop") %>%
  # Create the categorical bins explicitly on clean, summarized data
  mutate(comid_strata_bin = as.factor(ntile(catchment_p_med, 3))) %>%
  # Keep only the ID and the clean stratum bin column
  select(comid, comid_strata_bin)

# 2. Join the clean, non-variable bin column back into your main training data
test_data_prepared <- test.prep %>%
  inner_join(test_catchment_lookup, by = "comid")

#View(test_data_prepared)
```

## 8.1 Fit model to test data and evaluate accuracy

```{r}


#Predict on the independent test set
#create P strate column for test data

pred_test <- predict(final_fit, new_data = test.prep)
pred_test 
                    
Test_predictions<-cbind(test.prep, pred_test) %>% dplyr::rename(Predicted_P=`.pred`)

names(Test_predictions)

#Write predictions to file
write.csv(Test_predictions %>% dplyr::select(P_mgkg, Predicted_P), paste0(dir.final.model, "/Model_predictions_test_data/STREAMBANK_TWEEDIE_groupsplit_full_nostrata_RERUN092126.csv"), row.names=FALSE)


#model.name<-"/ModelBASICBOOST_groupsplit_nosoil_full_nostrata.csv"

#Read in predicted ata if necessary:
Test_predictions<-read.csv(paste0(dir.final.model, "/Model_predictions_test_data/STREAMBANK_MidwestRFfullgroupsplit_less1000.csv"))


#Read in test predictions if desired:
#Test_predictions<-read.csv(paste0(dir.final.model,"/Model_predictions_test_data", model.name))

#Calculate R-squared and RMSE
rsq_test <- Test_predictions %>%
  rsq(truth = P_mgkg, estimate = Predicted_P)

print(rsq_test)

rmse <- with(Test_predictions, sqrt(mean((P_mgkg - Predicted_P)^2)))
print(rmse)

#Scatter plot: actual vs predicted
test.plot<-ggplot(Test_predictions, aes(x = P_mgkg, y = Predicted_P)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=TRUE)+
  labs(
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test$.estimate, 2), ")")
  ) +
  xlab(bquote(Observed~TP~(mg~kg^-1)))+
  ylab(bquote(Predicted~TP~(mg~kg^-1)))+
  theme_minimal()+
 xlim(0, 1000)+
  ylim(0,1000)
test.plot 

#Save to plot to file
png(paste0(file.name=dir.figures, "/Fig_STREAMBANK_MidwestRFfullgroupsplit_less1000.png"), width = 4, height = 4, units="in", res = 300)    

grid.newpage() 
print(test.plot) 

dev.off() 


#Check for samples with P<1000
rsq_test.less1000 <- Test_predictions %>% filter(P_mgkg<=750) %>%
  rsq(truth = P_mgkg, estimate = Predicted_P)
rsq_test.less1000

rmse.less1000<-Test_predictions %>% 
    filter(P_mgkg<=750) %>% 
    mutate(rmse=sqrt(mean((P_mgkg - Predicted_P)^2))) %>% 
    dplyr::select(rmse) %>% distinct()
rmse.less1000

metrics <- yardstick::metric_set(
  yardstick::rsq, 
  yardstick::rmse, 
  yardstick::mae, 
  yardstick::mape, 
  yardstick::mpe
)

#Evaluate metrics for samples with very high TP excluded
lowerTP<-Test_predictions %>% filter(P_mgkg<=750)


lowerTP_results <- lowerTP %>%
  metrics(truth = P_mgkg, estimate = Predicted_P)

print(lowerTP_results)




#Plot Test data vs Predicted for samples <1000 mg/kg TP
ggplot(Test_predictions %>% filter(P_mgkg<1000), aes(x = P_mgkg, y = Predicted_P)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    x = "Actual P_mgkg",
    y = "Predicted P_mgkg",
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test.less1000$.estimate, 3), ")")
  ) +
  theme_minimal()+
  xlim(0,1000)+
  ylim(0,1000)


```

Evaluate predictions for log-transformed outcome

```{r}

names(test.prep)

#Predict on the independent test set
pred_test <- predict(final_fit, new_data = test.prep)
pred_test 

# 2. Combine and back-transform predictions to actual units (mg/kg)
Test_predictions <- test.prep %>%
  bind_cols(pred_test) %>%
  mutate(
    # Back-transform from log(y + 1) to original units: exp(z) - 1
    Predicted_P = exp(.pred) - 1 
  )

# 3. View actual vs predicted values
head(Test_predictions %>% select(P_mgkg, Predicted_P))

#Calculate R-squared and RMSE
rsq_test <- Test_predictions %>%
  rsq(truth = P_mgkg, estimate = Predicted_P)

print(rsq_test)

rmse <- with(Test_predictions, sqrt(mean((P_mgkg - Predicted_P)^2)))
print(rmse)


#Write predictions to file
write.csv(Test_predictions %>% dplyr::select(P_mgkg, Predicted_P), paste0(dir.final.model, "/Model_predictions_test_data/ModelBOOSTLOG_SOIL_FULL.csv"), row.names=FALSE)
```

## 8.2 Model Performance Benchmarks

Use yardstick package to compare metrics for multiple predictions: R2, RMSE, MAE, MAPE

TO ADD (per reviewer):

- more exploration of bias

- residuals inspection: Further residuals inspection (not just obs. vs. pred) would be helpful and possibly clarify oddities in the data; e.g. residuals against predicted TP, against the ‘most important’ predictors, a QQ-plot.  

Background on metrics:

Mean Squared Error (MSE): average of the squared difference between original and predicted values. Identifies outliers and imbalances. Most suitable when data has normal distribution.

Root Mean Squared Error (RMSE): square root of MSE. measures the standard deviation of the error. Affected by outliers, but to a lesser extent. Essential to ID outlieres. Measured in same units as predicted variable.

Mean Average Error (MAE): average absolute differences between the expected and actual values. Not as sensitive to outliers.

Mean Absolute Percentage Error (MAPE): absolute percentage differences between the models' predictions and actual values. Expresses average error as a percentage of actual value. Penalizes negative errors more heavily (when the predicted value exceeds the actual). Tends to favor models that underestimate rather than overestimate.

Mean Percentage Error (MPE): average percentage bias on a predictive model.

```{r}

library(readr)
library(yardstick)

#read in and concanate all model predictions

#location where predictions are located:
folder_path <- paste0(dir.final.model, "/Model_predictions_test_data")
folder_path

#list all prediction files
file_list <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)
file_list

#read in all predictions together
predictions<-read_csv(file_list, id = "filename") 
#clean up model name
predictions<-predictions %>% 
  mutate(filename=basename(filename)) %>% 
  mutate(filename = str_remove(filename, "\\.csv$")) |> 
  dplyr::rename(Model=filename)

head(predictions)
nrow(predictions)

#temporary workaround for now  - omit negative predictions (LASSO)

predictions<-predictions %>% filter(Predicted_P>0)


#Define metrics for comparison
metrics <- yardstick::metric_set(
  yardstick::rsq, 
  yardstick::rmse, 
  yardstick::mae, 
  yardstick::mape, 
  yardstick::mpe
)

#Evaluate metrics by model
evaluation_results <- predictions %>%
  group_by(Model) %>%
  metrics(truth = P_mgkg, estimate = Predicted_P)

print(evaluation_results)

View(evaluation_results)

#View wide for easier comparison

eval_wide<-evaluation_results |> 
  pivot_wider(names_from='.metric', values_from='.estimate')

View(eval_wide)

##Add model performance for values <1000 mg/kg, since model tends to perform poorly for high values

# 2. Evaluate on predictions under threshold (< 1000)
eval_sub1000 <- predictions %>%
  filter(P_mgkg < 1000) %>%  # Filter for threshold on true values (or Predicted_P < 1000 depending on needs)
  group_by(Model) %>%
  metrics(truth = P_mgkg, estimate = Predicted_P) %>%
  pivot_wider(
    names_from = .metric, 
    values_from = .estimate,
    names_glue = "{.metric}_under1000"  # Appends "_under1000" to new column names
  )

# 3. Combine both into a single summary table
eval_wide_full <- left_join(eval_wide, eval_sub1000, by = "Model")

View(eval_wide_full)

#write temporary table to file

write_xlsx(eval_wide, paste0(dir.final.model, "/model_comparison_table/DRAFT_model_comparison_table.xlsx"))

#Match to model specs
model.specs<-read_xlsx(paste0(dir.final.model, "/Revision_model_runs_groupsplit.xlsx"), sheet="group_split_models")

#Rename columns to merge
model.specs<-model.specs |> 
  dplyr::rename(Model_abrev=Model) |> 
  dplyr::rename(Model=Predictions_file)

names(model.specs)

View(model.specs)

#Add specs to model performance metrics

model_table<-merge(model.specs |> dplyr::select(-c("Test file", "Model_name", "Test_R2", "Test_RMSE", "Top_5_SHAP", "Notes")), eval_wide, by=c("Model"))

View(model_table)

#Write results to table
write_xlsx(model_table, paste0(dir.figures, "/Model_comparison_stats.xlsx"))
```

## 8.3 Older Benchmarks (first draft of manuscript)

Evaluating "acceptability" of model performance, based on Moriasi et al 2007

<https://swat.tamu.edu/media/1312/moriasimodeleval.pdf>

RSR - Based on the recommendation by Singh et al. (2004), a model evaluation statistic, named the RMSE-observations standard deviation ratio (RSR), was developed. RSR standardizes RMSE using the observations standard deviation, and it combines both an error index and the additional information recommended by Legates and McCabe (1999). RSR is calculated as the ratio of the RMSE and standard deviation of measured data

NSE ranges between −∞ and 1.0 (1 inclusive), with NSE = 1 being the optimal value. Values between 0.0 and 1.0 are generally viewed as acceptable levels of performance, whereas values \< 0.0 indicates that the mean observed value is a better predictor than the simulated value, which indicates unacceptable performance.

In general, model simulation can be judged as satisfactory if NSE \> 0.50 and RSR \< 0.70, and if PBIAS 25% for streamflow, PBIAS 55% for sediment, and PBIAS 70% for N and P

```{r}
#RMSE as a % of sample average
mean(P.Predictors$P_mgkg)
rmse/mean(P.Predictors$P_mgkg)

#RSR - RMSE-observations standard deviation ratio (RSR)
#< 0.7 is satisfactory
rmse/sd(P.Predictors$P_mgkg)

#Nash Sutcliffe efficiency (NSE)
#should be >0
Test_predictions %>% 
  mutate(Diff=(P_mgkg-.pred)^2) %>% 
  mutate(Diff.mean=(P_mgkg-mean(P_mgkg))^2) %>% 
  summarise(sum.diff=sum(Diff), sum.diff.mean=sum(Diff.mean)) %>% mutate(NSE=1-(sum.diff/sum.diff.mean))

```

# 9 Run Model on Additional Datasets

Steps:

- Match predictor variables of new dataset to model dataset

  - Values for predictors should be within range of model dataset

- Impute missing values for new dataset, using model dataset to help inform

- Run new dataset through the model

- Plot linear relationship

## 9.1 Load additional data you'd like to test or predict values for

**a. Predict values for another dataset**

```{r}

#1. Name additional dataset(s) you'd like to test:
#Note: to compare performance of larger Midwest model in predicting streambank P to streambank P specific models, they all must be evaluated on the same test dataset (ie, the same portion of test data used to build streambank models), not the full streambank P dataset

new.test.input<-"/Margenot_sample_attributes" #Margenot IL data

#2. Load data:

new.test.data<-read.csv(paste0(dir.final.model.data, new.test.input, '_for_modelling_wSSURGO_LIMITED_TEST_imputed_nosoil_08242026.csv'))
head(new.test.data)
nrow(new.test.data)

sort(names(new.test.data))

```

### b. Predict values at unknown locations

```{r}
#for predicting values at unknown locations

grid<-fread(paste0(dir.output, "/prediction_grid_points_SSURGO_attributes_mapunit_LIMITED.csv"))

#assign siteID as row number

grid$siteID <- paste0("grid_", 1:nrow(grid))

#assign empty P_mgkg column?

#Assign depth values
#Could assign 1st quartile, median, and 3rd quartile depth values from Streambank dataset?? And keep all other columns the same for a given location? (this will triple the size of the prediction dataset)

grid1<-grid |> mutate(Depth_cm=104)
head(grid1)

grid2<-grid |> mutate(Depth_cm=175)

grid3<-grid |> mutate(Depth_cm=260)

head(grid3)

grid.all<-bind_rows(grid1, grid2, grid3)
nrow(grid.all) #check number of rows 
nrow(grid)

new.test.data<-grid.all


new.test.data |> filter(siteID=="grid_1"&Depth_cm=="175")

names(new.test.data)

#fix minor format issues for compatibility with modelling datasets
new.test.data<-new.test.data |> dplyr::rename(NLCD_Land=nlcd_land, WETLAND_TYPE=wetland_type)

```

## 9.2 Load model you wish to use to predict values for new dataset

```{r}

#set directory for final saved models
dir.final.model<-paste0(dir.output, "/Final_model_comparisons")

#specify which model
model.input<-"/USGS_NCSS_soils_database_attributes_alldepths_for_modelling_wSSURGO_LIMITED_comid"

#read in model

final_fit<-readRDS(paste0(dir.output, "/Final_model_comparisons", model.input, "_RF_model_LIMITED_stepdummy_group_split_less1000_082726.rds"))

```

## 9.3 Make sure predictors in new dataset match predictors in the model

```{r}
#3. Make sure predictor columns of new data match predictor columns of data used to build the model

#set directory for final train/test datasets
dir.final.model.data<-paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets")



#Read in desired predictor dataset

P.Predictors<-read.csv(paste0(dir.final.model.data, model.input, "_train_imputed_less1000_groupsplit_08252026.csv"))


names(P.Predictors)

new.test.data2<-new.test.data %>%
  { if ("meanP_mgkg" %in% names(.)) rename(., P_mgkg = meanP_mgkg) else . } %>% #rename P column if took average of subsamples
  dplyr::select(any_of(names(P.Predictors))) #grab all columns that match P.Predictors 

names(new.test.data2)

#Do all columns match?
all(sort(names(new.test.data2)) == sort(names(P.Predictors))) # should be TRUE

#just in case, find missing columns:
setdiff(names(new.test.data2), names(P.Predictors))

setdiff(names(P.Predictors), names(new.test.data2))


names(new.test.data2)

#Match categorical levels where needed
levels(factor(new.test.data2$NLCD_Land))
levels(factor(P.Predictors$NLCD_Land))

#For Margenot Streambank samples, get NLCD levels to match soil TP dataset:
new.test.data2$NLCD_Land<-factor(new.test.data2$NLCD_Land,
                            labels=c("Barren Land", "Cultivated Crops", "Deciduous Forest", "Developed, High Intensity", "Developed, Low Intensity", "Developed, Medium Intensity", "Developed, Open Space", "Emergent Herbaceous Wetlands", "Evergreen Forest", "Pasture/Hay", "Herbaceous", "Mixed Forest", "Open Water", "Shrub/Scrub", "Woody Wetlands"))


levels(factor(new.test.data2$WETLAND_TYPE))
levels(factor(P.Predictors$WETLAND_TYPE))

#Set non wetland values to match soil TP dataset:

new.test.data2$WETLAND_TYPE<-factor(new.test.data2$WETLAND_TYPE, levels=c("", 
"Freshwater Emergent Wetland",
"Freshwater Forested/Shrub Wetland", 
"Freshwater Pond",
"Lake", 
"Other",
"Riverine"),  labels=c("Non wetland", 
                       "Freshwater Emergent Wetland",
                       "Freshwater Forested/Shrub Wetland",
                       "Freshwater Pond",
                       "Lake", 
                       "Other",
                       "Riverine"))
                                                                
 

#SET ATTRIBUTE TYPES
#Set categorical attributes to factor
#SET ATTRIBUTE TYPES
#Set categorical attributes to factor
#and set ID attributes to factor
cat_vars <- names(P.Predictors)[sapply(P.Predictors, function(x) is.factor(x) | is.character(x))]
cat_vars

P.Predictors<-P.Predictors %>% 
  dplyr::mutate(across(cat_vars, as.factor)) |> 
  dplyr::mutate(across(c(siteID, comid), as.factor))
new.test.data2<-new.test.data2 %>% 
  dplyr::mutate(across(cat_vars, as.factor)) |> 
  dplyr::mutate(across(c(siteID, comid), as.factor))


#Set remaining numeric and integer attributes to numeric
P.Predictors<-P.Predictors |> 
  dplyr::mutate(across(where(is.numeric), as.numeric))

new.test.data2<-new.test.data2 |> 
  dplyr::mutate(across(where(is.numeric), as.numeric))


#Check to make sure columns compare
library(janitor)
compare_df_cols(P.Predictors, new.test.data2, return = "mismatch")
```

## 9.4 Restrict data to samples with TP \< 1000 mg/kg (Optional)

```{r}

new.test.data2<-new.test.data2 %>% 
  filter(P_mgkg<1000)
```

## 9.5 Predict modeled values for the new dataset

```{r}
# 5. Predict on the independent test set
start.time<-Sys.time()
NEW.pred_test <- predict(final_fit, new_data = new.test.data2)
NEW.pred_test 

end.time<-Sys.time()

total.time<-end.time - start.time
total.time

NEW.Test_predictions<-cbind(NEW.DATA, NEW.pred_test)%>% rename(Predicted_P=`.pred`)
names(NEW.Test_predictions)


#Write predictions to file
#for additional dataset
write.csv(NEW.Test_predictions %>% dplyr::select(P_mgkg, Predicted_P), paste0(dir.final.model, "/Model_predictions_test_data/STREAMBANK_MidwestRFfullgroupsplit_less1000.csv"), row.names=FALSE)

View(NEW.Test_predictions %>% dplyr::select(P_mgkg, Predicted_P))

#for unknonw sample locations:
NEW.pred_test<-NEW.pred_test |> rename(TP=`.pred`)

#need to add x and y back in

summary(NEW.pred_test$TP)
names(new.test.data)
NEW.Test_predictions<-cbind(new.test.data |> dplyr::select(siteID, Depth_cm, x, y), NEW.pred_test)

#write to file
fwrite(NEW.Test_predictions, paste0(dir.final.model, "/Model_predictions_test_data/predicted_TP_streambank_30m_grid.csv"))

head(NEW.Test_predictions)
```

## 9.6 Evaluate prediction performance

```{r}
# 6. Calculate R-squared
rsq_test <- NEW.Test_predictions %>%
  rsq(truth = P_mgkg, estimate = Predicted_P)

print(rsq_test)

rmse <- with(NEW.Test_predictions, sqrt(mean((P_mgkg - Predicted_P)^2)))
print(rmse)

# 7. Scatter plot: actual vs predicted
streambank.test.plot<-ggplot(NEW.Test_predictions, aes(x = P_mgkg, y = Predicted_P)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test$.estimate, 2), ")")
  ) +
  xlab(bquote(Observed~soil~TP~(mg~kg^-1)))+
  ylab(bquote(Predicted~soil~TP~(mg~kg^-1)))+
  theme_minimal()+
  xlim(0,1500)+
  ylim(0,1500)
streambank.test.plot

#Check for samples with P<1000
rsq_test.less1000 <- NEW.Test_predictions %>% filter(P_mgkg<1000) %>%
  rsq(truth = P_mgkg, estimate = .pred)
rsq_test.less1000


ggplot(NEW.Test_predictions %>% filter(P_mgkg<1000), aes(x = P_mgkg, y = .pred)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    x = "Actual streambank TP (mgkg)",
    y = "Predicted streambank TP (mg/kg)",
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test.less1000$.estimate, 2), ")")
  ) +
  theme_minimal()


#Save to plot to file
png(paste0(file.name=dir.figures, "/Fig_pred_vs_obs_soilTP_xgboost_reduced_with_WS_included_less1000_reducedcor_021126_STREAMBANK_PREDICTED.png"), width = 4, height = 4, units="in", res = 300)    

grid.newpage() 
print(streambank.test.plot) 

dev.off()
```

## 9.7 Compare to baseline performance for streambank data

Compare streambank test dataset to mean of streambank train dataset

```{r}

streambank.train<-read.csv(paste0(dir.final.model.data, new.test.input, '_for_modelling_wSSURGO_LIMITED_train_imputed_nosoil_08242026.csv'))
                           
 
Model_MEAN_STREAMBANK<-new.test.data2 %>% dplyr::select(P_mgkg) %>% 
  mutate(Predicted_P=mean(streambank.train$P_mgkg))

head(Model_MEAN_STREAMBANK)

write.csv(Model_MEAN_STREAMBANK %>% dplyr::select(P_mgkg, Predicted_P), paste0(dir.final.model, "/Model_predictions_test_data/STREAMBANKMEAN_groupsplit.csv"), row.names=FALSE)

rmse <- with(Model_MEAN_STREAMBANK, sqrt(mean((P_mgkg - Predicted_P)^2)))
print(rmse)

```
