---
title: "module_tune_models"
author: "Christy Dolph"
date: "2025-08-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

# 1 Install and load the required libraries

```{r}

knitr::opts_chunk$set(echo = TRUE)

# Clean up workspace 
rm(list=ls())


#list of necessary packages 
want <- c('tidyverse', 'data.table', 'xgboost', 'caret', 'dplyr', 'ggplot2', 'scales', 'tidymodels', 'missRanger', 'doParallel', 'DALEXtra', 'vip', 'grid', 'stringr')

need <- want[!(want %in% installed.packages()[,"Package"])]
if (length(need)) install.packages(need)
lapply(want, function(i) require(i, character.only=TRUE))
rm(want, need)

#Main directory
dir.root<-getwd()

#Input data
dir.data<-paste(dir.root, "/Streambank_P_measured_data", sep="")

#Spatial data
dir.spatial<-paste(dir.root, "/Spatial_data", sep="")

#Output data (produced by script)
dir.output<-paste0(dir.root, "/output_data")

dir.figures<-paste0(dir.root, '/Maps_figures')

#set directories for final saved models + data

dir.final.model.data<-paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets")

dir.final.model<-paste0(dir.output, "/Final_model_comparisons")

```

# 2 Load the dataset

```{r}

#Select the input data you'd like to use: 

#Margenot et al IL streambank P data, with limited ssurgo attributes to retain all data, for independent streambank model or for running through existing soil TP model (predictors are matched below): 
#model.input<-"/Margenot_sample_attributes_for_modelling_wSSURGO_LIMITED"


#Modern upland Soil P database:
#model.input<-'/USGS_NCSS_soils_database_attributes_alldepths_for_modelling'

#Modern upland soil P database but with limited ssurgo attributes that should encompass water areas?
model.input<-'/USGS_NCSS_soils_database_attributes_alldepths_for_modelling_wSSURGO_LIMITED'

P_dataset <-read.csv(paste0(dir.output, model.input, '.csv'))
                    
names(P_dataset) 
nrow(P_dataset)
#View(P_dataset)

#Check class of all attributes
sapply(P_dataset, class)


#Convert characters to factors:
P_dataset <- P_dataset %>% 
  type.convert(as.is=FALSE)

#Convert integer attributes to numeric:
integer_cols <- sapply(P_dataset, is.integer)
P_dataset[integer_cols] <- lapply(P_dataset[integer_cols], as.numeric)

# Check the class of columns
sapply(P_dataset, class)

#Make name of P_mgkg column consistent
#NOTE: need to fix this later! (not working)
P_dataset<-P_dataset %>%
  { if ("meanP_mgkg" %in% names(.)) rename(., P_mgkg = meanP_mgkg) else . }
```

## 2.1 Option to add pH and other site-specific data for streambank samples

```{r}

ph<-readxl::read_excel(paste0(dir.data, "/Margenot streambank soil data to Christy Dolph_01202026_wpH.xlsx"))
nrow(ph)
#match pH to existing attributes based on lat long

names(P_dataset)
names(ph)

#Clean up columns
ph.clean<-ph %>% 
  separate_wider_delim(cols = 'Depth (cm)', delim = "-", names = c("Depth_min_cm", "Depth_max_cm"),cols_remove = FALSE, too_few="debug") %>%
  dplyr::rename(x=Long, y=Lat, pct.sand=`%Sand`, pct.silt=`%Silt`, pct.clay=`%Clay`, Depth_cm=Depth_max_cm) %>% 
   dplyr::select(x,y,pH,pct.silt, pct.sand, pct.clay, Depth_cm) %>% 
  na.omit() #drop samples with missing data
names(ph.clean)

#View(ph.clean)
nrow(ph.clean)

#View(P_dataset %>% dplyr::select(x,y,P_mgkg, Depth_cm))

class(ph$x)
class(P_dataset$x)

P_dataset.ph<-merge(P_dataset, ph.clean, by=c('x', 'y', 'Depth_cm'))

nrow(P_dataset.ph) #Note 34 sites were dropped after merge 

#To use pH data:
P_dataset<-P_dataset.ph
```

# 3 Explore the data

## 3.1 Distribution of streambank P values

```{r}

nrow(P_dataset)

#Histogram of P
P.hist<-
  ggplot(P_dataset, aes(P_mgkg))+
  #geom_histogram(binwidth = 1,colour = "lightgray", fill = "darkgrey") +
  geom_histogram(binwidth=10, colour = "black", fill ="black") +
  #geom_vline(aes(xintercept=mean(P_mgkg)), linetype="dashed", color="blue")+
  geom_vline(aes(xintercept=mean(P_mgkg)), linetype="dashed", color="purple", linewidth=1.5)+
  #geom_vline(aes(xintercept=1000), linetype="dotdash", color="black")+
  #geom_vline(aes(xintercept=1000), linetype="dotdash", #color="darkgray", linewidth=1.5)+
  xlab("P (mg/kg)") +
  ylab ("Frequency") +
  #ggtitle("Near Channel Soil P distribution")+
  #ggtitle("Soil P Samples distribution") +
  theme(axis.title = element_text(size=20))+
  theme(axis.text=element_text(size=20))+
  #theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  #theme(plot.title = element_text(size = 26, face = "bold", hjust=0.5))
P.hist

summary(P_dataset$P_mgkg)

#Percent of values over 1000 mg/kg
nrow(P_dataset)

P_dataset %>% filter(P_mgkg>1000) %>% nrow()


```

Mean P value

```{r}

#Including and excluding samples with P >1000 mg/kg

P_dataset %>% 
  filter(P_mgkg<1000) %>% 
  summarise(meanP=mean(P_mgkg))

```

Plot streambank P vs bulk density

```{r}

ggplot(P_dataset) +
  geom_point(aes(meanBD_gcm3, P_mgkg))

```

## 3.2 Check class of all variables

```{r}
summary(P_dataset)
class(P_dataset)

```

Create correct/matching levels for NLCD attribute

```{r}
#Set NLCD_Land to categorical

levels(factor(P_dataset$NLCD_Land))

#For Margenot Streambank samples, get levels to match soil TP:
P_dataset$NLCD_Land<-factor(P_dataset$NLCD_Land,
                       labels=c("Barren Land", "Cultivated Crops", "Deciduous Forest", "Developed, Low Intensity", "Developed, Open Space", "Emergent Herbaceous Wetlands", "Pasture/Hay", "Mixed Forest", "Open Water", "Woody Wetlands"))

#For big soil TP dataset, set factor levels:
P_dataset$NLCD_Land<-factor(P_dataset$NLCD_Land, 
                            levels=c('11','21','22','23','24','31','41','42','43','52', '71','81','82','90','95'),
                            labels=c('Open Water', 'Developed, Open Space', 'Developed, Low Intensity','Developed, Medium Intensity', 'Developed, High Intensity', 'Barren Land','Deciduous Forest','Evergreen Forest','Mixed Forest','Shrub/Scrub','Grassland/Herbaceous', 'Pasture/Hay','Cultivated Crops','Woody Wetlands','Emergent Herbaious Wetlands'))

levels(P_dataset$NLCD_Land)
```

Assign correct variable type to remaining variables:

```{r}
#Set remaining categorical attributes to factor
cat_vars <- names(P_dataset)[sapply(P_dataset, function(x) is.factor(x) | is.character(x))]
cat_vars

P_dataset<-P_dataset %>% 
  mutate(across(cat_vars, as.factor))


#Set remaining numeric attributes to numeric
dep_var="P_mgkg"
num_vars <- names(P_dataset)[sapply(P_dataset, is.numeric) & names(P_dataset) != dep_var]
num_vars

P_dataset <- P_dataset %>%
  mutate(across(num_vars, as.numeric))

#Check all columns
sapply(P_dataset, class)
```

## 3.3 Relationship to other variables

### a. Categorical attributes

```{r}

#For categorical attributes, plot boxplots:  

#full dataset:
df<-P_dataset

dep_var="P_mgkg"
cat_vars <- names(df)[sapply(df, function(x) is.factor(x) | is.character(x))]


# Open a PDF device
pdf(paste0(dir.output, model.input, "_boxplots_by_category.pdf"))

# Loop through categorical variables and plot to PDF (one per page)
walk(cat_vars, function(cat_var) {
  p <- ggplot(df, aes_string(x = cat_var, y = dep_var)) +
    geom_boxplot(fill = "skyblue", color = "black") +
    labs(title = paste("Boxplot of", dep_var, "by", cat_var)) +
    theme_minimal()+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 
  
  print(p)  # important to print inside loop when saving
})

# Close PDF device
dev.off()

```

### b. Numeric attributes

```{r}

#full dataset:
df<-P_dataset

num_vars <- names(df)[sapply(df, is.numeric) & names(df) != dep_var]

# Open a PDF device (all plots go here, one per page)
pdf(paste0(dir.output, model.input, "_scatterplots_", dep_var, ".pdf"))

# Loop through numeric variables and plot
walk(num_vars, function(num_var) {
  
  # Fit linear model
  formula <- as.formula(paste(dep_var, "~", num_var))
  model <- lm(formula, data = df)
  r2 <- summary(model)$r.squared
  
  # Create scatterplot with R² annotation
  p <- ggplot(df, aes_string(x = num_var, y = dep_var)) +
    geom_point(color = "steelblue", alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = paste("Scatterplot of", dep_var, "vs", num_var)) +
    annotate("text", 
             x = Inf, y = Inf, 
             label = paste0("R² = ", round(r2, 3)), 
             hjust = 1.1, vjust = 1.5, size = 5, color = "black") +
    theme_minimal()
  
  print(p)  # each plot on a new page
})

# Close PDF device
dev.off()
```

### c. Manuscript figure (Appendix): Examine relationships among parent material: glacial till vs loess

"pcteolfinews" = % of AOI area classified as lithology type: eolian sediment, fine-textured (glacial loess)

```{r}
P_dataset %>% order() %>% names()

P_dataset %>% 
 select(order(colnames(P_dataset))) %>% names()

#Parent material:

parent<-c("pctglactilclayws",
"pctglaclakefinews",
"pctglaclakecrsws",
"pcteolcrsws",
"pcteolfinews",
"pctglactilloamws",
"pctglactilcrsws",
"pctalluvcoastws",
"pctcolluvsedws",
"clayws",
"pctglactilcrscat",
"pctglactilclaycat")

soils<-P_dataset %>% dplyr::select(all_of(parent))

ggplot(soils)+
  geom_point(aes(pctglactilloamws, pcteolfinews))

ggplot(soils)+
  geom_point(aes(pctglactilclayws, pctglactilloamws))

reg<-with(soils, lm(pctglactilloamws ~ pctglactilclayws))
summary(reg)
```

# 4 Data Preparation

### 4.1 Remove any remaining administrative attributes:

```{r}
P_dataset<-P_dataset %>% dplyr::select(-musumcpct, -musumcpcta, -musumcpcts)

```

### 4.2 Option to remove x and y as predictors

```{r}
P_dataset<-P_dataset %>% dplyr::select(-x, -y)
```

### 4.3 Option to restrict to samples with TP \<1000 mg/kg

```{r}
P_dataset<-P_dataset %>% filter(P_mgkg<1000)
```

### 4.4 Option to remove outlier(s)

Note we did not remove outliers for the streambank model.

```{r}
# Calculate the 0.5th and 99.5th percentiles 
data_vector<-P_dataset$P_mgkg 

lower_bound <- quantile(data_vector, 0.005) 
lower_bound  

upper_bound <- quantile(data_vector, 0.995) 
upper_bound  

# Remove outliers 
data_without_outliers <- P_dataset %>%    
  filter(P_mgkg>lower_bound & P_mgkg<upper_bound) 
nrow(data_without_outliers)

#Option to set data as full dataset, or without outliers
#without outliers:
P_dataset<-data_without_outliers

summary(P_dataset$P_mgkg)
```

### 4.5 Omit attributes with mostly NA values or with no information (standard deviation of 0)

Machine learning models can't handle NA, so need to either drop attributes that have NA or impute missing values (below)

BUG: this may not be dropping columns that have blank rather than NA??

```{r}


#Identify columns containing NA values:
cols_with_na <- P_dataset %>%
  dplyr::select_if(~any(is.na(.))) %>%
  colnames()

# Print the names of columns with NA values
print(cols_with_na)

#Identify columns that are >50% NA values

cols_to_remove<-P_dataset %>%
  dplyr::select_if(~ mean(is.na(.)) > 0.5) %>% 
  names()

print(cols_to_remove)

#Remove columns that are >50% NA

P_dataset<-P_dataset %>% 
  dplyr::select(-all_of(cols_to_remove))

names(P_dataset)

#ID columns with no information (numeric)
cols_no_info<-P_dataset %>% 
   dplyr::select(where(is.numeric)) %>% 
dplyr::select(where(~ sd(.x, na.rm = TRUE) == 0)) %>% names()

print(cols_no_info)

#Omit predictors with no information (standard deviation of 0)
P_dataset <-P_dataset %>% 
  dplyr::select(-all_of(cols_no_info))

```

### 4.6 Drop or reclassify factor levels with only one representative (one data point):

ID factors that have a level with only one data point. If these factors are included ML models typically won't run.

```{r}
cols_with_single_point_levels <- P_dataset %>%
  dplyr::select(where(is.factor)) %>% # Select only factor columns
  sapply(function(x) any(table(x) == 1)) %>% # Check if any level has count 1
  names(.)[.] # Get the names of columns where the condition is TRUE

cols_with_single_point_levels
```

Reclassify or drop factors:

```{r}


#Drop remaining attributes
cols_with_single_point_levels <- P_dataset %>%
  dplyr::select(where(is.factor)) %>% # Select only factor columns
  sapply(function(x) any(table(x) == 1)) %>% # Check if any level has count 1
  names(.)[.] # Get the names of columns where the condition is TRUE

cols_with_single_point_levels

P_dataset<-P_dataset %>% 
  dplyr::select(-any_of(cols_with_single_point_levels))

names(P_dataset)
```

### 4.7 Identify & remove highly correlated predictors

While xgboost and other ML models are typically robust to multi-collinearity, reducing the number of features in the model by eliminating redundant/correlated variables can help with 1) model training time, 2) interpretation of the model in terms of variable importance and 3) may offer some performance gains.

Note: you can only do this for numeric variables where data isn't missing! So will need to exclude NA rows and categorical variables

Use the findCorrelation() function from the caret package to remove correlated variables while keeping the one of a correlated pair:

-   From the function documentation: "If two variables have a high correlation, the function looks at the mean absolute correlation of each variable and removes the variable with the largest mean absolute correlation." In other words, it considers both negative and positive correlations by using the absolute value

-   the function chooses one of the two variables from a pair based on how correlated it is with all the other variables

-   Note: this correlation matrix includes the dependent variable, so need to think about if you want to include that

```{r}
# Calculate the correlation matrix

set.seed(123)
cor_matrix <- P_dataset %>% 
  dplyr::select(where(is.numeric)) %>% 
  na.omit() %>% 
  dplyr::select(where(~ sd(.) > 0)) %>% 
  cor(use = "pairwise.complete.obs") # 'use' argument handles remaining NAs safely

# View highly correlated pairs

cor_matrix_view <- cor_matrix 

# Set upper triangle to NA in the copy only
cor_matrix_view[upper.tri(cor_matrix_view, diag = TRUE)] <- NA

# Flatten and filter the copy
high_correlations <- cor_matrix_view %>%
  as.table() %>%
  as.data.frame() %>%
  filter(!is.na(Freq)) %>%          # Remove the NAs we just made
  mutate(Freq_abs = abs(Freq)) %>% 
  filter(Freq_abs > 0.8) %>% 
  arrange(desc(Freq_abs))

#View(high_correlations)

View(high_correlations %>% filter(Var2=="pctglactilloamws"))

#Use findCorrelation() to drop highly correlated variables - 
#first ID indices of highly correlated predictors

highly_correlated_indices <- findCorrelation(cor_matrix, cutoff = 0.80, exact = FALSE)

# Extract names to remove
cor_cols_to_remove <- colnames(cor_matrix)[highly_correlated_indices]
print(cor_cols_to_remove)

#Remove highly correlated variabes
P_dataset <- P_dataset %>%
  dplyr::select(-all_of(cor_cols_to_remove))

P_dataset %>% 
    dplyr::select(order(colnames(.))) %>% 
  names()

```

# 5 Split data into training and testing

For remaining attributes, impute missing values. But first, must split data into training and testing datasets (to prevent data leakage).

```{r}
set.seed(123)
P_split <- initial_split(P_dataset, strata = P_mgkg)
#Note - Stratification ensures that the distribution of a variable is preserved across training and testing sets.
#Default split is 75% data in train, 25% in test

P_train <- training(P_split)
nrow(P_train)
names(P_train)

P_test <- testing(P_split)
nrow(P_test)
names(P_test)
```

# 6 Impute missing values:

```{r}
#Separate out predictor variables from dependent variable (P_mgkg):
#Also dropping bulk density for now

Predictors<-P_train %>% 
  dplyr::select(-any_of(c("P_mgkg", "BD_gcm3")))

#R with parameters set to run faster

start_time <- Sys.time()
start_time
dataImputed <- missRanger(Predictors,
                          formula = . ~ . ,
                          num.trees = 100, 
                          verbose = 1, seed = 111)

end_time <- Sys.time()
end_time - start_time

nrow(dataImputed)
#View(P_train)
#View(dataImputed)

#rematch to soil P
P.Predictors<-cbind(P_train$P_mgkg, dataImputed) 
P.Predictors<-P.Predictors %>% 
  dplyr::rename(P_mgkg = `P_train$P_mgkg`)

#View(P.Predictors)

###
#Impute *testing data* separately (for use in model testing later)
#Combine with predictor attributes to have more data to impute missing values

dataImputed<-P.Predictors %>% 
  dplyr::select(-P_mgkg)
names(dataImputed)


Test.impute<-rbind(
  (P_test %>% 
    dplyr::select(-any_of(c("P_mgkg", "BD_gcm3")))),
dataImputed)
nrow(Test.impute)

start_time <- Sys.time()
start_time
data.Test.Imputed <- missRanger(Test.impute,
                                formula = . ~ . ,
                                num.trees = 100, 
                                verbose = 1, seed = 111)

end_time <- Sys.time()
end_time - start_time

#select out test data; and attach to soil P
test<-data.Test.Imputed[c(1:nrow(P_test)),]
P.test<-P_test %>% dplyr::select(P_mgkg)
P.test
test.prep<-cbind(P.test, test)
nrow(test.prep)

View(P_test)
View(test.prep)

names(P_test)
names(test.prep)
```

## 6.1 Remove factors that only have 1 data point in any level

This is to ensure the model does not crash because there is a level in the test dataset that does not occur in the training dataset.

We also need to check if all levels of factors in the test data occur in the training dataset.

```{r}
#Do all factor levels in test dataset have an example in train dataset?

#eg, all factor levels in test data should occur in train data:
levels(factor(test.prep$NLCD_Land))
levels(factor(P.Predictors$NLCD_Land))

#Check if all levels of all factors in test data occur in train data: 
df1<-test.prep
df2<-P.Predictors

# Identify factor columns in df1
factor_cols_df1 <- names(df1)[sapply(df1, is.factor)]
factor_cols_df1


results <- lapply(factor_cols_df1, function(col_name) {
  if (col_name %in% names(df2) && is.factor(df2[[col_name]])) {
    levels_not_in_df2 <- setdiff(levels(df1[[col_name]]), levels(df2[[col_name]]))
    if (length(levels_not_in_df2) > 0) {
      return(paste0("Levels in '", col_name, "' of df1 not found in df2: ",
                    paste(levels_not_in_df2, collapse = ", ")))
    } else {
      return(paste0("All levels of '", col_name, "' in test dataset are present in predictor dataset."))
    }
  } else {
    return(paste0("Column '", col_name, "' is not a factor in predictor dataset or does not exist in predictor dataset."))
  }
})

# Print the results
for (res in results) {
  print(res)
}


#ID factors in training data with only a single data point in a given level: 

train_with_single_point_levels <- P.Predictors %>%
  dplyr::select(where(is.factor)) %>% # Select only factor columns
  sapply(function(x) any(table(x) == 1)) %>% # Check if any level has count 1
  names(.)[.] # Get the names of columns where the condition is TRUE
train_with_single_point_levels

#Drop factors from train and test with only a single data point in a given level 

P.Predictors <- P.Predictors %>%
  dplyr::select(-any_of(train_with_single_point_levels))

names(P.Predictors)

#Drop same factors from test data:

test.prep<-test.prep %>%
  dplyr::select(-any_of(train_with_single_point_levels))

names(test.prep)    

#check column match - should be 0 columns in test that are not in Predictors
check.col <- names(test.prep)[!names(test.prep) %in% names(P.Predictors)]
check.col
```

Write imputed and cleaned data to file

```{r}
#WRITE IMPUTED DATA TO FILE
#(note this is training data only)
write.table(P.Predictors, paste0(dir.output, "/Final_model_comparisons", model.input, "_train_imputed_less1000_03252026.csv"), sep=",", row.names=FALSE)

write.table(test.prep, paste0(dir.output, "/Final_model_comparisons", model.input, "_TEST_imputed_less1000_03252026.csv"), sep=",", row.names=FALSE)

```

# 7 Tune model hyperparameters

## 7.1 Random Forest

#### a. Run Version of Original Model from Dolph et al 2023

```{r}
#Define recipe to use in both model types 

P_recipe_RF <- 
  # which consists of the formula (outcome ~ predictors)
  recipe(P_mgkg ~ ., data = P.Predictors) %>%
  step_center(all_numeric_predictors()) %>%
  step_scale(all_numeric_predictors())
  #Other options excluded for now, eg:
  #%>%
  #step_dummy(all_nominal()) #%>%
  #step_impute_bag(all_predictors(), trees=100) #option to impute missing values, using all other predictors
P_recipe_RF


P_train_preprocessed <- P_recipe_RF %>%
  # apply the recipe to the training data
  prep(P.Predictors) %>%
  # extract the pre-processed training dataset
  juice()
P_train_preprocessed
names(P_train_preprocessed)

#So need to bake() and not juice
P_test_preprocessed <- P_recipe_RF %>%
  # apply the recipe to the training data
  prep(P.Predictors) %>%
  # extract the pre-processed testing dataset
  bake(new_data=test.prep)
nrow(P_test_preprocessed)
names(P_test_preprocessed)

short_model_RF<-
  rand_forest(
    mtry = 50, 
    min_n=3, 
    trees = 1000) %>%
  set_mode("regression") %>%
  set_engine("randomForest")

#put into workflow
short_wf_RF<-workflow()%>%
  add_recipe(P_recipe_RF) %>%
  add_model(short_model_RF)


#run models
RF_fit<-short_wf_RF %>%
  fit(P_train_preprocessed)
RF_fit
summary(RF_fit)


#Check performance
names(P_test_preprocessed)

randfor_pred<-predict(RF_fit, new_data=P_test_preprocessed) 
#add predictions to test data:
RF.pred<-as.data.frame(cbind(P_test_preprocessed, randfor_pred))

names(RF.pred)

#check R2 value for actual vs predicted
reg.AO<- with(RF.pred,lm(P_mgkg~.pred))
summary(reg.AO)


RMSE = mean((RF.pred$P_mgkg - RF.pred$.pred)^2) %>% sqrt()
RMSE

#Variable Importance, previous method:
##RUN CONDITIONAL PERMUTATION IMPORTANCE:
#read in saved model if needed;


#Need to rerun model using randomforest() package

start_time <- Sys.time()
start_time

set.seed(542863)
rfP_model <- randomForest(P_mgkg ~ ., data = as.data.frame(P_train_preprocessed), mtry = 50, ntree=1000, replace = FALSE, 
                          nodesize = 3
                          , keep.forest = TRUE, keep.inbag = TRUE)
end_time <- Sys.time()
end_time - start_time

summary(rfP_model)

#Note: permimp does require training data to run!

library(permimp)

start_time <- Sys.time()
start_time

CPI_permimpRF <- permimp(rfP_model, 
conditional = TRUE, progressBar = TRUE, do_check=FALSE)

CPI_permimpRF
end_time <- Sys.time()
end_time - start_time

#Rank importance values and plot a subset:
#Make a dot plot:
library(dplyr)
library(ggplot2)
VarImp<-as.data.frame(sort(CPI_permimpRF$values, decreasing=TRUE), optional=T)
colnames(VarImp)[1]<-"Imp"
rownames(VarImp)
VarImp2 <- cbind(rownames(VarImp), data.frame(VarImp, row.names=NULL))
colnames(VarImp2)[1]<-"Var"
head(VarImp2)
View(VarImp)
VarImp2[1:75,]


#save permimp object!
saveRDS(CPI_permimpRF, paste0(dir.output, model.input, "_permimp_results_less1000.rds"))


```

#### b. Tune New Random Forest Models

```{r}

library(foreach)
library(ranger)


P_recipe_RF <- 
  # which consists of the formula (outcome ~ predictors)
  recipe(P_mgkg ~ ., data = P.Predictors) %>%
  #step_center(all_numeric_predictors()) %>%
  #step_scale(all_numeric_predictors()) %>% 
  #Other options excluded for now, eg:
  step_dummy(all_nominal()) 

doParallel::registerDoParallel()

tune_spec <- rand_forest(
  mtry = tune(),
  trees = 1000,
  min_n = tune()
) %>%
  set_mode("regression") %>%
  set_engine("ranger")

#specify cross validation for the training dataset, note that default is k=10 folds
set.seed(234)
P_folds <- vfold_cv(P.Predictors)

###
#PUT EVERYTHING TOGETHER IN A WORKFLOW

tune_wf <- workflow() %>%
  add_recipe(P_recipe_RF) %>%
  add_model(tune_spec)


###
#TUNE THE PARAMETERS

#Check start time for initial tuning
start_time <- Sys.time()
start_time

set.seed(345)
tune_res <- tune_grid(
  tune_wf,
  resamples = P_folds,
  grid = 20,
  control = control_grid(save_pred = TRUE, verbose=TRUE)
)
tune_res


#Check end time for initial tuning: 
end_time <- Sys.time()
end_time - start_time #took ~3 min


#Check out initial tuning results
tune_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  select(mean, min_n, mtry) %>%
  pivot_longer(min_n:mtry,
               values_to = "value",
               names_to = "parameter") %>%
  arrange(mean) %>%
  print(n=40) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "rsq")


#find the best set:
best_rmse1<-select_best(tune_res, metric="rmse")
best_rmse1


####
#FINALIZE THE WORKFLOW

final_rf <- finalize_model(
  tune_spec,
  best_rmse1
)

final_rf

final_wf <- workflow() %>%
  add_recipe(recipe(P_mgkg ~ ., data = P.Predictors)) %>%
  add_model(final_rf)

###
#FIT THE MODEL BASED ON TRAINING DATA
rf_fit <- final_wf %>%
  fit(P.Predictors)
rf_fit

#Save the final model 
saveRDS(rf_fit, paste0(dir.output, model.input, "_RF_model_LIMITED_stepdummy_less1000_031326.rds"))

```

#### c. Check RF Model Performance

```{r}
#Predict on the independent test set
pred_test <- predict(rf_fit, new_data = test.prep)
pred_test 
                    
Test_predictions<-cbind(test.prep, pred_test)
names(Test_predictions)

#Calculate R-squared
rsq_test <- Test_predictions %>%
  rsq(truth = P_mgkg, estimate = .pred)

print(rsq_test)

rmse <- with(Test_predictions, sqrt(mean((P_mgkg - .pred)^2)))
print(rmse)

#Scatter plot: actual vs predicted
test.plot<-ggplot(Test_predictions, aes(x = P_mgkg, y = .pred)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    x = "Observed soil TP (mg/kg)",
    y = "Predicted soil TP (mg/kg)",
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test$.estimate, 3), ")")
  ) +
  theme_minimal()+
  xlim(0, 1500)+
  ylim(0,1500)
test.plot 

#Save to plot to file
png(paste0(file.name=dir.figures, "/Fig_pred_vs_obs_xgboost_reduced_with_WS_included.png"), width = 4, height = 4, units="in", res = 300)    

grid.newpage() 
print(test.plot) 

dev.off() 


#Check for samples with P<1000
rsq_test.less1000 <- Test_predictions %>% filter(P_mgkg<1000) %>%
  rsq(truth = P_mgkg, estimate = .pred)
rsq_test.less1000

rmse.less1000<-Test_predictions %>% 
    filter(P_mgkg<1000) %>% 
    mutate(rmse=sqrt(mean((P_mgkg - .pred)^2))) %>% 
    dplyr::select(rmse) %>% distinct()
rmse.less1000


#Plot Test data vs Predicted for samples <1000 mg/kg TP
ggplot(Test_predictions %>% filter(P_mgkg<1000), aes(x = P_mgkg, y = .pred)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    x = "Actual P_mgkg",
    y = "Predicted P_mgkg",
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test.less1000$.estimate, 3), ")")
  ) +
  theme_minimal()+
  xlim(0,1000)+
  ylim(0,1000)
```

#### d. Variable Importance

```{r}
#Run Model in randomForest
library(randomForest)

start_time <- Sys.time()
start_time

#check parameter values
best_rmse1

set.seed(542863)
rfP_model <- randomForest(mean.SRP ~ ., data = as.data.frame(P_train_preprocessed), 
                          mtry = best_rmse1$mtry, ntree=1000, replace = FALSE, #specify mtry and nodesize(same as min_n)
                          nodesize = best_rmse1$min_n,
                          keep.forest = TRUE, keep.inbag = TRUE)
end_time <- Sys.time()
end_time - start_time

summary(rfP_model)

#SAVE FINAL RANDOM FOREST MODEL!
saveRDS(rfP_model, "./FINALMODEL_randomForest_latesummer.rds")

#LOAD IN MODEL (if needed)
rfP_model<-readRDS("./FINALMODEL_randomForest_latesummer.rds")

print(rfP_model)
names(rfP_model)

rfP_model$mtry

#Check performance of the model (should match performance of the random_Forest() model tuned above, using tidymodels?)
rf.pred<-predict(rfP_model, newdata=P_test_preprocessed)
head(rf.pred)

#Match actual to predicted data:
predicted.P<-cbind(as.data.frame(P_test_preprocessed), rf.pred)
head(predicted.P)

#actual vs predicted
reg.check<-with(predicted.P, lm(mean.SRP~rf.pred))
summary(reg.check)
#calculate rmse
RMSE = mean((predicted.P$mean.SRP - predicted.P$rf.pred)^2) %>% sqrt()
RMSE

names(predicted.P)
names(P.Predictors)


##RUN CONDITIONAL PERMUTATION IMPORTANCE:
#read in saved model if needed;
setwd(output_dir)
rfP_model<-readRDS("./FINALMODEL_randomforest_latesummer.rds")
rfP_model

#Note: permimp does require training data to run!

start_time <- Sys.time()
start_time

CPI_permimpRF <- permimp(rfP_model, 
                         conditional = TRUE, progressBar = TRUE, do_check=FALSE)

CPI_permimpRF
end_time <- Sys.time()
end_time - start_time #took 14 minutes

#save permimp object!
saveRDS(CPI_permimpRF, "./permimp_results_latesummer.rds")


#read in permimp results if needed
CPI_permimpRF<-readRDS("./permimp_results_latesummer.rds")


#Rank importance values and plot a subset:
#Make a dot plot:
#library(dplyr)
#library(ggplot2)
VarImp<-as.data.frame(sort(CPI_permimpRF$values, decreasing=TRUE), optional=T)
colnames(VarImp)[1]<-"Imp"
rownames(VarImp)
VarImp2 <- cbind(rownames(VarImp), data.frame(VarImp, row.names=NULL))
colnames(VarImp2)[1]<-"Var"
head(VarImp2)
View(VarImp)
VarImp2[1:75,]
```

## 7.2 XGBoost

### a. Tune XGBoost Model

Helpful resources:

<https://juliasilge.com/blog/xgboost-tune-volleyball/>

Note that grid_space_filling() is the newer preferred way to create space-filling hyperparameter grids, it uses a latin hypercube by default

XGBoost model is based on trees - don’t need much preprocessing for our data, ie don’t need to worry about factors or centering or scaling our data. Can just go straight to setting up our model specification.

Note that in tidy models, "xgboost does not have a means to translate factor predictors to grouped slplits. Factor/categorical predictors need to be converted to numeric values (dummy or indicator variables) for this engine. When using the formula via fit.model_spec(), parsnip will convert factor columns to indicators using a one-hot encoding"

Citation:<https://parsnip.tidymodels.org/reference/details_boost_tree_xgboost.html>

```{r}

#Specify model type and hyperparameters to tune 

xgb_spec <- boost_tree(
  trees = 1000,
  tree_depth = tune(), min_n = tune(),
  loss_reduction = tune(), ## first three: model complexity
  sample_size = tune(), mtry = tune(),  ## randomness
  learn_rate = tune()                   ## step size
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_spec
```

Because there are a lot of hyperparameters to tune, reduce computational complexity by setting up possible values for hyperparameters that cover the hyperparameter space.

Note on setting the size of the search space: \~ 10 observations per parameter being tuned are needed to adequately cover the search space - **need a citation for this!**

Here we are tuning 6 parameters (`tree_depth`, `min_n`, `loss_reduction`, `sample_prop`, `mtry`, `learn_rate`)

Recommended Size: 6 x 10 = 60. We will search 80 just to be extra safe. (Tho note this will substantially increase computational time!)

```{r}

#NEW WAY 
#Set up hyperparameter space to search 

set.seed(123)
xgb_grid <- grid_space_filling(
  parameters(
    tree_depth(),
    min_n(),
    loss_reduction(),
    sample_prop(),
    finalize(mtry(), P.Predictors),
    learn_rate()
  ),
  size = 60
)

xgb_grid
```

Put model into a workflow

```{r}

#Original:
#xgb_wf <- workflow() %>%
#  add_formula(P_mgkg ~ .) %>%
#  add_model(xgb_spec)

#xgb_wf

#With recipe (for use with shapviz later): 

P_recipe <- P.Predictors %>%
 recipe(P_mgkg ~ .) %>% 
 #step_integer(all_nominal()) #integer encode factors
 step_dummy(all_nominal_predictors(), one_hot = TRUE) 

#Model workflow:
xgb_wf <- workflow() %>%
  add_recipe(P_recipe) %>%
  add_model(xgb_spec)


```

Create cross validation resamples for tuning

```{r}
set.seed(123)
P_folds <- vfold_cv(P.Predictors, strata = P_mgkg)

P_folds
```

Tune model:

-Update: takes \~1-2 hours for full USGS/USDA soil P database, 30 min for IL streambank samples, for a search space of 30 and full variables

-Update 2: for a reduced dataset (dropping co-linear variables), and a larger search space (n=80), takes about 13-24 minutes for IL streambank samples

Note for small datasets: The following warning might arise if a dataset is so small that trees can't split according to the hyperparameter values in parameter space: 'A correlation computation is required, but 'estimate' is constant and has 0 standard deviation, resulting in a divide by 0 error. 'NA' will be returned'

See below for a possible alternative solution for small datasets (Bayesian tuning)

```{r}

#Read in training data (if not already loaded above)
#P.Predictors<-read.csv(paste0(dir.output, model.input, "_train_imputed.csv"), sep=",")
#names(P.Predictors)
nrow(P.Predictors) 

#Tune model: 
start.time<-Sys.time()
start.time

doParallel::registerDoParallel()

set.seed(234)
xgb_res <- tune_grid(
  xgb_wf,
  resamples = P_folds,
  grid = xgb_grid,
  control = control_grid(save_pred = TRUE, verbose=TRUE)
)

xgb_res

end.time<-Sys.time()
end.time

total.time<-end.time-start.time
total.time
```

Explore XGBoost model results:

```{r}
collect_metrics(xgb_res)
```

Visualize tuning results:

```{r}
names(collect_metrics(xgb_res))

View(collect_metrics(xgb_res))

xgb_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  dplyr::select(mean, mtry:sample_size) %>%
  pivot_longer(mtry:sample_size,
               values_to = "value",
               names_to = "parameter"
  ) %>%
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "RMSE")
```

Find best set of parameters:

```{r}
best_rmse<-show_best(xgb_res, metric="rmse", n=1)
best_rmse
View(best_rmse)

best_rsq<-show_best(xgb_res, metric="rsq", n=1)
best_rsq
```

Finalize tunable workflow with the best parameters:

```{r}
final_xgb <- finalize_workflow(
  xgb_wf,
  best_rmse
)

final_xgb
```

### b. Save final model object

```{r}
#set directories for final saved models + data

dir.final.model.data<-paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets")

dir.final.model<-paste0(dir.output, "/Final_model_comparisons")

#Run the final model:
set.seed(123)
final_fit <- final_xgb %>%
  fit(data = P.Predictors)

# Save the trained model object
saveRDS(final_fit, file = paste0(dir.final.model, model.input, "_xgb_model_LIMITED_stepdummy_less1000_03252026.rds"))

```

### c. Load Saved Model (as needed)

And load saved model input and validation data

```{r}

model.input

# Load saved model
#final_fit <- readRDS(paste0(dir.output, model.input, "_xgb_model_LIMITED_less1000_corReduced_stepdummy_021926.rds"))

final_fit<-readRDS(paste0(dir.final.model, model.input, "_xgb_model_LIMITED_stepdummy_less1000_reducedCor_pH_03182026.rds"))


summary(final_fit)
#summary(final_fit_all)


#READ IMPUTED DATA
dir.final.model.data<-paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets")
#(note this is training data only)
P.Predictors<-read.csv(paste0(dir.final.model.data, model.input, "_train_imputed_less1000_reducedCor_03182026_pH.csv"))

P.Predictors$NLCD_Land<-factor(P.Predictors$NLCD_Land)

#View(P.Predictors)

#(test data)
test.prep<-read.csv(paste0(dir.final.model.data, model.input, "_TEST_imputed_less1000_reduced_Cor_03182026_pH.csv"))

test.prep$NLCD_Land<-factor(test.prep$NLCD_Land)

#View(test.prep)

#SET ATTRIBUTE TYPES
#Set categorical attributes to factor
cat_vars <- names(P.Predictors)[sapply(P.Predictors, function(x) is.factor(x) | is.character(x))]
cat_vars

P.Predictors<-P.Predictors %>% 
  mutate(across(cat_vars, as.factor))
test.prep<-test.prep %>% 
  mutate(across(cat_vars, as.factor))

#Set remaining numeric attributes to numeric
dep_var="P_mgkg"
num_vars <- names(P.Predictors)[sapply(P.Predictors, is.numeric) & names(P.Predictors) != dep_var]
num_vars

P.Predictors <- P.Predictors %>%
  mutate(across(num_vars, as.numeric))
test.prep<-test.prep %>% 
  mutate(across(num_vars, as.numeric))

sapply(P.Predictors, class)
```

### d. Check model hyperparameters

```{r}

#Different approaches for examing hyperparameter values:

print(final_fit) 

xgb_engine_fit <- extract_fit_engine(final_fit)

print(xgb_engine_fit$params)

```

### e. Variable importance: XGBoost

#### i. Built in variable importance for XGBoost

Based on "Gain" - the improvement in accuracy (loss reduction) brought by a feature to the branches it is used in.It measures how much each feature contributes to the model's predictive power, with higher values indicating more important features. Can also evaluate based on Cover or Frequency

```{r}

#use vip() to automatically plot importance:

final_fit %>%
  extract_fit_parsnip() %>%
  vip(geom = "point", num_features = 10)+
  theme_bw()+
  theme(panel.grid=element_blank())+
  #scale_y_discrete(labels = my_labels)+
  xlab("Variable Importance")


#A more adjustable plot:


#extract model first 

xgb_booster_object <- extract_fit_engine(final_fit)  

#importance measures
importance <- xgb.importance(model = xgb_booster_object) 

importance

#ID top 10 most important predictors 
top_vars <- importance$Feature[1:10] 
top_vars 

#make nicer labels
#Cat scale predictors only:
labels = list(
  "Water input",
  "% Glacial till, loamy (watershed)",
  "Soil erodibility factor",
  "Depth (cm)",
  "% Glacial till, loamy (catchment)",
  "WWTP density",
  "% Ag drainage",
  "% Hay",
  "Hydraulic conductivity",
  "% Mixed forest"
              
            #"Net anthropogenic N inputs",
             # "Annual min air temp, 1981-2010",
              #"Mean topographic wetness index",
              #"Sediment regulation index",
              #"Elevation",
              #"Slope Length - USLE",
              #"Population density",
              #"Soil organic carbon, 100-150cm"
)

#Watershed and Cat scale predictors:
labels = list("Depth",
              "Water input",
              "WWTP density",
              "Mean annual precip",
              "% ag on high slopes",
              "Elevation",
              "% Glacial till, loamy",
              "Soil erodibility",
              "N from rock weathering",
              "Water withdrawals"
              )


#select top 10 by gain and assign better labels
imp.data<-importance %>% 
  arrange(desc(as.numeric(Gain))) %>% 
   slice_max(Gain, n = 10)

imp.data$labels<-as.character(labels)
imp.data


#Make nicer plot: 
imp.plot<-ggplot(imp.data)+
  geom_point(aes(x=Gain, y = reorder(labels, Gain)), size=3)+
  theme_bw()+
  theme(panel.grid=element_blank(), axis.text=element_text(size=16), axis.title=element_text(size=18))+
  #scale_y_discrete(labels = my_labels)+
  xlab("Variable Importance")+
  ylab("Model Predictors")
imp.plot

#save plot to file

png(paste0(file.name=dir.figures, "/Fig_streambank_importance_plot_WSincluded_corReduced_less1000.png"), width = 2900, height = 1800, res = 300)    

grid.newpage() 
print(imp.plot) 

dev.off()  
```

#### ii. Shapley Importance Values

There are multiple packages/approaches for illustrating SHAP values:

1.  shapviz
2.  SHAPforxgboost
3.  xgboost has a native shap function (xgb.plot.shap)

```{r}

#####################################
#1. Using shapviz method
#install.packages("shapviz")
library(shapviz)


#Recall P recipe
P_recipe <- P.Predictors %>%
 recipe(P_mgkg ~ .) %>% 
 #step_integer(all_nominal()) #integer encode factors
 step_dummy(all_nominal_predictors(), one_hot = TRUE) 


# The 'X_pred' argument in shapviz() is prepared predictor matrix
set.seed(123)
shap <- shapviz(extract_fit_engine(final_fit), 
                X_pred = bake(
      prep(P_recipe),
      has_role("predictor"),
      new_data = NULL,
      composition = "matrix"
    )
  )
shap

View(shap)

#Importance plot for top predictors: 
sv_importance(shap, kind = "beeswarm", show_numbers = TRUE, max_display=15)

#Importance plot for all features for Appendix:
shap.plot.all<-sv_importance(shap, kind = "beeswarm", show_numbers = TRUE, max_display=Inf, number_size=5)+
  theme(axis.text=element_text(size=16), legend.text=element_text(size=16), legend.title=element_text(size=16))
shap.plot.all
#save plot to file

png(paste0(file.name=dir.figures, "/Fig_streambank_SHAP_importance_plot_1_WSincluded_corReduced_less1000_pH_03182026_ALL_FEATURES.png"), width = 4000, height = 6000, res = 300)    

grid.newpage() 
shap.plot.all
dev.off() 




#top 10 features with nicer labels:
shap.plot1<-sv_importance(shap, kind = "beeswarm", show_numbers = TRUE, max_display=10)+
  scale_y_discrete(labels = c(
    "waterinputcat" = "Water input",
    "pctglactilloamws"= "% Glacial till, loamy (watershed)",
    "kffactws" = "Soil erodibility factor",
    "Depth_cm" = "Depth (cm)",
    "hydricrating_No" = "Not hydric soils",
    "wwtpmajordensws" = "WWTP density",
    "pctglactilloamcat" = "% Glacial till, loamy (catchment)",
  
  "tk100_150s" = "Soil thickness, 100-150cm",
  "rocknws" = "N from rock weathering", 
  "pcthay2011ws" = "% Hay",
  "popden2010ws" = "Population density",
  "precip2009ws" = "Mean annual precipitation (mm) ",
  "omws" = "Organic matter"))+
  theme_bw()
shap.plot1

#top 10 with site specific characteristics included:
shap.plot2<-sv_importance(shap, kind = "beeswarm", show_numbers = TRUE, max_display=10)+
  scale_y_discrete(labels = c(
    "pct.silt" = "% Silt",
    "pct.clay" = "% Clay",
    "waterinputcat" = "Water input",
    "pctglactilloamws"= "% Glacial till, loamy (watershed)",
    "pH" = "pH",
    "manurews" = "Manure inputs (kg N/ha/yr)",
    "kffactws" = "Soil erodibility factor",
    "omws" = "Organic matter",
    "pctglactilloamcat" = "% Glacial till, loamy (catchment)",
    "hydricrating_No" = "Not hydric soils"
    ))+
  theme_bw()
shap.plot2



#save plot to file

png(paste0(file.name=dir.figures, "/Fig_streambank_SHAP_importance_plot_1_WSincluded_corReduced_less1000_pH_03182026.png"), width = 2200, height = 1400, res = 300)    

grid.newpage() 
shap.plot2
dev.off()  

```

SHAP Dependency Plots:

```{r}
#Dependency plots using shapviz:

#For a specific variable:

sv_dependence(shap, v="pctmxfst2011cat", share_y=TRUE, color_var=NULL)+xlab("% Mixed Foreste")

#Original Streambank P model (no site specific features)

#plot3<-sv_dependence(shap, v="Depth_cm", share_y=TRUE, color_var=NULL)+xlab("Depth (cm)")

#plot4<-sv_dependence(shap, v="kffactws", share_y=TRUE, color_var=NULL)+xlab("Soil erodibility")

#plot5<-sv_dependence(shap, v="pctglactilloamcat", share_y=TRUE, color_var=NULL)+xlab("% Glacial till, loamy \n (catchment)")

#plot6<-sv_dependence(shap, v="hydricrating_No", share_y=TRUE, color_var=NULL)+ xlab("Not hydric soils")

#for some reason, this plot coming out incorrectly (are no negative values for wwtpmajordensws)
#plot5<-sv_dependence(shap, v="wwtpmajordensws", share_y=TRUE, color_var=NULL)+xlab("WWTP density")
summary(P.Predictors$wwtpmajordensws)
#plotting manually:
sv_dependence(shap, v="wwtpmajordensws", share_y=TRUE)

wwtp.shap.data<-as.data.frame(shap[["S"]]) %>% dplyr::select(wwtpmajordensws)
head(wwtp.shap.data)

og.data<-as.data.frame(shap[["X"]][["wwtpmajordensws"]])
head(og.data)
colnames(og.data)[1]<-"WWTP_dens"

wwtp.shap<-cbind(wwtp.shap.data, og.data)

#plot7<-ggplot(wwtp.shap)+
  geom_point(aes(WWTP_dens, wwtpmajordensws), col="#3b528b")+
  xlab("WWTP density")+
  ylab("SHAP value")
#plot7


#plot8<-sv_dependence(shap, v="precip2009ws", share_y=TRUE, color_var=NULL)+xlab("Mean Precipitation (mm)")

#plot9<-sv_dependence(shap, v="omws", share_y=TRUE, color_var=NULL)+xlab ("Organic Matter")

#plot10<-sv_dependence(shap, v = "rocknws", share_y = TRUE, color_var=NULL)+ xlab("N from rock weathering")

-----------------------
## Site Specific features included

plot1<-sv_dependence(shap, v="pct.silt", share_y=TRUE, color_var=NULL)+xlab("% Silt")

plot2<-sv_dependence(shap, v="pct.clay", share_y=TRUE, color_var=NULL)+xlab("% Clay")
  
plot3<-sv_dependence(shap, v="waterinputcat", share_y=TRUE, color_var=NULL)+xlab("Water input")

plot4<-sv_dependence(shap, v="pctglactilloamws", share_y=TRUE, color_var=NULL)+xlab("% Glacial till, loamy \n (watershed)")
  
plot5<-sv_dependence(shap, v="pH", share_y=TRUE, color_var=NULL)+xlab("pH")

plot6<-sv_dependence(shap, v="manurews", share_y=TRUE, color_var=NULL)+xlab("Manure (kg N/ha/yr)")

plot7<-sv_dependence(shap, v="kffactws", share_y=TRUE, color_var=NULL)+xlab("Soil erodibility")

plot8<-sv_dependence(shap, v="omws", share_y=TRUE, color_var=NULL)+xlab("Organic Matter")

plot9<-sv_dependence(shap, v="pctglactilloamcat", share_y=TRUE, color_var=NULL)+xlab("% Glacial till, loamy \n (catchment)")

plot10<-sv_dependence(shap, v="hydricrating_No", share_y=TRUE, color_var=NULL)+ xlab("Not hydric soils")




library(gridExtra)

grid.arrange(plot1,
plot2,
plot3,
plot4,
plot5,
plot6,
plot7,
plot8,
plot9,
plot10, nrow=5, ncol=2)

#save plot to file 
png(paste0(file.name=dir.figures, "/Fig_streambank_SHAP_dependency_WSincluded_corReduced_less1000_pH_03182026.png"), width = 1400, height = 3000, res = 300)    

grid.arrange(plot1,
plot2,
plot3,
plot4,
plot5,
plot6,
plot7,
plot8,
plot9,
plot10, nrow=5, ncol=2)

dev.off()

##
#Other plot types:

sv_force(shap, row_id = 1)
sv_waterfall(shap, row_id = 1)

#####################################
#2. Using SHAPforxgboost method
#(produces same results as above)

library(SHAPforxgboost)

game_shap <-
  shap.prep(
    xgb_model = extract_fit_engine(final_fit),
    X_train = bake(
      prep(P_recipe),
      has_role("predictor"),
      new_data = NULL,
      composition = "matrix"
    )
  )

shap.plot2<-shap.plot.summary(game_shap) #this shows SHAP values for all predictors 
shap.plot2

png(paste0(file.name=dir.figures, "/Fig_streambank_SHAP_importance_plot_2+WSincluded_corReduced_less1000.png"), width = 1800, height = 2900, res = 300)    
grid.newpage() 
shap.plot2
dev.off()  

##################################
#3. Using native xgboost method 

#One dependency plot for all top variables of interest
#**Note: this is showing slightly different variable importance order than previous two methods! **

xgb.plot.shap(data = bake(
      prep(P_recipe),
      has_role("predictor"),
      new_data = NULL,
      composition = "matrix"
    ), 
    model = extract_fit_engine(final_fit), top_n = 5)          
             
#save plot to file

png(paste0(file.name=dir.figures, "/Fig_streambank_SHAP_importance_plot_3_WSincluded_corReduced_less1000.png"), width = 1800, height = 2900, res = 300)    

grid.newpage() 
xgb.plot.shap(data = bake(
      prep(P_recipe),
      has_role("predictor"),
      new_data = NULL,
      composition = "matrix"
    ), 
    model = extract_fit_engine(final_fit), top_n = 15)  
dev.off()  


```

### f. Partial dependence plots:

Use DALEX for partial dependence plots when using tidymodels

Example: <https://juliasilge.com/blog/mario-kart/>

```{r}
#Use a model_explainer from DALEXtra to create partial dependence plots:  
model_explainer <- explain_tidymodels(   
  final_fit,   
  data = dplyr::select(P.Predictors, -P_mgkg),   
  y = P.Predictors$P_mgkg,   
  verbose = FALSE )  

#Plot PDP plots for top 10 most important variables:  
pdp_time <- model_profile(   
  model_explainer,   
  variables = top_vars,   
  N = NULL )  

plot(pdp_time)
```

Make fancier Partial Dependence Plots for manuscript:

```{r}

#Needs: variables in correct order
#be able to control plot aesthetics 

names(pdp_time)

#look at top 10 most important predictors:
top_vars

#Extract relationship between variables and predicted dependent variable:

PDP<-as.data.frame(pdp_time$agr_profiles)
names(PDP)

#Remove funky formatting and 
#Select top 10 variables:
PDP.sel<-PDP %>% 
  rename_with(~ str_remove_all(.x, "^_|_$"))%>% 
  filter(vname %in% top_vars)

#Make nicer labels for predictors and put them in right order: 
PDP.sel$PDP.label<-factor(PDP.sel$vname, levels=top_vars, labels=labels)
head(PDP.sel)

#Plot using ggplot: 

PDP.Plot<-ggplot(PDP.sel, aes(x = x, y = yhat)) +
  geom_line(color = "blue", linewidth = 1.2) + # Change line style here
  facet_wrap(~ PDP.label, scales = "free_x", ncol = 4, labeller = label_wrap_gen(width = 20)) + # 'free_x' allows different units; wrap strip text
  labs(
    y = "Predicted P (mg/kg)",
    x = "Predictor Value",
    title = "Partial Dependence Plots"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.background = element_rect(fill = "gray95"), # Box behind variable names
    strip.text = element_text(face = "bold", size = 10)
  )
PDP.Plot


#save plot to file

png(paste0(file.name=dir.figures, "/Fig_PDP_plot_streambank_WS_included_corReduced_less1000.png"), width = 2000, height = 2000, res = 300)    
grid.newpage() 
print(PDP.Plot) 

dev.off()  

```

### g. Model Performance: Independent validation data

Fit model to test data and evaluate accuracy

```{r}

#Predict on the independent test set
pred_test <- predict(final_fit, new_data = test.prep)
pred_test 
                    
Test_predictions<-cbind(test.prep, pred_test)
names(Test_predictions)

#Calculate R-squared
rsq_test <- Test_predictions %>%
  rsq(truth = P_mgkg, estimate = .pred)

print(rsq_test)

rmse <- with(Test_predictions, sqrt(mean((P_mgkg - .pred)^2)))
print(rmse)

#Scatter plot: actual vs predicted
test.plot<-ggplot(Test_predictions, aes(x = P_mgkg, y = .pred)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test$.estimate, 2), ")")
  ) +
  xlab(bquote(Observed~soil~TP~(mg~kg^-1)))+
  ylab(bquote(Predicted~soil~TP~(mg~kg^-1)))+
  theme_minimal()+
  xlim(0, 1000)+
  ylim(0,1000)
test.plot 

#Save to plot to file
png(paste0(file.name=dir.figures, "/Fig_STREAMBANK_pred_vs_obs_xgboost_reduced_with_WS_included_less1000_reducedcor_pH_031826.png"), width = 4, height = 4, units="in", res = 300)    

grid.newpage() 
print(test.plot) 

dev.off() 


#Check for samples with P<1000
rsq_test.less1000 <- Test_predictions %>% filter(P_mgkg<1000) %>%
  rsq(truth = P_mgkg, estimate = .pred)
rsq_test.less1000

rmse.less1000<-Test_predictions %>% 
    filter(P_mgkg<1000) %>% 
    mutate(rmse=sqrt(mean((P_mgkg - .pred)^2))) %>% 
    dplyr::select(rmse) %>% distinct()
rmse.less1000


#Plot Test data vs Predicted for samples <1000 mg/kg TP
ggplot(Test_predictions %>% filter(P_mgkg<1000), aes(x = P_mgkg, y = .pred)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    x = "Actual P_mgkg",
    y = "Predicted P_mgkg",
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test.less1000$.estimate, 3), ")")
  ) +
  theme_minimal()+
  xlim(0,1000)+
  ylim(0,1000)


```

Evaluating "acceptability" of model performance, based on Moriasi et al 2007

<https://swat.tamu.edu/media/1312/moriasimodeleval.pdf>

RSR - Based on the recommendation by Singh et al. (2004), a model evaluation statistic, named the RMSE-observations standard deviation ratio (RSR), was developed. RSR standardizes RMSE using the observations standard deviation, and it combines both an error index and the additional information recommended by Legates and McCabe (1999). RSR is calculated as the ratio of the RMSE and standard deviation of measured data

NSE ranges between −∞ and 1.0 (1 inclusive), with NSE = 1 being the optimal value. Values between 0.0 and 1.0 are generally viewed as acceptable levels of performance, whereas values \< 0.0 indicates that the mean observed value is a better predictor than the simulated value, which indicates unacceptable performance.

In general, model simulation can be judged as satisfactory if NSE \> 0.50 and RSR \< 0.70, and if PBIAS 25% for streamflow, PBIAS 55% for sediment, and PBIAS 70% for N and P

```{r}
#RMSE as a % of sample average
mean(P.Predictors$P_mgkg)
rmse/mean(P.Predictors$P_mgkg)

#RSR - RMSE-observations standard deviation ratio (RSR)
#< 0.7 is satisfactory
rmse/sd(P.Predictors$P_mgkg)

#Nash Sutcliffe efficiency (NSE)
#should be >0
Test_predictions %>% 
  mutate(Diff=(P_mgkg-.pred)^2) %>% 
  mutate(Diff.mean=(P_mgkg-mean(P_mgkg))^2) %>% 
  summarise(sum.diff=sum(Diff), sum.diff.mean=sum(Diff.mean)) %>% mutate(NSE=1-(sum.diff/sum.diff.mean))

```

### h. Alternative tuning method for small datasets - try a Bayesian tuning approach

Define search space with safer bounds for small dataset:

```{r}
param_set <- parameters(   
  tree_depth(range = c(2L, 6L)),   
  min_n(range = c(2L, 20L)),   
  loss_reduction(range = c(0, 5)),   # gamma, no negative   
  sample_size = sample_prop(range = c(0.6, 1.0)),   
  finalize(mtry(), P.Predictors),   
  learn_rate(range = c(0.01, 0.3)) )
```

Bayesian tuning:

Start with some random trials (`initial`) and then let Bayesian optimization adapt:

Note: this is slow, even for a small dataset

```{r}
set.seed(345) 
xgb_bayes <- tune_bayes(   
  xgb_wf,   
  resamples = P_folds,   
  param_info = param_set,   
  metrics = metric_set(rmse, mae),   
  initial = 10, # initial random search   
  iter = 20, # number of Bayesian iterations   
  control = control_bayes(verbose = TRUE, no_improve = 10, 
  save_pred=TRUE) )
```

Plot tuning improvements over iterations

```{r}
bayes_results <- xgb_bayes %>% collect_metrics() bayes_params <- xgb_bayes %>% collect_predictions()  xgb_bayes %>% 
  show_best(metric = "rmse", n = 20) 
# top-performing iterations  

# Plot RMSE over iterations 
bayes_results %>%   
  filter(.metric == "rmse") %>%   
  mutate(iter = 1:n()) %>%  # iteration order   
  
  ggplot(aes(x = iter, y = mean)) +   
  geom_point(color = "blue") +   
  geom_line(color = "darkblue") +   
  labs(     
    x = "Iteration",     
    y = "Mean RMSE",     
    title = "Bayesian Optimization Progress"   ) +   
  theme_minimal()
```

Pick best model:

```{r}
best_bayes <- select_best(xgb_bayes, metric = "rmse")
best_bayes 

final_xgb <- finalize_workflow(xgb_wf, best_bayes)  

final_fit <- fit(final_xgb, data = P.Predictors)

library(vip)  
final_fit %>%   
  extract_fit_engine() %>%   
  vip(geom = "point")
```

# 8 Run Model on Additional Test Datasets

Steps:

-   Match predictor variables of new dataset to model dataset

    -   Values for predictors should be within range of model dataset

-   Impute missing values for new dataset, using model dataset to help inform

-   Run new dataset through the model

-   Plot linear relationship

## 8.1 Load additional data you'd like to test

```{r}

#1. Name additional dataset(s) you'd like to test:

new.test.input<-"/Margenot_sample_attributes" #Margenot IL data

#2. Load data:

list.files(dir.output)

new.test.data<-read.csv(paste0(dir.output, new.test.input, '_for_modelling_wSSURGO_LIMITED.csv'))
head(new.test.data)
nrow(new.test.data)

sort(names(new.test.data))

```

## 8.2 Load model you wish to use to predict values for new dataset

```{r}

#set directory for final saved models
dir.final.model<-paste0(dir.output, "/Final_model_comparisons")

#specify which model
model.input<-"/USGS_NCSS_soils_database_attributes_alldepths_for_modelling_wSSURGO_LIMITED"

#read in model

final_fit<-readRDS(paste0(dir.output, "/Final_model_comparisons", model.input, "_xgb_model_LIMITED_less1000_corReduced_stepdummy_03032026.rds"))

```

## 8.3 Make sure predictors in new dataset match predictors in the model

```{r}
#3. Make sure predictor columns of new data match predictor columns of data used to build the model

#set directory for final train/test datasets
dir.final.model.data<-paste0(dir.output, "/Final_model_comparisons/Final_train_test_datasets")



#Read in desired predictor dataset

P.Predictors<-read.csv(paste0(dir.final.model.data, model.input, "_train_imputed_less1000_reducedCor_03032026.csv"))

names(P.Predictors)

new.test.data2<-new.test.data %>%
  { if ("meanP_mgkg" %in% names(.)) rename(., P_mgkg = meanP_mgkg) else . } %>% #rename P column if took average of subsamples
  dplyr::select(any_of(names(P.Predictors))) #grab all columns that match P.Predictors 

names(new.test.data2)

#Do all columns match?
all(sort(names(new.test.data2)) == sort(names(P.Predictors))) # should be TRUE

#just in case, find missing columns:
setdiff(names(P.Predictors), names(new.test.data2))

setdiff(names(new.test.data2), names(P.Predictors))


names(new.test.data2)

```

## 8.4 Restrict data to samples with TP \< 1000 mg/kg (Optional)

```{r}

new.test.data2<-new.test.data2 %>% 
  filter(P_mgkg<1000)
```

## 8.5 Impute missing values for new data

```{r}
#4. Impute missing values separately
#Combine with predictor attributes to have more data to impute missing values

dataImputed<-P.Predictors %>% 
  dplyr::select(-P_mgkg)
names(dataImputed)


NEW.Test.impute<-rbind(
  (new.test.data2 %>% 
    dplyr::select(-any_of(c("P_mgkg", "meanBD_gcm3")))),
dataImputed)
nrow(NEW.Test.impute)

start_time <- Sys.time()
start_time
NEW.data.Test.Imputed <- missRanger(NEW.Test.impute,
                                formula = . ~ . ,
                                num.trees = 100, 
                                verbose = 1, seed = 111)

end_time <- Sys.time()
end_time - start_time

#select out test data; and attach to soil P
NEW.test<-NEW.data.Test.Imputed[c(1:nrow(new.test.data2)),]
NEW.P<-new.test.data2 %>% dplyr::select(P_mgkg)
NEW.P
NEW.DATA<-cbind(NEW.P, NEW.test)
nrow(NEW.DATA)


#View(new.test.data2)
#View(NEW.test)
#View(NEW.DATA)
```

## 8.6 Predict modeled values for the new dataset

```{r}
# 5. Predict on the independent test set
NEW.pred_test <- predict(final_fit, new_data = NEW.DATA)
NEW.pred_test 
                    
NEW.Test_predictions<-cbind(NEW.DATA, NEW.pred_test)
names(NEW.Test_predictions)


```

Evaluate prediction performance

```{r}
# 6. Calculate R-squared
rsq_test <- NEW.Test_predictions %>%
  rsq(truth = P_mgkg, estimate = .pred)

print(rsq_test)

# 7. Scatter plot: actual vs predicted
streambank.test.plot<-ggplot(NEW.Test_predictions, aes(x = P_mgkg, y = .pred)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test$.estimate, 2), ")")
  ) +
  xlab(bquote(Observed~soil~TP~(mg~kg^-1)))+
  ylab(bquote(Predicted~soil~TP~(mg~kg^-1)))+
  theme_minimal()+
  xlim(0,1000)+
  ylim(0,1000)
streambank.test.plot

#Check for samples with P<1000
rsq_test.less1000 <- NEW.Test_predictions %>% filter(P_mgkg<1000) %>%
  rsq(truth = P_mgkg, estimate = .pred)
rsq_test.less1000


ggplot(NEW.Test_predictions %>% filter(P_mgkg<1000), aes(x = P_mgkg, y = .pred)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_smooth(method="lm", color="black", se=FALSE)+
  labs(
    x = "Actual streambank TP (mgkg)",
    y = "Predicted streambank TP (mg/kg)",
    title = paste0("XGBoost Predictions (R² = ", round(rsq_test.less1000$.estimate, 2), ")")
  ) +
  theme_minimal()


#Save to plot to file
png(paste0(file.name=dir.figures, "/Fig_pred_vs_obs_soilTP_xgboost_reduced_with_WS_included_less1000_reducedcor_021126_STREAMBANK_PREDICTED.png"), width = 4, height = 4, units="in", res = 300)    

grid.newpage() 
print(streambank.test.plot) 

dev.off() 
```
